### 2023-04-25 paper/paper.tex
add paper

\documentclass[9pt,twocolumn,twoside]{gsajnl_modified}
% Use the documentclass option 'lineno' to view line numbers

\usepackage[htt]{hyphenat}  % https://tex.stackexchange.com/a/543
\usepackage[export]{adjustbox}
\usepackage{xurl}
\usepackage{stfloats}
\usepackage{multirow}
\usepackage{makecell}
 \renewcommand\theadfont{}
\usepackage[leftcaption]{sidecap}
\sidecaptionvpos{figure}{t}

\renewcommand{\topfraction}{0.9}	% max fraction of floats at top
    \renewcommand{\bottomfraction}{0.8}	% max fraction of floats at bottom
    %   Parameters for TEXT pages (not float pages):
    \setcounter{topnumber}{2}
    \setcounter{bottomnumber}{2}
    \setcounter{totalnumber}{4}     % 2 may work better
    \setcounter{dbltopnumber}{2}    % for 2-column pages
    \renewcommand{\dbltopfraction}{0.9}	% fit big float above 2-col. text
    \renewcommand{\textfraction}{0.07}	% allow minimal text w. figs
    %   Parameters for FLOAT pages (not text pages):
    \renewcommand{\floatpagefraction}{0.7}	% require fuller float pages
	% N.B.: floatpagefraction MUST be less than topfraction !!
    \renewcommand{\dblfloatpagefraction}{0.7}	% require fuller float pages

\newcommand\jdbcomment[1]{\textcolor{red}{[#1]}}
\newcommand\rancomment[1]{\textcolor{blue}{[#1]}}

\title{Association between SARS-CoV-2 and metagenomic content of samples from the Huanan Seafood Market}

\author[]{\Large Jesse D. Bloom}

\affil[]{Fred Hutchinson Cancer Center; Howard Hughes Medical Institute

}

\keywords{}

\runningtitle{} % For use in the footer
\runningauthor{}

\begin{abstract}
The role of the Huanan Seafood Market in the early SARS-CoV-2 outbreak remains unclear.
Recently the Chinese CDC released data from deep sequencing of environmental samples collected from the market after it was closed on January-1-2020~\citep{liu2023surveillance}.
Prior to this release, \citet{crits2023genetic} analyzed data from a subset of the samples.
Both studies concurred that the samples contained genetic material from a variety of species, including some like raccoon dogs that are susceptible to SARS-CoV-2.
However, neither study systematically analyzed the relationship between the amount of genetic material from SARS-CoV-2 and different animal species.
Here I implement a fully reproducible computational pipeline that jointly analyzes the number of reads mapping to SARS-CoV-2 and the mitochondrial genomes of chordate species across the full set of samples.
I validate the presence of genetic material from numerous species, and calculate mammalian mitochondrial compositions similar to those reported by \citet{crits2023genetic}.
However, the number of SARS-CoV-2 reads is not consistently correlated with reads mapping to any non-human susceptible species.
For instance, 14 samples have $>$20\% of their chordate mitochondrial DNA from raccoon dogs, but only one of these samples contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
Instead, SARS-CoV-2 reads are most correlated with reads mapping to various fish, such as catfish and largemouth bass.
These results suggest that while metagenomic analysis of the environmental samples is useful for identifying animals or animal products sold at the market, co-mingling of animal and viral genetic material is unlikely to reliably indicate whether any animals were infected by SARS-CoV-2.
\end{abstract}

\begin{document}

\maketitle
\thispagestyle{firststyle}
%\marginmark
\firstpagefootnote

\correspondingauthoraffiliation{}{\href{mailto:jbloom@fredhutch.org}{jbloom@fredhutch.org}}
\vspace{-33pt}% Only used for adjusting extra space in the left column of the first page

\lettrine[lines=2]{\color{color2}I}{}nitial reports from Chinese officials about the outbreak that eventually became the SARS-CoV-2 pandemic described patients associated with the Huanan Seafood Market, and said there was no evidence of significant human-to-human transmission~\citep{ProMED2109, healthcommission2019, healthcommission2020, who2020}.
These two claims implied that the infections likely originated from a non-human source within the market.

But by mid to late January of 2020, it was clear SARS-CoV-2 was spreading from human to human, and had been for some time~\citep{chan2020familial,phan2020importation,li2020early,nishiura2020initial}.
In addition, Chinese scientists published papers reporting that some of the earliest identified human cases in December 2019 had no link to the Huanan Seafood Market~\citep{huang2020clinical,chen2020epidemiological}.
Thus began a debate that continues to this day: was the market an initial source of zoonotic infections from animals, or was it simply a superspreading site that amplified earlier human infections from another source~\citep{cohen2020wuhan}?

It is universally agreed that there were human cases of SARS-CoV-2 at the Huanan Seafood Market by mid December of 2019~\citep{li2020early}.
It is also known that some animals susceptible to SARS-CoV-2 were sold at the market~\citep{xiao2021animal}.
The unanswered question is if any animals were infected, and if so whether they infected humans or were infected by them.

In early 2022, the Chinese CDC posted a preprint on \textit{Research Square} describing their sampling of the market beginning immediately after its closure on January-1-2020~\citep{liu2022surveillance}.
The Chinese CDC collected samples from both the environment and animals / animal products.
They reported that none of the 457 animal samples tested positive for SARS-CoV-2, but that 73 of 923 environmental samples tested positive.
A significant caveat of the Chinese CDC study is that all the samples were collected on January-1-2020 or later, which is \emph{at least} a month after the first human infections in Wuhan~\citep{odni2022,zhang2020viral,van2020emergence,pipes2021assessing}.
Despite this significant caveat, the data reported in the Chinese CDC study has been variously interpreted to support arguments that virus in the market was human derived~\citep{liu2022surveillance}, that the outbreak originated from live animals sold in the market~\citep{worobey2022huanan}, or that the virus spread among humans in Mahjong rooms or toilets in the market~\citep{courtier2022sars}.

One aspect of the data gathered by the Chinese CDC that was \emph{not} thoroughly analyzed in their 2022 preprint was the metagenomic content of the environmental samples. 
The 2022 Chinese CDC preprint simply reported that the number of SARS-CoV-2 reads in deep sequencing of the samples was correlated with the number of human reads (fourth figure of \citet{liu2022surveillance}), but did not specify correlations for other species and did not provide the raw sequencing data.
Other scientists pointed out that if the raw data were shared, it would be possible to expand upon the analysis in the 2022 Chinese CDC preprint to determine if the abundance of SARS-CoV-2 genetic material correlated with material from other species~\citep{cohen2022anywhere,cohen2022studies}.

At some point after posting their 2022 preprint, the Chinese CDC uploaded raw sequencing data for some environmental samples to the GISAID database, where they were subsequently downloaded and analyzed by another group of scientists.
News of this analysis leaked to the media, which published stories emphasizing the co-mingling of genetic material from raccoon dogs and SARS-CoV-2 in one of the samples~\citep{wu2023atlantic,mueller2023nytimes}.
The next week, the scientists published their report~\citep{crits2023genetic}, which described bioinformatic analyses showing that some environmental samples contained genetic material from animals susceptible to SARS-CoV-2, including raccoon dogs.
However, although \citet{crits2023genetic} analyzed the mammalian metagenomic content of the partial set of samples they obtained from GISAID, they did not report any analysis of the SARS-CoV-2 content of the samples.

A week later, the Chinese CDC posted an updated version of their preprint on the \textit{ChinaXiv} server~\citep{liu2023surveillance-chinaxiv} and released the full set of raw sequencing data in public databases.
This updated preprint, and a version published by \textit{Nature} the next week~\citep{liu2023surveillance} emphasized different aspects of the data than \citet{crits2023genetic}.
Specifically, although the Chinese CDC concurred that some of the environmental samples contained material from raccoon dogs and other susceptible species, they stated that material from many species was found, and that raccoon dog material was more commonly identified in SARS-CoV-2 negative samples.
However, like \citet{crits2023genetic}, the new version of the Chinese CDC paper~\citep{liu2023surveillance} did not analyze the association between the amount of SARS-CoV-2 and animal material in the samples, and even removed the figure from the original preprint~\citep{liu2022surveillance} that showed that the number of SARS-CoV-2 reads was correlated with reads from both humans and several unidentified species.

Here I systematically analyze the relationship between the metagenomic and SARS-CoV-2 content of all environmental samples for which deep sequencing data are available.
This analysis confirms that samples contain genetic material from a wide range of species, including some that are susceptible to SARS-CoV-2.
However, the samples that contain abundant material from raccoon dogs and other non-human susceptible species generally contain little or no SARS-CoV-2 reads.
Furthermore, the number of SARS-CoV-2 reads is not consistently correlated with reads mapping to any non-human susceptible species, but is instead most correlated with reads from species that are implausible candidates for having been infected with SARS-CoV-2.
These results suggest that SARS-CoV-2 was widespread at the market by January 2020, and therefore that co-mingling of viral and animal genetic material in environmental samples collected at that time is unlikely to be informative about the original source of the outbreak.

\section{Results}

\subsection{Deep sequencing data deposited in public databases on March-29-2023 by the Chinese CDC are a superset of the data analyzed by Crits-Christoph et al earlier that month}
On March-29-2023, the Chinese CDC released raw deep sequencing data for samples taken from the Huanan Market in early 2020.
These data were released on the NGDC database under accession CRA010170 (\url{https://ngdc.cncb.ac.cn/gsa/browse/CRA010170}).

I created a fully reproducible computational pipeline that downloaded and analyzed the full dataset from the NGDC (my pipeline is available at \url{https://github.com/jbloom/Huanan_market_samples}).
The dataset consists of 696 FASTQ files from 395 deep sequencing runs of 176 samples, with the total size of the gzipped files exceeding three terabytes.

Prior to the full posting of the dataset on the NGDC by the Chinese CDC, \citet{crits2023genetic} downloaded and analyzed 227 FASTQ files that had been uploaded to the GISAID database by the Chinese CDC.
When news of the analysis by \citet{crits2023genetic} became public, GISAID reportedly revoked access to the data and stated that the set of files that had been downloaded was incomplete~\citep{cohen2023new,gisaid_statement}.

To determine the relationship between the 696 FASTQ files released by the Chinese CDC on the NGDC on March-29-2023 and the 227 files downloaded earlier that month by \citet{crits2023genetic}, I computed the SHA-512 hashes for all the NGDC FASTQ files and compared them to the hashes reported in Table S1 of \citet{crits2023genetic}.
This analysis confirmed that the dataset released on the NGDC contained unmodified versions of all 227 FASTQ files downloaded by \citet{crits2023genetic}, plus an additional 469 FASTQ files (Tables~\ref{tab:sha526} and \ref{tab:metadata}).

\subsection{Analysis of mitochondrial sequences validates mammalian metagenomics by Crits-Christoph et al, and also finds abundant non-mammalian chordates}
To determine the metagenomic content of the samples, I performed an analysis that conceptually parallels that described by \citet{crits2023genetic}.
Throughout this paper, I report results only for the 169 samples described by the Chinese CDC as ``RNA sequencing of total nucleic acids from environmental swabs for metagenomics,'' and exclude the 4 samples from amplicon-based sequencing of SARS-CoV-2 and the 3 samples that involved RNA sequencing for viral whole-genome assembly (Table~\ref{tab:metadata}).

Briefly, the deep sequencing data were pre-processed to retain only high-quality reads.
These reads were aligned to the SARS-CoV-2 genome and a representative set of chordate mitochondrial genomes.
%The set of mitochondrial genomes was assembled by downloading all NCBI RefSeq genomes, filtering for those from chordates, and removing genomes that were highly similar to other retained genomes (within a Mash distance of $<$0.07).
This set included all the mammalian mitochondrial genomes from all species for which alignments were reported in \citet{crits2023genetic}.
The accessions for the 4,170 chordate mitochondrial genomes that formed this set are listed in Table~\ref{tab:mito_retained}.
The counts of reads mapping to each mitochondrial genome for each sequencing run and sample are provided in Tables~\ref{tab:mito_composition_by_run} and \ref{tab:mito_composition_by_sample}, respectively.

\begin{figure*}
\includegraphics[width=\linewidth]{figures/crits_christoph_vs_current_run_corr.pdf}
\caption{
Correlation of the number of read counts mapping to mitochondrial genomes of mammalian species between the current study and \citet{crits2023genetic} for each sequencing run shared across the studies.
The correlation is calculated only across species for which \citet{crits2023genetic} provide read counts in the third supplementary figure of their report, and only for runs for which both studies find $\ge$50 reads mapping to that set of species.
See \url{https://jbloom.github.io/Huanan_market_samples/crits_christoph_vs_current_run_corr.html} for an interactive version of this plot that enables mouseover of points for details on specific runs, and adjustment of the threshold for how many total aligned mitochondrial reads a run must contain to be shown in this plot.
\label{fig:crits_christoph_vs_current_run_corr}
}
\end{figure*}

I compared the mitochondrial compositions of the samples determined by my analysis to those reported by \citet{crits2023genetic}.
Note that while \citet{crits2023genetic} describe aligning the sequencing data to all metazoa mitochondria, they only report alignment counts for mammals.
The aligned read counts for mammals from my analysis are highly correlated with those reported by \citet{crits2023genetic} for all sequencing runs with a reasonable number of total counts (Figure~\ref{fig:crits_christoph_vs_current_run_corr}).
My analysis and that of \citet{crits2023genetic} are independently implemented, and use different alignment programs, reference genome sets, and quality filtering criteria---so the fact that the two analyses give highly correlated results is reassuring.

Despite the strong correlation between my analysis and that by \citet{crits2023genetic} on a common set of mammalian mitochondrial genomes, it is important to emphasize that the results are dependent on the genomes to which the reads are aligned.
For instance, Figure~\ref{fig:composition} shows the composition of sample Q61 as determined by aligning reads to different genome sets.
My analysis recapitulates the finding of \citet{crits2023genetic} that raccoon dogs make the largest contribution to the mammalian mitochondrial genetic material in this sample---but if the analysis is expanded to all chordates, then duck mitochondrial genetic material is more abundant (Figure~\ref{fig:composition}A).
But raccoon dog is more abundant than duck if the analysis is instead performed by assembling contigs and then aligning to the full genomes of raccoon dogs, ducks, and several other species, in line with the findings of \citet{crits2023genetic}.
These results are a reminder that the measured metagenomic composition is contingent upon the method and reference set used.
Throughout the rest of this paper, I use compositions calculated from reads mapping to mitochondrial genomes rather than contigs mapped to the full genomes for two reasons: (1) this parallels the approach of \citet{crits2023genetic} who only reported mitochondrial composition for samples other than Q61, and (2) a number of relevant species only have mitochondrial but not full genomes available in NCBI.

The interactive figure at \url{https://jbloom.github.io/Huanan_market_samples/mito_composition.html} enables the reader to examine the mitochondrial composition of each sample over both mammals and chordates.

\begin{figure*}
\includegraphics[width=\linewidth]{figures/composition.pdf}
\caption{
Metagenomic composition of sample Q61.
{\bf (A)}
Composition as determined by aligning reads to mitochondrial genomes.
From left to right: composition determined in current study across all chordates, composition determined in current study across all mammals, composition reported in \citet{crits2023genetic} for just mammals shown in the first figure of their report, and composition reported \citet{crits2023genetic} across all mammals in the third supplementary table of their report.
{\bf (B)}
Composition determined in the current study by aligning assembled contigs to the four indicated genomes; this composition is similar to that reported in the third figure of \citet{crits2023genetic}.
See \url{https://jbloom.github.io/Huanan_market_samples/mito_composition.html} for an interactive version of (A) that allows similar pie charts to be viewed for any sample.
See \url{https://jbloom.github.io/Huanan_market_samples/genomic_contig_composition.html} for an interactive version of (B).
\label{fig:composition}
}
\end{figure*}

\subsection{Most environmental samples contain little or no SARS-CoV-2 reads}
To quantify the SARS-CoV-2 content of the samples, I plotted the percentage of all high-quality reads that aligned to SARS-CoV-2 (Figure~\ref{fig:sars2_aligned}, Table~\ref{tab:sars2_aligned_by_run}, Table~\ref{tab:sars2_aligned_by_sample}).
Over three-quarters of the samples had no reads aligning to SARS-CoV-2, and most of the rest only had a small number of viral reads.

Despite the fact that most samples were sequenced to depths that exceeded 100,000,000 reads, only two samples had $>$1,000 reads mapping to SARS-CoV-2 (F100 and B5).
These samples were both collected on January-1-2020 and had a chordate mitochondrial composition dominated by catfish and largemouth bass (see interactive version of Figure~\ref{fig:sars2_aligned} at \url{https://jbloom.github.io/Huanan_market_samples/sars2_aligned_vertical.html}).

\begin{figure*}
\centerline{\includegraphics[width=0.85\linewidth]{figures/sars2_aligned.pdf}}
\caption{
Percentage of high-quality reads that align to SARS-CoV-2 for each sample.
Points are colored according to the date that the sample was collected.
Note the x-axis uses a symlog scale.
See \url{https://jbloom.github.io/Huanan_market_samples/sars2_aligned_vertical.html} for an interactive version of this plot where you can mouseover points for details including the mitochondrial composition of each sample, and select only samples from specific dates or from locations.
\label{fig:sars2_aligned}
}
\end{figure*}

Most samples with the highest SARS-CoV-2 content were collected on the Chinese CDC's first sampling date of January-1-2020, but some samples with SARS-CoV-2 reads were collected on later dates (Figure~\ref{fig:sars2_aligned}).
The non-January-1 sample with the most SARS-CoV-2 was RLC-3, which was collected on February-2-2022 and had chordate mitochondrial composition dominated by rat snake, spotted dove, and human (interactive version of Figure~\ref{fig:sars2_aligned}).

\begin{table}
\caption{
Reads mapping to SARS-CoV-2 out of all high-quality (pre-processed) reads for samples with $\ge$20\% of their chordate mitochondrial composition from a susceptible non-human species as defined in \citet{crits2023genetic}.
Samples with non-zero SARS-CoV-2 reads are in bold.
\label{tab:susceptible}
}
{\scriptsize
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{llrrr}
\toprule
species      &   sample              & \makecell[r]{mitochondrial \\ reads from \\ species} &  \makecell[r]{reads \\ aligning \\ to SARS2} & \makecell[r]{ total pre-\\processed \\ reads} \\ 
\midrule
\multirow{14}{*}{raccoon dog} & HJ200048-20200112-1 &                                80\% &                        0 &                   $1.2\times 10^8$ \\
                 & HJ200050-20200112-1 &                                69\% &                        0 &                     $1.0 \times 10^8$ \\
                 & HJ200017-20200112-1 &                                61\% &                        0 &                   $1.1 \times 10^8$ \\
                 & HJ200023-20200112-1 &                                58\% &                        0 &                   $6.9 \times 10^7$ \\
                 & HJ200011-20200112-1 &                                41\% &                        0 &                   $5.8 \times 10^7$ \\
                 & HJ200012-20200112-1 &                                39\% &                        0 &                   $1.3 \times 10^8$ \\
                 & {\bf Q61} &                                {\bf 32\%} &                        {\bf 1} &                   $\mathbf{2.1 \times 10^8}$ \\
                 & HJ200019-20200112-1 &                                30\% &                        0 &                     $7.0 \times 10^7$ \\
                 & HJ200006-20200112-1 &                                29\% &                        0 &                   $1.3 \times 10^8$ \\
                 & HJ200001-20200112-1 &                                28\% &                        0 &                   $1.2 \times 10^8$ \\
                 & HJ200018-20200112-1 &                                26\% &                        0 &                   $1.4 \times 10^8$ \\
                 & HJ200044-20200112-1 &                                25\% &                        0 &                   $1.2 \times 10^8$ \\
                 & HJ200047-20200112-1 &                                22\% &                        0 &                   $1.4 \times 10^8$ \\
                 & 629-3-C &                                22\% &                        0 &                   $2.5 \times 10^8$ \\
\cline{1-5}
\multirow{6}{*}{hoary bamboo rat} & HJ200065-20200112-1 &                                48\% &                        0 &                   $7.3 \times 10^7$ \\
                 & HJ200062-20200112-1 &                                40\% &                        0 &                   $1.8 \times 10^8$ \\
                 & 629-5-L4 &                                35\% &                        0 &                   $1.4 \times 10^8$ \\
                 & 629-13-L &                                33\% &                        0 &                   $1.5 \times 10^8$ \\
                 & 629-1-L1 &                                30\% &                        0 &                   $2.5 \times 10^8$ \\
                 & HJ200049-20200112-1 &                                23\% &                        0 &                     $1.0 \times 10^8$ \\
\cline{1-5}
\multirow{6}{*}{Amur hedgehog} & W-8-25-L2 &                                56\% &                        0 &                   $3.1 \times 10^8$ \\
                 & HJ200040-20200112-1 &                                51\% &                        0 &                   $1.5 \times 10^8$ \\
                 & HJ200039-20200112-1 &                                30\% &                        0 &                   $1.2 \times 10^8$ \\
                 & {\bf 8-25-M1} &                                {\bf 30\%} &                      {\bf 24} &                   $\mathbf{4.4 \times 10^8}$ \\
                 & HJ200038-20200112-1 &                                23\% &                        0 &                     $1.0 \times 10^8$ \\
                 & W-8-25-D2 &                                22\% &                        0 &                   $3.3 \times 10^8$ \\
\cline{1-5}
Malayan porcupine & {\bf Q70} &                                {\bf 85\%} &                        {\bf 2} &                   $\mathbf{1.5 \times 10^8}$ \\
\cline{1-5}
Himalayan marmot & HJ200005-20200112-1 &                                30\% &                        0 &                   $1.2 \times 10^8$ \\
\bottomrule
\end{tabular}
}
\end{table}

The Q61 sample that was the focus of the raccoon-dog centered media coverage~\citep{wu2023atlantic,mueller2023nytimes} preceding the report by \citet{crits2023genetic} is one of 70 samples collected on January-12-2020 from the west wing of the market.
Six of these samples contained SARS-CoV-2 reads, while the other 64 had no SARS-CoV-2 reads (Figure~\ref{fig:sars2_aligned}).
Sample Q61, which has a mitochondrial metagenomic composition dominated by duck and raccoon dog, had 1 of $\sim2.1 \times 10^8$ high-quality reads mapping to SARS-CoV-2 (interactive version of Figure~\ref{fig:sars2_aligned}).
The samples from January-12-2020 with the most SARS-CoV-2 reads were Q68, Q37, and Q64; species that contributed $\ge$10\% of chordate mitochondrial reads in these samples were chicken, dog, duck, Chinese salamander, rabbit, and various snakes.

 Overall, there were SARS-CoV-2 reads in just 3 of the 28 samples with at least 20\% of their chordate mitochondrial composition from the non-human susceptible species thought to have been sold live in the market (Table~\ref{tab:susceptible}).
There were 14 samples with at least 20\% chordate mitochondrial composition from raccoon dogs, but only sample Q61 contained any SARS-CoV-2 reads (1 of $\sim2.1 \times 10^8$ high-quality reads).
None of the six samples with at least 20\% chordate mitochondrial composition from bamboo rats contained any SARS-CoV-2 reads.
There was one sample each with at least 20\% chordate mitochondrial composition from Amur hedgehog and Malayan porcupine that contained SARS-CoV-2 reads (Table~\ref{tab:susceptible}).

\subsection{Correlations of abundance of SARS-CoV-2 to mitochondrial genetic material from various species}
To more systematically examine the relationship between SARS-CoV-2 and genetic material from different chordates, I calculated the correlation between the number of reads mapping to SARS-CoV-2 versus the mitochondrial genome of each species across all samples.
If the correlation is calculated on a log-log scale as was done in the original Chinese CDC preprint~\citep{liu2022surveillance}, then the five species whose genetic material is most correlated with the abundance of SARS-CoV-2 are (in order) largemouth bass, catfish, cow, carp, and snakehead fish (Figure~\ref{fig:per_species_corr_faceted}).
None of these species are likely hosts for SARS-CoV-2: non-mammals are not thought to be infectable, and cows were probably sold as animal products rather than live animals.
There is a modest correlation between the abundance of SARS-CoV-2 and human mitochondrial material, but this correlation is weaker than for several other species (Figure~\ref{fig:per_species_corr_faceted}).
There is a negative correlation between the abundance of SARS-CoV-2 and mitochondrial material from raccoon dogs and hoary bamboo rats, and at most a very weak positive correlation for other susceptible species thought to have been sold live at the market (Figure~\ref{fig:per_species_corr_faceted}).

Furthermore, the correlations are highly contingent on the sample set and details of how the statistics are calculated.
For instance, if we exclude the first sampling timepoint and only look at samples calculated on January-12-2020 or later, then the most correlated species is the Oriental rat snake, with the correlation for humans becoming even weaker and that for raccoon dogs remaining negative (Figure~\ref{fig:per_species_corr_faceted_later_dates}).
The correlations also change if only consider samples containing a non-zero number of SARS-CoV-2 reads, or calculate correlations on a linear rather than log scale (Figure~\ref{fig:different_corrs}).
The interactive versions of the correlation plots at \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html} and \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_single.html} enable the reader to explore the correlations for different sample subsets and methods of calculating the statistics.
Exploration of these interactive plots shows that it takes a great deal of cherry picking of parameter combinations to put any plausibly infectable hosts at the top of the list of species whose mitochondrial material is most correlated with the abundance of SARS-CoV-2.

\begin{figure*}
\includegraphics[width=\linewidth]{figures/per_species_corr_faceted.png}
\caption{
Correlation between percent of all reads mapping to SARS-CoV-2 and the mitochondrial genome of each of the indicated species.
Each point represents a different environmental sample, and the orange text shows the Pearson correlation.
The scales are log10, and values of zero (which cannot be plotted on a log scale) are shown as half the minimum non-zero value observed across all samples.
See \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html} for an interactive version of this plot that enables mouseover of points for sample details, selection only of samples collected on specific dates or containing at least one SARS-CoV-2 reads, adjustment of scales from log to linear, and adjustment of mitochondrial percent to be of reads mapping to any mitochondria rather than of all reads.
See \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_single.html} for similar plots for individual species.
\label{fig:per_species_corr_faceted}
}
\end{figure*}

\subsection{Fully annotated correlation figure analogous to that in the original Chinese CDC preprint}
The original Chinese CDC preprint from 2022 had in its fourth figure a plot showing the correlation between the amount of SARS-CoV-2 and genetic material from various animal species in the environmental samples~\citep{liu2022surveillance}.
The preprint annotated one point in the plot to indicate there was a correlation between the amount of SARS-CoV-2 and human genetic material.
However, the preprint failed to annotate the identity of any other species with genetic material that correlated with SARS-CoV-2 abundance.
This omission of full species annotations in the correlation plot was widely noted, including in two news articles in \textit{Science} in 2022~\citep{cohen2022anywhere,cohen2022studies}, and a third article in 2023 that re-printed the incompletely annotated plot~\citep{cohen2023unearthed}.
However, neither of the two more recent studies of the metagenomic data have remedied this omission: \citet{crits2023genetic} did not report any analysis of the abundance of SARS-CoV-2 reads, and the published 2023 version of the Chinese CDC preprint simply dropped the plot altogether~\citep{liu2023surveillance}.

\begin{figure}
\includegraphics[width=\linewidth]{figures/overall_corr/overall_corr.pdf}
\caption{
Correlations between SARS-CoV-2 content and mitochondrial content for all species calculated for samples containing at least one SARS-CoV-2 read from any sampling date (left) or just just the January-12-2020 date when most of the wildlife sampling occurred.
This plot is designed to mimic the fourth figure of \citet{liu2022surveillance}, and so only shows samples with at least one SARS-CoV-2 read and calculates correlations between the log number of SARS-CoV-2 reads versus the log number of species mitochondrial reads for consistency with \citet{liu2022surveillance}.
See the interactive version at \url{https://jbloom.github.io/Huanan_market_samples/overall_corr.html} to mouseover all points for details, select different subsets of samples, and calculate the correlations on a linear or log scale.
\label{fig:overall_corr}
}
\end{figure}

To remedy this omission, I used the chordate mitochondrial compositions calculated in the current study to generate plots analogous to that in the original Chinese CDC preprint.
When the correlations are taken across only samples containing at least one SARS-CoV-2 read (as done in the original Chinese CDC preprint), then the genetic material of several non-human species (largemouth bass, catfish, cow, sheep, pig) is more correlated with SARS-CoV-2 reads than is human material (Figure~\ref{fig:overall_corr}).
It is also informative to take the correlations only across samples collected on January-12-2020, which was the date of most intensive sampling of the wildlife stalls (this is when Q61 was collected).
 For January-12-2020 samples, humans are among the species whose genetic material is most correlated with SARS-CoV-2 reads, but goat and spotted dove have roughly equivalent correlations (Figure~\ref{fig:overall_corr}).
Raccoon dog and bamboo rat genetic material are not positively correlated with SARS-CoV-2 reads in either sample set.
See the interactive version of Figure~\ref{fig:overall_corr} at \url{https://jbloom.github.io/Huanan_market_samples/overall_corr.html} to explore other sample subsets.

Note that the correlation plots in Figure~\ref{fig:overall_corr} are not exactly identical to those in the original Chinese CDC preprint.
Since \citet{liu2022surveillance} do not provide sufficiently detailed methods to fully reproduce their analysis, it is impossible to definitively determine the source of the discrepancy---but it is likely due to the inclusion of different genomes in the reference sets used to calculate the metagenomic composition of the samples.

\subsection{The low SARS-CoV-2 read counts raise questions about the consistency of the approach use to call sample positivity}
The Chinese CDC study includes a table that classifies which environmental samples were positive for SARS-CoV-2 (this is first main table in \citet{liu2022surveillance} and the second supplementary table in \citet{liu2023surveillance}).
Subsequent studies that have re-analyzed the Chinese CDC data have re-used their classifications of sample positivity~\citep{worobey2022huanan,courtier2022sars,crits2023genetic}.

\begin{table}
\caption{
Inconsistency in criteria used to classify SARS-CoV-2 positivity in Chinese CDC study, illustrated with four example samples.
There is no consistent rationale for classifying Q61 as positive but E-10-29-2 and A1 as negative: all three were reported negative by RT-qPCR, and A1 was not analyzed by sequencing while the difference in the number of SARS-CoV-2 reads between Q61 and E-10-29-2 is not statistically significant.
RT-qPCR results are from the second supplementary table of \citet{liu2023surveillance} (or equivalently the first table of \citet{liu2022surveillance}).
\label{tab:consistency}
}
{\footnotesize
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{llll}
\toprule
sample & \makecell[l]{RT-qPCR test \\ result (Ct)} & \makecell[l]{sequencing reads \\ mapping to SARS-CoV-2 \\ out of high-quality reads} & \makecell[l]{classification in \\ Chinese CDC study} \\
\midrule
F100 & positive (34.7) & 7,200 out of $2.6 \times 10^8$ & positive \\
Q61 & negative & 1 out of $2.1 \times 10^8$ & positive \\
E-10-29-2 & negative & 0 out of $1.9 \times 10^8$ & negative \\
A1 & negative & not sequenced & negative \\
\bottomrule
\end{tabular}
}
\end{table}

Now that the full data are available, we can examine the criteria used to classify samples as positive or negative for SARS-CoV-2.
It appears that the Chinese CDC classified samples as positive if they met \emph{either} of two criteria: they tested positive for SARS-CoV-2 by RT-qPCR, \emph{or} they were metagenomically sequenced and contained at least one read mapping to SARS-CoV-2.
However, these criteria are not consistent because not all samples were analyzed by both methods, and the differences in the number of SARS-CoV-2 reads in the sequencing data are often not statistically significant between samples classified as positive versus negative.
This inconsistency is illustrated in Table~\ref{tab:consistency} for four example samples: F100 is clearly positive (it both tested positive by RT-qPCR and contained thousands of reads mapping to SARS-CoV-2), but there is no statistical rationale for classifying Q61 as positive but E-10-29-2 and A1 as negative.
See Table~\ref{tab:rt-qpcr} for merged sequencing and RT-qPCR data for all samples that were either sequenced or called as positive by the Chinese CDC.

\section{Discussion}
I have described the first full analysis of the association between the abundance of SARS-CoV-2 and genetic material from different animal species in environmental samples collected by the Chinese CDC from the Huanan Seafood Market.
This analysis expands upon prior work by \citet{liu2023surveillance} and \citet{crits2023genetic} in several ways.

First, I establish that the full set of sequencing data uploaded to public databases by the Chinese CDC at the end of March 2023 contains unmodified versions of all files earlier analyzed by \citet{crits2023genetic}, as well as a large number of additional files.
Therefore, regardless of the disputes about the original access to these files~\citep{cohen2023new,gisaid_statement,crits2023genetic}, all the files earlier posted on GISAID are now available without alteration in public databases.

Second, I validate the finding of both \citet{crits2023genetic} and \citet{liu2023surveillance} that the environmental samples contain genetic material from many species, including humans, various fish, various snakes, cows, goats, pigs, sheep, birds such as ducks and spotted doves, raccoon dogs, bamboo rats, and a long list of other animals.
I am only able to quantitatively compare my analyses of metagenomic compositions to the mammalian compositions reported by \citet{crits2023genetic}, since \citet{liu2023surveillance} provide neither numerical results nor relevant computer code, and \citet{crits2023genetic} only provide numerical results for mammalian species.
For mammalian species, the compositions determined by my analyses are highly correlated with those reported by \citet{crits2023genetic}, which provides a reassuring robustness check on both analyses.
However, it is important to emphasize that metagenomic compositions depend on the methods and reference genomes used, so future analyses using different approaches (e.g., full versus mitochondrial genomes) would likely yield somewhat different results.
I suggest future work should provide full computer code and tabulated results for all species (as done in the current study) to facilitate reproducibility and comparability across studies.

Third, I perform the first comprehensive analysis of the association between the abundance of SARS-CoV-2 and mitochondrial genetic material across all environmental samples.
This analysis reveals that the greatest co-mingling of viral and animal material involves species that were almost certainly not infected by SARS-CoV-2, such as fish (e.g., largemouth bass, catfish) and livestock (e.g., cows, sheep, goats).
Consistent with analyses by the Chinese CDC~\citep{liu2022surveillance}, I find some correlation between the abundance of SARS-CoV-2 and human genetic material, but this correlation is weaker than for several non-infectable animals and so on its own is insufficient to identify the source of the viral material.
Mitochondrial material from most susceptible non-human species sold live at the market is negatively correlated with the presence of SARS-CoV-2: 13 of the 14 samples with at least 20\% of their chordate mitochondrial material from raccoon dogs contain no SARS-CoV-2 reads, and the other sample contains just 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
Likewise, none of the six samples with at least 20\% bamboo rat mitochondrial material contain any SARS-CoV-2 reads.
These findings are compatible with the results in \citet{crits2023genetic}, since that study did not report any analysis of SARS-CoV-2 content.
However, they are somewhat inconsistent with related media articles that emphasized co-mingling of raccoon dog and viral material~\citep{wu2023atlantic,mueller2023nytimes}---in fact, raccoon dogs are one of the species with the least co-mingling of their genetic material and SARS-CoV-2.
The basic finding that SARS-CoV-2 material is not associated with material from susceptible non-human species sold live at the market is largely robust to examining subsets of the samples (such as just those collected on specific dates), and I provide interactive plots to facilitate visualizing the data in different ways.

Fourth, I create a version of the SARS-CoV-2 versus species correlation plot in the fourth figure of the original Chinese CDC preprint~\citep{liu2022surveillance}, but with full annotation of all species.
The identity of the non-human species correlated with SARS-CoV-2 content had previously been the subject of speculation~\citep{cohen2022anywhere, cohen2022studies, cohen2023unearthed}.
It turns out these species are fish and livestock that are unlikely candidates for having been infected with SARS-CoV-2.
Even if the plot is restricted to just samples collected on the date of most intense sampling of wildlife stalls (January-12-2020), no non-human susceptible species is among those most correlated with SARS-CoV-2 content.

Fifth, my analysis calls into question the consistency of the criteria used to classify environmental samples as positive versus negative.
For instance, sample Q61 became the subject of widespread media coverage~\citep{wu2023atlantic, mueller2023nytimes} because it contains raccoon dog genetic material and was classified as SARS-CoV-2 positive by the Chinese CDC.
However, this sample tested negative by RT-qPCR and appears to have been called positive on the basis of containing 1 of $\sim$200,000,000 reads that mapped to SARS-CoV-2.
But all environmental samples contain a mix of genetic material from numerous sources, so it is not consistent to classify this particular sample as positive when hundreds of other samples that also tested negative by RT-qPCR are classified as negative because they were never sequenced or had SARS-CoV-2 content statistically indistinguishable from 1 of $\sim$200,000,000 reads.
I suggest that future work analyzing the spatial distribution of SARS-CoV-2 across the Huanan Seafood Market~\citep{worobey2022huanan,courtier2022sars,crits2023genetic} consider the quantitative content of samples (such as determined from deep sequencing in the current study, or the Ct values in \citet{liu2023surveillance}), and only include negative samples subjected to the full set of measurements used to classify other samples as positive.

Overall my study validates the approach of \citet{crits2023genetic} of using metagenomic analysis of environmental samples to identify animals and animal products sold at the Huanan Seafood Market.
As noted by \citet{crits2023genetic}, this innovative approach could usefully inform tracing of animals supplied to the market.
However, the results described here suggest that the utility of these metagenomic analyses does not extend to indicating whether animals at the market were actually infected by virus.
For instance, the presence of 1 in $\sim$200,000,000 sequencing reads mapping to SARS-CoV-2 in a single sample containing raccoon dog genetic material does not suggest raccoon dogs were infected, given that material from many other species including fish is far more consistently co-mingled with SARS-CoV-2.

When considered in larger context, the inability of the environmental samples to inform on the origins of the virus is unsurprising.
These samples were all collected on January-1-2020 or later, which is at least several weeks after the Huanan Seafood Market became a superspreading site for human infections~\citep{li2020early}.
Therefore, by the time the samples were collected, SARS-CoV-2 had been spread widely across the market by humans regardless of its original source---as evidenced by the results reported here, which show viral genetic material coincident with material from myriad animals ranging from fish to snakes to mammals. 
The first human infections with SARS-CoV-2 in Wuhan occurred no later than November of 2019~\citep{odni2022,zhang2020viral,van2020emergence,pipes2021assessing}, which is over a month before the Chinese CDC reports that it began to collect samples from the market.
For this reason, further insight into the origins and early spread of SARS-CoV-2 will likely require learning more about events or cases that occurred no later than November or early December of 2019.

\section{Limitations of this study}
This study has limitations related to the data, methodology, and samples themselves.

For the data, all sequencing files and related annotations derive from information shared by the Chinese CDC.
The description of how samples were processed prior to sequencing lacks detail: for instance, \citet{liu2023surveillance} say ``human nucleic acid was removed'' but not precisely how this was done.
Additionally, the data were released after China's State Council ordered in March of 2020 that all publications and information related to COVID-19 be reviewed by a centralized task force~\citep{kang2020china}.
It is possible that this centralized Chinese government review influenced which data were released~\citep{hvistendahl2023}.

For the methodology, the metagenomic compositions were calculated by alignment to a reference set of mitochondrial genomes from a large but still incomplete set of species.
It is possible that some species that deposited genetic material do not have exact matches in this reference set, so slightly different results might be obtained if a different reference set was used.
Similarly, this study analyzed chordate mitochondrial genomes: different results would be obtained if the reference genome set was expanded (e.g., to all metazoa) or shrunk (e.g., only to mammals), or if the reads were aligned to full rather than mitochondrial genomes.
In addition, the results might slightly change with different parameters for read alignment, quality filtering, etc.
To ensure transparency of the methodology in this study, I have provided a fully reproducible computational pipeline (\url{https://github.com/jbloom/Huanan_market_samples}).
In addition, I have provided interactive plots of the results to help the reader explore the effects of different parameter choices on the final results (\url{https://jbloom.github.io/Huanan_market_samples/}).

The major limitation of the samples is that they were all collected on January-1-2020 or later, which is well after the first human SARS-CoV-2 infections in Wuhan.
The lateness of the sampling relative to the origin of the outbreak limits the conclusions that can be drawn.

{\small

\section{Methods}
\subsection{Code and data availability}
See \url{https://github.com/jbloom/Huanan_market_samples} for a GitHub repository containing a fully reproducible Snakemake~\citep{molder2021sustainable} computational pipeline implementing the analysis described in this paper, starting with downloading of the raw data from NGDC and proceeding all the way through to rendering of the plots shown in the figures.
That GitHub repository also includes files containing key numerical results.
See \url{https://jbloom.github.io/Huanan_market_samples/} for interactive plots of the results rendered using Altair~\citep{vanderplas2018altair}.

\subsection{Processing and alignment of deep sequencing data}
The FASTQ files with the raw deep sequencing data for all samples were downloaded from the NGDC project CRA010170 (\url{https://ngdc.cncb.ac.cn/gsa/browse/CRA010170}).
The FASTQ files were pre-processed with \texttt{fastp}~\citep{chen2018fastp} to remove low-quality reads; all tabulations of number of reads in this paper refer to the high-quality reads that passed this pre-processing step.

To create an mitochondrial genome alignment reference set, I followed a procedure partially analogous to that described by \citet{crits2023genetic}.
All mitochondrial genomes in the RefSeq database were downloaded, and then filtered to retain only genomes from the phylum Chordata (in contrast, \citet{crits2023genetic} describe retaining all metazoa mitochondrial genomes).
The raccoon dog mitochondrial genome was not present in this set, so it was separately downloaded from Genbank accession KX964606.
All the mitochondrial genomes were then filtered to remove highly similar ones.
To do this, \texttt{Mash}~\citep{ondov2016mash} was used to compute the mash distances between all pairs of mitochondrial genomes.
To ensure the most relevant mitochondrial genomes were retained, I first specified for manual retention the genomes for all the species listed in the third supplementary file of \citet{crits2023genetic}, as well as genomes for some additional relevant species (these species are listed under the \textit{mitochondrial\_genomes\_to\_keep} key in \url{https://github.com/jbloom/Huanan_market_samples/blob/main/config.yaml}).
After retaining all of these manually specified mitochondrial genomes, the pipeline then greedily iterated through all other genomes choosing for retention each genome that was not within a mash distance of 0.07 of another already retained genome.
The full set of retained mitochondrial genomes is listed in Table~\ref{tab:mito_retained}.

The retained mitochondrial genomes were then concatenated, and added to the SARS-CoV-2 reference genome (Genbank accession NC\_045512v2) with its polyA tail trimmed to avoid spurious alignments.
This concatenation of the SARS-CoV-2 and mitochondrial genomes served as the reference for all sequencing read alignment.

The sequencing reads were then aligned to this genome using \texttt{minimap2}~\citep{li2018minimap2} in the \texttt{sr} (short-read) mode.
The resulting alignments were filtered to only retain primary alignments with mapping quality of at least 4.

I then used \texttt{CoverM} (\url{https://github.com/wwood/CoverM}) to compute the number of aligned reads for each run, requiring alignment lengths of at least 40 nucleotides with at least 95\% identity, and excluding the 100 nucleotides at the contig ends.
The resulting statistics on alignment counts to SARS-CoV-2 and the mitochondrial genomes for each sequencing run are in Tables~\ref{tab:mito_composition_by_run} and \ref{tab:sars2_aligned_by_run}.
To get per-sample alignment counts, I aggregated counts for all runs for each sample to get the statistics in Tables~\ref{tab:mito_composition_by_sample} and \ref{tab:sars2_aligned_by_sample}.
For the mitochondrial genome compositions and all other analyses in this paper, I only included only the samples from metagenomic sequencing, which are annotated with the description ``RNA sequencing of total nucleic acids from environmental swabs for metagenomics'' in the metadata provided by \citet{liu2023surveillance}.
Further, compositions were only computed for runs with at least 500 reads aligning across all mitochondrial, and species were only retained if they had at least 20\% of all aligned mitochondrial reads and at least 4000 covered based for at least one run.

\subsection{Assembling and aligning contigs for sample Q61}
For sample Q61, contigs were also assembled and aligned to full genomes to facilitate comparison to \citet{crits2023genetic}, who also performed a similar analysis only for this sample.
The contigs were assembled using \texttt{Trinity}~\citep{grabherr2011trinity}, and then aligned to using \texttt{minimap2}~\citep{li2018minimap2} to concatenated full genomes for chicken, dog, raccoon dog, and duck.
Contig alignments were only reported if they had a mapping quality of at least 10, an alignment length of at least 300 nucleotides, and an identity of at least 98\%.

\section{Acknowledgments}
I thank \citet{liu2023surveillance} for publicly sharing via the NGDC database the deep sequencing data analyzed in this study.
This study utilized the Fred Hutch Scientific Computing infrastructure, which is supported in part by NIH grants S10-OD-020069 and S10-OD-028685.
The author is an Investigator of the Howard Hughes Medical Institute.

\section{Competing interests}
JDB is on the scientific advisory boards of Apriori Bio, Aerium Therapeutics, Invivyd, the Vaccine Company, and Oncorus; consults for GSK; and receives royalty payments as an inventor on Fred Hutch licensed patents related to deep mutational scanning of viral proteins.

\bibliography{references}
}

\newgeometry{left=1in, right=1in, top=48pt, bottom=60pt}
\onecolumn
\renewcommand{\thepage}{S\arabic{page}}
\setcounter{page}{1}
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}

\clearpage

\begin{figure*}
\includegraphics[width=\linewidth]{figures/per_species_corr_faceted_later_dates.png}
\caption{
Correlation between percent of all reads mapping to SARS-CoV-2 and the mitochondrial genome of each of the indicated species only among samples collected after the first (Jan-1-2020) sampling date.
Except for the fact that this figure excludes samples from the first date, this figure is otherwise the same as Figure~\ref{fig:per_species_corr_faceted}.
See \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html} for an interactive version of this plot with numerous options, including allowing further refinement of the sample date range.
\label{fig:per_species_corr_faceted_later_dates}
}
\end{figure*}

\clearpage

\begin{figure*}
\includegraphics[width=\linewidth]{figures/different_corrs/correlations.pdf}
\caption{
The correlations between the amount of SARS-CoV-2 and mitochondrial genetic material in the samples depends on how the correlation is calculated.
The top row shows correlations for human, and the bottom row shows correlations for raccoon dog.
Each column shows the correlations calculated different ways: with a log or linear scale, calculating the mitochondrial percent to be of all reads or just reads mapping to the mitochondria, and showing all samples or only those with at least one SARS-CoV-2 read.
See \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_single.html} for interactive versions of the plot panels.
\label{fig:different_corrs}
}
\end{figure*}

\clearpage

\begin{table}
\caption{
See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/crits_christoph_data/check_sha512_vs_crits_christoph.csv} for the correspondence between the FASTQ files analyzed by \citet{crits2023genetic} and the files deposited in the NGDC by \citet{liu2023surveillance}.
The files are matched based on their SHA-526 hashes, which were computed directly for the NGDC files as part of the current study, and were taken from Table S1 of \citet{crits2023genetic} for that study.
\label{tab:sha526}
}
\end{table}

\begin{table}
\caption{
See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/metadata/merged_metadata.csv} for the metadata for all of the samples and files uploaded to the NGDC by \citet{liu2023surveillance} under accession CRA010170.
\label{tab:metadata}
}
\end{table}

\begin{table}
\caption{
See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/mitochondrial_genomes/retained.csv} for the NCBI accessions of the set of chordate mitochondrial genomes that was used as the alignment reference for the metagenomic analysis.
\label{tab:mito_retained}
}
\end{table}

\begin{table}
\caption{
See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/aggregated_counts/mito_composition_by_run.csv} for the counts of reads aligning to each mitochondrial genome for each sequencing run.
\label{tab:mito_composition_by_run}
}
\end{table}

\begin{table}
\caption{
See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/aggregated_counts/mito_composition_by_sample.csv} for the counts of reads aligning to each mitochondrial genome for each sample.
\label{tab:mito_composition_by_sample}
}
\end{table}

\begin{table}
\caption{
See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/aggregated_counts/sars2_aligned_by_run.csv} for the counts of reads aligning to SARS-CoV-2 for each sequencing run.
\label{tab:sars2_aligned_by_run}
}
\end{table}

\begin{table}
\caption{
See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/aggregated_counts/sars2_aligned_by_sample.csv} for the counts of reads aligning to SARS-CoV-2 for each sequencing sample.
\label{tab:sars2_aligned_by_sample}
}
\end{table}

\begin{table}
\caption{
See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/rt_qpcr/rt_qpcr.csv} for combined RT-qPCR data and number of SARS-CoV-2 reads for all environmental samples that were either RT-qPCR positive or metagenomically sequenced.
This table was constructed by merging the results for all sequenced samples with all samples listed in the second supplementary table of \citet{liu2023surveillance} (or equivalently the first main table of \citet{liu2022surveillance}), which lists all ``positive samples.''
Since the Chinese CDC says all environmental samples were tested by RT-qPCR, samples not listed as RT-qPCR positive are presumed to have tested negative by that assay.
Note that this table does \emph{not} list samples that tested negative by RT-qPCR and were not sequenced.
\label{tab:rt-qpcr}
}
\end{table}

\end{document}

### 2023-04-25 paper/paper.tex
tweak abstract

@@
 However, the number of SARS-CoV-2 reads is not consistently correlated with reads mapping to any non-human susceptible species.
-For instance, 14 samples have $>$20\% of their chordate mitochondrial DNA from raccoon dogs, but only one of these samples contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
+For instance, 14 samples have $>$20\% of their chordate mitochondrial material from raccoon dogs, but only one of these samples contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
 Instead, SARS-CoV-2 reads are most correlated with reads mapping to various fish, such as catfish and largemouth bass.

### 2023-05-02 scripts/process_metadata.py
some minor fixes (samples sequenced 2 ways, more consistent `fastp` preprocessing) (#1)
* trim polyG / polyX from paired FASTQ

* for samples sequenced 2 ways, retain metagenomic

Previously, samples that were sequenced both metagenomically and for virus assembly (like A20) were filtered. Now the pipeline retains the metagenomic data for these samples and just ignores the virus assembly data. This adds a few more samples.

This commit also includes the new results after re-running the pipeline after the bug fix to apply consistent `fastp` processing to the paired-end and single-end reads.

Nei

@@ -25,13 +25,18 @@ assert set(samples.columns).intersection(experiments.columns) == {"Accession", "
 
 metadata = (
     samples
-    .drop(columns="ID")
+    .drop(columns=["ID", "Public description"])
     .rename(columns={"Accession": "BioSample accession"})
     .merge(
         (
             experiments
             .drop(columns="ID")
-            .rename(columns={"Accession": "Experiment accession"})
+            .rename(
+                columns={
+                    "Accession": "Experiment accession",
+                    "Library Construction / Experimental Design": "description",
+                },
+            )
         ),
         on="BioSample accession",
         validate="one_to_many",
@@ -94,7 +99,7 @@ metadata = (
 first_cols = [
     "Sample name",
     "Sample title",
-    "Public description",
+    "description",
     "Collection date",
     "Run accession",
     "BioSample accession",

### 2023-05-02 paper/paper.tex
add new tables

@@
 Samples with non-zero SARS-CoV-2 reads are in bold.
+See Table~\ref{tab:susceptible_mammal} for a similar table that shows mammalian rather than chordate composition.
+This table uses a 20\% cutoff due to space considerations; to see similar data tabulated for all samples with no cutoff, see the much larger Table~\ref{tab:raccoon_dog_long} (for raccoon dog) and Table~\ref{tab:mito_composition_by_sample} (for all species).
 \label{tab:susceptible}
@@
 \toprule
-species      &   sample              & \makecell[r]{mitochondrial \\ reads from \\ species} &  \makecell[r]{reads \\ aligning \\ to SARS2} & \makecell[r]{ total pre-\\processed \\ reads} \\ 
+species      &   sample              & \makecell[r]{chordate \\ mitochondrial \\ reads from \\ species} &  \makecell[r]{reads \\ aligning \\ to SARS2} & \makecell[r]{ total pre-\\processed \\ reads} \\
 \midrule
@@
 There was one sample each with at least 20\% chordate mitochondrial composition from Amur hedgehog and Malayan porcupine that contained SARS-CoV-2 reads (Table~\ref{tab:susceptible}).
+If mitochondrial composition is instead analyzed only among mammals rather than chordates, Q61 is still the only sample with at least 20\% raccoon dog composition that contains any SARS-CoV-2 reads (Table~\ref{tab:susceptible_mammal}).
+There is one sample with at least 20\% of its mammalian (but not chordate) mitochondrial composition from bamboo rat that contains a small number of SARS-CoV-2 reads (Table~\ref{tab:susceptible_mammal}).
+
+Note that the 20\% cutoff applied in Tables~\ref{tab:susceptible} and \ref{tab:susceptible_mammal} is not integral to the analysis, but is just a way to subset on samples containing the most material from the species of interest to make the tables small enough to easily visualize.
+See Tables~\ref{tab:raccoon_dog_long} and \ref{tab:mito_composition_by_sample} for much larger tables that show comparable data for all samples with respect to the raccoon dog or overall mitochondrial composition, respectively.
+For instance, Table~\ref{tab:raccoon_dog_long} shows that there are a few samples with lower raccoon dog mitochondrial composition that also contain some SARS-CoV-2 reads.
+However, ultimately these very large tables are difficult to visualize, and so the relationship between SARS-CoV-2 and mitochondrial content for each species for all samples is probably more facilely visualized using the interactive versions of the scatter plots described in the next subsection.
 
@@
 \caption{
+Reads mapping to SARS-CoV-2 out of all high-quality (pre-processed) reads for samples with $\ge$20\% of their mammalian mitochondrial composition from a susceptible non-human species as defined in \citet{crits2023genetic}.
+Samples with non-zero SARS-CoV-2 reads are in bold.
+See Table~\ref{tab:susceptible} for a similar table that shows chordate rather than mammalian composition.
+This table uses a 20\% cutoff due to space considerations; to see similar data tabulated for all samples with no cutoff, see the much larger Table~\ref{tab:raccoon_dog_long} (for raccoon dog) and Table~\ref{tab:mito_composition_by_sample} (for all species).
+\label{tab:susceptible_mammal}
+}
+{\scriptsize
+\setlength{\tabcolsep}{5pt}
+\renewcommand{\arraystretch}{1.3}
+\begin{tabular}{llrrr}
+\toprule
+species      &   sample              & \makecell[r]{mammalian \\ mitochondrial \\ reads from \\ species} &  \makecell[r]{reads \\ aligning \\ to SARS2} & \makecell[r]{ total pre-\\processed \\ reads} \\
+\midrule
+\multirow{28}{*}{raccoon dog} & HJ200050-20200112-1 &                                       94\% &                        0 &                     1e+08 \\
+                  & HJ200023-20200112-1 &                                       93\% &                        0 &                   6.9e+07 \\
+                  & HJ200048-20200112-1 &                                       92\% &                        0 &                   1.2e+08 \\
+                  & HJ200019-20200112-1 &                                       90\% &                        0 &                     7e+07 \\
+                  & {\bf Q61} &                                       {\bf 81\%} &                        {\bf 1} &                   {\bf 2.1e+08} \\
+                  & HJ200017-20200112-1 &                                       81\% &                        0 &                   1.1e+08 \\
+                  & 629-3-C &                                       78\% &                        0 &                   2.5e+08 \\
+                  & HJ200044-20200112-1 &                                       76\% &                        0 &                   1.2e+08 \\
+                  & HJ200014-20200112-1 &                                       70\% &                        0 &                   1.1e+08 \\
+                  & HJ200047-20200112-1 &                                       66\% &                        0 &                   1.4e+08 \\
+                  & HJ200011-20200112-1 &                                       61\% &                        0 &                   5.8e+07 \\
+                  & HJ200018-20200112-1 &                                       53\% &                        0 &                   1.4e+08 \\
+                  & HJ200013-20200112-1 &                                       50\% &                        0 &                   1.1e+08 \\
+                  & HJ200012-20200112-1 &                                       50\% &                        0 &                   1.3e+08 \\
+                  & HJ200043-20200112-1 &                                       48\% &                        0 &                   8.3e+07 \\
+                  & HJ200027-20200112-1 &                                       43\% &                        0 &                     9e+07 \\
+                  & HJ200024-20200112-1 &                                       42\% &                        0 &                   9.8e+07 \\
+                  & HJ200001-20200112-1 &                                       38\% &                        0 &                   1.2e+08 \\
+                  & HJ200016-20200112-1 &                                       33\% &                        0 &                   1.2e+08 \\
+                  & HJ200007-20200112-1 &                                       31\% &                        0 &                   1.4e+08 \\
+                  & HJ200020-20200112-1 &                                       31\% &                        0 &                   7.9e+07 \\
+                  & HJ200006-20200112-1 &                                       30\% &                        0 &                   1.3e+08 \\
+                  & HJ200009-20200112-1 &                                       27\% &                        0 &                   1.5e+08 \\
+                  & HJ200045-20200112-1 &                                       27\% &                        0 &                   1.1e+08 \\
+                  & HJ200026-20200112-1 &                                       25\% &                        0 &                     1e+08 \\
+                  & HJ200022-20200112-1 &                                       23\% &                        0 &                   1.2e+08 \\
+                  & HJ200004-20200112-1 &                                       23\% &                        0 &                   2.3e+08 \\
+                  & HJ200065-20200112-1 &                                       22\% &                        0 &                   7.3e+07 \\
+\cline{1-5}
+\multirow{14}{*}{hoary bamboo rat} & 629-13-L &                                       85\% &                        0 &                   1.5e+08 \\
+                  & HJ200041-20200112-1 &                                       56\% &                        0 &                   1.8e+08 \\
+                  & HJ200015-20200112-1 &                                       56\% &                        0 &                   1.4e+08 \\
+                  & HJ200065-20200112-1 &                                       56\% &                        0 &                   7.3e+07 \\
+                  & 629-1-L1 &                                       55\% &                        0 &                   2.5e+08 \\
+                  & 629-5-L4 &                                       54\% &                        0 &                   1.4e+08 \\
+                  & HJ200062-20200112-1 &                                       42\% &                        0 &                   1.8e+08 \\
+                  & HJ200013-20200112-1 &                                       37\% &                        0 &                   1.1e+08 \\
+                  & HJ200049-20200112-1 &                                       31\% &                        0 &                     1e+08 \\
+                  & HJ200046-20200112-1 &                                       28\% &                        0 &                   1.1e+08 \\
+                  & HJ200030-20200112-1 &                                       28\% &                        0 &                   1.1e+08 \\
+                  & HJ200067-20200112-1 &                                       27\% &                        0 &                   7.7e+07 \\
+                  & HJ200025-20200112-1 &                                       21\% &                        0 &                   1.2e+08 \\
+                  & {\bf Q68} &                                       {\bf 20\%} &                        {\bf 6} &                   {\bf 8.7e+07} \\
+\cline{1-5}
+\multirow{15}{*}{Amur hedgehog} & W-8-25-L2 &                                       76\% &                        0 &                   3.1e+08 \\
+                  & HJ200040-20200112-1 &                                       75\% &                        0 &                   1.5e+08 \\
+                  & HJ200038-20200112-1 &                                       63\% &                        0 &                     1e+08 \\
+                  & HJ200039-20200112-1 &                                       62\% &                        0 &                   1.2e+08 \\
+                  & 8-25-CK &                                       62\% &                        0 &                   1.5e+09 \\
+                  & W-8-25-L &                                       59\% &                        0 &                   2.9e+08 \\
+                  & HJ200035-20200112-1 &                                       59\% &                        0 &                   7.2e+07 \\
+                  & {\bf 8-25-M1} &                                      {\bf 53\%} &                      {\bf 24} &                   {\bf 4.4e+08} \\
+                  & HJ200033-20200112-1 &                                       51\% &                        0 &                   1.2e+08 \\
+                  & W-8-25-D2 &                                       50\% &                        0 &                   3.3e+08 \\
+                  & W-8-25-D1 &                                       50\% &                        0 &                   2.9e+08 \\
+                  & HJ200045-20200112-1 &                                       41\% &                        0 &                   1.1e+08 \\
+                  & HJ200032-20200112-1 &                                       35\% &                        0 &                   1.4e+08 \\
+                  & 8-25-M2 &                                       24\% &                        0 &                   1.6e+09 \\
+                  & HJ200036-20200112-1 &                                       21\% &                        0 &                   2.2e+08 \\
+\cline{1-5}
+\multirow{2}{*}{Malayan porcupine} & {\bf Q70} &                                      {\bf 93\%} &                       {\bf 2} &                  {\bf 1.5e+08} \\
+                  & HJ200001-20200112-1 &                                       23\% &                        0 &                   1.2e+08 \\
+\cline{1-5}
+Himalayan marmot & HJ200005-20200112-1 &                                       36\% &                        0 &                   1.2e+08 \\
+\cline{1-5}
+masked palm civet & HJ200042-20200112-1 &                                       22\% &                        0 &                   1.9e+08 \\
+\bottomrule
+\end{tabular}
+}
+\end{table}
+
+\begin{table}
+\caption{
+See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/plots/raccoon_dog_long.csv} for a table giving the raccoon dog mitochondrial composition (both among chordates and mammals) along with the SARS-CoV-2 content for all samples.
+\label{tab:raccoon_dog_long}
+}
+\end{table}
+
+\begin{table}
+\caption{
 See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/rt_qpcr/rt_qpcr.csv} for combined RT-qPCR data and number of SARS-CoV-2 reads for all environmental samples that were either RT-qPCR positive or metagenomically sequenced.

### 2023-05-02 paper/paper.tex
some clarifications to paper text

@@
 To determine the metagenomic content of the samples, I performed an analysis that conceptually parallels that described by \citet{crits2023genetic}.
-Throughout this paper, I report results only for the 169 samples described by the Chinese CDC as ``RNA sequencing of total nucleic acids from environmental swabs for metagenomics,'' and exclude the 4 samples from amplicon-based sequencing of SARS-CoV-2 and the 3 samples that involved RNA sequencing for viral whole-genome assembly (Table~\ref{tab:metadata}).
+Throughout this paper, I report results only for sequencing described by the Chinese CDC as ``RNA sequencing of total nucleic acids from environmental swabs for metagenomics,'' and exclude the small number of samples that involve amplicon-based sequencing of SARS-CoV-2 or RNA sequencing for viral whole-genome assembly (Table~\ref{tab:metadata}).
 
@@
 See \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_single.html} for similar plots for individual species.
+The plots shown here include only samples with at least 200 aligned mitochondrial reads; that option can be adjusted in the interactive plots.
 \label{fig:per_species_corr_faceted}
@@
 For the mitochondrial genome compositions and all other analyses in this paper, I only included only the samples from metagenomic sequencing, which are annotated with the description ``RNA sequencing of total nucleic acids from environmental swabs for metagenomics'' in the metadata provided by \citet{liu2023surveillance}.
-Further, compositions were only computed for runs with at least 500 reads aligning across all mitochondrial, and species were only retained if they had at least 20\% of all aligned mitochondrial reads and at least 4000 covered based for at least one run.
+Species were only retained in the annotated reference set if they had at least 20\% of all aligned mitochondrial reads and at least 4000 covered based for at least one run, or if they are one of the species specified under the \textit{mitochondrial\_genomes\_to\_keep} key in \url{https://github.com/jbloom/Huanan_market_samples/blob/main/config.yaml}.
+The species specified under that key include all susceptible species studied in \citet{crits2023genetic}, so this criteria does not lead too dropping of any of the species in their study.
 
@@
 See \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html} for an interactive version of this plot with numerous options, including allowing further refinement of the sample date range.
+The plots shown here include only samples with at least 200 aligned mitochondrial reads; that option can be adjusted in the interactive plots.
 \label{fig:per_species_corr_faceted_later_dates}
@@
 See \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_single.html} for interactive versions of the plot panels.
+The plots shown here include only samples with at least 200 aligned mitochondrial reads; that option can be adjusted in the interactive plots.
 \label{fig:different_corrs}

### 2023-05-03 paper/paper.tex
update `overall_corr` figure

@@
 I validate the presence of genetic material from numerous species, and calculate mammalian mitochondrial compositions similar to those reported by \citet{crits2023genetic}.
-However, the number of SARS-CoV-2 reads is not consistently correlated with reads mapping to any non-human susceptible species.
+However, the number of SARS-CoV-2 reads is not consistently correlated with reads mapping to non-human susceptible species.
 For instance, 14 samples have $>$20\% of their chordate mitochondrial material from raccoon dogs, but only one of these samples contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
@@
 This analysis confirms that samples contain genetic material from a wide range of species, including some that are susceptible to SARS-CoV-2.
-However, the samples that contain abundant material from raccoon dogs and other non-human susceptible species generally contain little or no SARS-CoV-2 reads.
-Furthermore, the number of SARS-CoV-2 reads is not consistently correlated with reads mapping to any non-human susceptible species, but is instead most correlated with reads from species that are implausible candidates for having been infected with SARS-CoV-2.
+However, the samples that contain abundant material from raccoon dogs and other non-human susceptible species usually contain little or no SARS-CoV-2 reads.
+Furthermore, the number of SARS-CoV-2 reads is not consistently correlated with reads mapping to non-human susceptible species, but is instead generally most correlated with reads from species that are implausible candidates for having been infected with SARS-CoV-2.
 These results suggest that SARS-CoV-2 was widespread at the market by January 2020, and therefore that co-mingling of viral and animal genetic material in environmental samples collected at that time is unlikely to be informative about the original source of the outbreak.
@@
 There is a modest correlation between the abundance of SARS-CoV-2 and human mitochondrial material, but this correlation is weaker than for several other species (Figure~\ref{fig:per_species_corr_faceted}).
-There is a negative correlation between the abundance of SARS-CoV-2 and mitochondrial material from raccoon dogs and hoary bamboo rats, and at most a very weak positive correlation for other susceptible species thought to have been sold live at the market (Figure~\ref{fig:per_species_corr_faceted}).
+There is a negative correlation between the abundance of SARS-CoV-2 and mitochondrial material from raccoon dogs and hoary bamboo rats, and at most a weak positive correlation for other susceptible species thought to have been sold live at the market (Figure~\ref{fig:per_species_corr_faceted}).
 
@@
 This omission of full species annotations in the correlation plot was widely noted, including in two news articles in \textit{Science} in 2022~\citep{cohen2022anywhere,cohen2022studies}, and a third article in 2023 that re-printed the incompletely annotated plot~\citep{cohen2023unearthed}.
-However, neither of the two more recent studies of the metagenomic data have remedied this omission: \citet{crits2023genetic} did not report any analysis of the abundance of SARS-CoV-2 reads, and the published 2023 version of the Chinese CDC preprint simply dropped the plot altogether~\citep{liu2023surveillance}.
+However, neither of the two more recent studies of the metagenomic data have addressed this omission: \citet{crits2023genetic} did not report any analysis of the abundance of SARS-CoV-2 reads, and the published 2023 version of the Chinese CDC preprint simply dropped the plot altogether~\citep{liu2023surveillance}.
 
@@
 \caption{
-Correlations between SARS-CoV-2 content and mitochondrial content for all species calculated for samples containing at least one SARS-CoV-2 read from any sampling date (left) or just just the January-12-2020 date when most of the wildlife sampling occurred.
-This plot is designed to mimic the fourth figure of \citet{liu2022surveillance}, and so only shows samples with at least one SARS-CoV-2 read and calculates correlations between the log number of SARS-CoV-2 reads versus the log number of species mitochondrial reads for consistency with \citet{liu2022surveillance}.
-See the interactive version at \url{https://jbloom.github.io/Huanan_market_samples/overall_corr.html} to mouseover all points for details, select different subsets of samples, and calculate the correlations on a linear or log scale.
+Correlations between SARS-CoV-2 content and mitochondrial content for all species.
+Top row show just samples containing at least one SARS-CoV-2 read and bottom row shows samples regardless of SARS-CoV-2 content; left shows samples from all sampling dates and right shows just samples from the January-12-2020 date when most of the wildlife sampling occurred.
+This plot is designed to mimic the fourth figure of \citet{liu2022surveillance}.
+Some key species are labeled; see the interactive version at \url{https://jbloom.github.io/Huanan_market_samples/overall_corr.html} to mouseover all points for details, select different subsets of samples, and calculate the correlations on a linear or log scale.
 \label{fig:overall_corr}
@@
 To remedy this omission, I used the chordate mitochondrial compositions calculated in the current study to generate plots analogous to that in the original Chinese CDC preprint.
-When the correlations are taken across only samples containing at least one SARS-CoV-2 read (as done in the original Chinese CDC preprint), then the genetic material of several non-human species (largemouth bass, catfish, cow, sheep, pig) is more correlated with SARS-CoV-2 reads than is human material (Figure~\ref{fig:overall_corr}).
+When the correlations are taken across all sampling dates only for samples containing at least one SARS-CoV-2 read (as done in the original Chinese CDC preprint), then the genetic material of several non-human species (largemouth bass, catfish, cow, sheep, pig) is more correlated with SARS-CoV-2 reads than is human material (Figure~\ref{fig:overall_corr}, upper left).
+If the plot is expanded to all samples (regardless of whether or not they contain SARS-CoV-2 reads), the trends are broadly similar, with fish remaining the most correlated species (Figure~\ref{fig:overall_corr}, lower left).
 It is also informative to take the correlations only across samples collected on January-12-2020, which was the date of most intensive sampling of the wildlife stalls (this is when Q61 was collected).
- For January-12-2020 samples, humans are among the species whose genetic material is most correlated with SARS-CoV-2 reads, but goat and spotted dove have roughly equivalent correlations (Figure~\ref{fig:overall_corr}).
-Raccoon dog and bamboo rat genetic material are not positively correlated with SARS-CoV-2 reads in either sample set.
+ For January-12-2020 samples that contain SARS-CoV-2 reads, humans are among the species whose genetic material is most correlated with SARS-CoV-2 reads, but goat and spotted dove have roughly equivalent correlations (Figure~\ref{fig:overall_corr}, upper right).
+ If we include January-12-2020 samples regardless of SARS-CoV-2 content, then snakehead fish and Malayan porcupine are among the most correlated species.
+Raccoon dog and bamboo rat genetic material are not positively correlated with SARS-CoV-2 reads in any of the sample sets.
 See the interactive version of Figure~\ref{fig:overall_corr} at \url{https://jbloom.github.io/Huanan_market_samples/overall_corr.html} to explore other sample subsets.
@@
 However, the results described here suggest that the utility of these metagenomic analyses does not extend to indicating whether animals at the market were actually infected by virus.
-For instance, the presence of 1 in $\sim$200,000,000 sequencing reads mapping to SARS-CoV-2 in a single sample containing raccoon dog genetic material does not suggest raccoon dogs were infected, given that material from many other species including fish is far more consistently co-mingled with SARS-CoV-2.
+For instance, the presence of 1 in $\sim$200,000,000 sequencing reads mapping to SARS-CoV-2 in a sample containing raccoon dog genetic material does not suggest raccoon dogs were infected, given that material from many other species that certainly were not infected (such as fish) is far more consistently co-mingled with SARS-CoV-2.
+Of course, the lack of association also does not disprove the possibility of infected animals at the market, particularly at a date substantially preceding the Chinese CDC's collection of samples---it simply suggests that analysis of the combined viral and animal content of the available environmental samples is not informative for shedding light on this question either way.
 

### 2023-05-03 paper/paper.tex
revise preprint

@@
 All mitochondrial genomes in the RefSeq database were downloaded, and then filtered to retain only genomes from the phylum Chordata (in contrast, \citet{crits2023genetic} describe retaining all metazoa mitochondrial genomes).
+The reason I limited to chordate rather than metazoa mitochondrial genomes is that all chordates have very similar length mitochondrial genomes, but at the level of metazoa there is wide variation in mitochondrial genome length.
 The raccoon dog mitochondrial genome was not present in this set, so it was separately downloaded from Genbank accession KX964606.
@@
 I thank \citet{liu2023surveillance} for publicly sharing via the NGDC database the deep sequencing data analyzed in this study.
+An initial version of this preprint was revised in part based on helpful comments posted on \textit{bioRxiv} by Florence Débarre and Alex Crits-Christoph (see appendix posted as supplementary material).
 This study utilized the Fred Hutch Scientific Computing infrastructure, which is supported in part by NIH grants S10-OD-020069 and S10-OD-028685.
@@
 
-\newgeometry{left=1in, right=1in, top=48pt, bottom=60pt}
+\newgeometry{left=1.2in, right=1.2in, top=48pt, bottom=60pt}
 \onecolumn
@@
 
+\clearpage
+
+\renewcommand{\thepage}{A\arabic{page}}
+\setcounter{page}{1}
+\renewcommand{\thefigure}{A\arabic{figure}}
+\setcounter{figure}{0}
+\renewcommand{\thetable}{A\arabic{table}}
+\setcounter{table}{0}
+
+\clearpage
+\normalsize
+
+\section{Appendix: response to comments and revision of first version of preprint}
+After the original version of this preprint was posted on \textit{bioRxiv} at \url{https://www.biorxiv.org/content/10.1101/2023.04.25.538336v1}, Florence Débarre and Alex Crits-Christoph posted a set of helpful comments at \url{https://www.biorxiv.org/content/10.1101/2023.04.25.538336v1#comments} using the \textit{bioRxiv} comment feature.
+
+This appendix provides a response to those comments, some of which were used to inform revisions incorporated into this new version of the preprint.
+Note that the computer code / data / paper from the initial version are all also available on GitHub under tag \texttt{bioRxiv\_v1} (\url{https://github.com/jbloom/Huanan_market_samples/tree/bioRxiv_v1}) and the code / data / paper for this revised version are under the tag \texttt{bioRxiv\_v2} (\url{https://github.com/jbloom/Huanan_market_samples/tree/bioRxiv_v2}).
+To see all differences between the two versions, go to \url{https://github.com/jbloom/Huanan_market_samples/compare/bioRxiv_v1...bioRxiv_v2} which shows the differences between the GitHub tags.
+
+In addition to the revisions made in response to the comments below, I also made a minor tweak to ensure that the same \texttt{fastp} pre-processing filters were being applied to both the single- and paired-end sequencing data.
+Overall, none of the modifications to the code led to anything more than minor changes in the overall results.
+
+Below the comments posted on \textit{bioRxiv} by Florence Débarre and Alex Crits-Christoph are in {\bf \color{purple} purple text}, and my responses are in {\bf black text}.
+
+\color{purple}
+
+\subsection{Comments and responses}
+
+In this preprint, Bloom re-analyses a dataset of metagenomic data recently shared on open repositories by \citet{liu2023surveillance}.
+A subset of these data, previously made available on GISAID, was also previously analyzed by \citet{crits2023genetic}---of which we are co-authors.
+
+\begin{enumerate}
+\item Bloom’s analysis largely confirms the genetic identification of wildlife species by \citet{crits2023genetic}.
+
+The identification of animal species differed between \citet{liu2023surveillance} and \citet{crits2023genetic}, in particular regarding the abundance of raccoon dog genetic material in sample Q61.
+Bloom’s analysis, done with similar but independent methods from \citet{crits2023genetic}, largely replicates their findings.
+
+A comprehensive Github repository allows any interested reader to conveniently check Bloom’s results by themselves.
+
+{\color{black}
+I agree that my analysis largely confirm the analysis of the mammalian mitochondrial genetic material in your study.
+As I note in my preprint, it is reassuring to have two independent studies reach largely consistent results on this point.
+As I also note, I think the main value of your earlier work was to provide genetic confirmation of which animals were at the market, and my study basically independently confirms your findings on that point.
+
+I am glad you find the GitHub repository useful!
+}
+
+\item The correlation analysis presented by Bloom contains numerous flaws.
+
+Bloom reproduced the correlation analysis initially presented in \citet{liu2022surveillance}, but Bloom’s analysis, as conducted, is inappropriate to answer the question of which animal hosts shed the viral material detected.
+\citet{crits2023genetic} were missing data to carry out such an analysis, but warned against its use.
+Here we repeat and expand the arguments presented in \citet{crits2023genetic}.
+These points would need to be taken into account for a correlation analysis to be valid, but were not by Bloom (and often just cannot be):
+
+{\color{black}
+At a high level, I agree that the lack of any correlation between SARS-CoV-2 and the genetic material of susceptible animals does not \emph{disprove} the possible existence of an infected animal (for reasons you list below).
+However, it does indicate that the genetic content of these samples is insufficient to reliably indicate whether any animal was infected. That is the claim my preprint makes.
+In the end, if the SARS-CoV-2 content is not associated with the genetic content from susceptible animals, then certainly the genetic material cannot reliably indicate that the animals were infected and shedding the virus.
+
+I think we are in agreement that co-mingling of viral and animal genetic material in these samples is insufficient to indicate if any animals were infected.
+Regardless of perceptions/discussion on social or popular media, your report never claimed proof of infected animals, and my preprint never claimed disproof of infected animals.
+Rather, my understanding is that by fully analyzing the data we are converging on the view that the genetic content of these samples is simply insufficient to reliably determine if any animals were infected.
+}
+
+\begin{enumerate}[label=\alph*)]
+
+\item The outbreak had been ongoing for weeks in the market when the samples were collected, and multiple humans had been infected.
+By then, most virus in the market will come from human cases, especially as stalls containing human cases were targeted first, when sampling was performed on January 1.
+The large number of samples from these stalls with sick humans will highly influence correlational analyses.
+As shown in Worobey et al.
+(2022) on similar data obtained via qPCR, both the distance from known human cases and the distance from wildlife stalls independently contributed to the spatial variation in viral positive samples.
+The number of human cases at the time of sampling means that even if infected animals were present at the market, an underpowered correlation analysis of the sequencing data would be unlikely to have revealed it.
+
+{\color{black} I agree that the late date of sample collection is a major limitation, and mention that several times in my preprint.
+As a say in the preprint, by the time the samples were collected humans had spread the virus widely across the market.
+That means the environmental samples are not particularly informative on the original source of the virus in the market in either direction, which is the central conclusion I draw.}
+
+\item The correlation analysis treats all animals of a given species as an homogeneous group.
+There were however multiple stalls selling the same animal species, but likely not from the same supply sources.
+The presence of uninfected animals in one stall would affect the result of a correlation analysis, but would not invalidate the potential presence of infected animals in another stall.
+
+{\color{black} Certainly I agree that this is a limitation. As I say in the preprint, these environmental samples are not sufficient for reliably indicating if any animals were infected.}
+
+\item The samples themselves are not homogeneous: for instance, within-sample animal diversity is different when the swab was taken directly from fish packaging than from the ground.
+
+In addition, wildlife stall samples tend to be composed of a diverse menagerie of animal species RNA/DNA (a consequence of animals housed directly on top of each other and equipment shared between them), while samples related to human cases elsewhere in the market tend to be fairly simple.
+Therefore, by correlating abundances within just the positive samples, human-shed samples tend to have very high relative abundances for human DNA (or, e.g., human and fish), while animal shed samples tend to have lower abundances for any given species.
+This is a natural consequence of sequencing abundances being relative, not absolute.
+
+{\color{black}
+There is no correlation between the amount of SARS-CoV-2 genetic material and that from potentially susceptible animals like raccoon dogs or bamboo rats regardless of whether the correlation is taken over just SARS-CoV-2 positive samples or all samples.
+This is shown in the new Figure~\ref{fig:overall_corr} and can also be seen in Figure~\ref{fig:per_species_corr_faceted} and perhaps most easily in the interactive version of that figure linked in its legend.
+}
+
+\item SARS-CoV-2 read counts can differ for multiple reasons: there can be different quantities of virus shed in the first place, and different times since virus deposition will influence detection, as RNA degrades over time.
+While it is not possible to precisely date the time of viral deposition, there are two important considerations.
+First, if wildlife introduced the virus first to the market, animal-shed virus would (on average) occur before human sheddings.
+Second, wildlife stalls were sampled multiple days after stalls with reported cases (January 12 vs January 1).
+Put together, this means there would have been substantially more time for virus shed by animals to decay than virus shed by humans.
+
+{
+\color{black}
+There is no correlation between the amount of SARS-CoV-2 and genetic material from potentially susceptible animals like raccoon dogs or bamboo rats regardless of whether the analysis is performed on all samples or restricted to just samples collected on later dates (such as January 12).
+This can be seen in Figure~\ref{fig:overall_corr}, Figure~\ref{fig:per_species_corr_faceted_later_dates}, and perhaps most easily in the interactive plots at  \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html}, which enable subsetting to samples collected on arbitrary dates.
+Of course this fact does not rule out the possibility that animals shed virus earlier and that viral RNA decayed.
+It is also possible animals never shed virus.
+Both possibilities are consistent with my conclusion that these samples are not sufficient to reliably indicate whether any animals were infected.
+}
+
+\item Virus presence is detected through RNA, host presence through mtDNA, which degrade differently over time.
+Relative abundances therefore depend on the time difference between when the different genetic materials were deposited and when they were collected.
+
+{\color{black}
+See response to prior point.
+}
+
+\item Stalls were sampled multiple times. A correlation analysis would need to take into account spatial proximity and in particular model the per-stall effect.
+
+{\color{black}
+At a qualitative level (just for instance looking at the scatter plots), it is obvious that the samples that contain substantial amounts of material from potentially susceptible animals like raccoon dogs or bamboo rats contain no or very little SARS-CoV-2 (see interactive plots at \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_single.html}).
+This is going to remain true no matter how the samples are partitioned or modeled.
+Given this fact, it seems clear that the samples do not suggest shedding by an infected animal. 
+Of course there could have been infected animals earlier or that shed virus that was not collected in these samples, but in either case that makes these samples uninformative for determining if any animal was infected.
+}
+
+\item The correlations reported in Bloom’s figure 5 are calculated solely on samples that contained any positive reads for SARS-CoV-2, and were therefore sequencing-positive for the virus.
+Therefore, this is not a test of which species were more likely to be in samples that tested positive for the virus, but how abundant each species was in the samples that did test positive.
+This is an important distinction.
+
+{\color{black}
+The correlations in Figure~\ref{fig:per_species_corr_faceted} and its interactive version (\url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html}) show all samples, and I have also added panels with all samples to Figure~\ref{fig:overall_corr}.
+It remains the case even when looking at all samples that SARS-CoV-2 content is mostly associated with material from animals that definitely were not infected.
+}
+
+\end{enumerate}
+
+For these reasons, both positive and negative correlations are largely uninterpretable:
+\begin{itemize}
+
+\item A positive correlation is not indicative of infection.
+As an enlightening example (also highlighted in \citet{crits2023genetic}), the most correlated species in Bloom’s analysis is a fish species that is not susceptible to SARS-CoV-2.
+
+\item A negative correlation is not indicative of lack of infection (see above).
+
+\end{itemize}
+
+{\color{black}
+I agree with this essential point: my preprint simply concludes that the genetic content of these samples is insufficient to reliably determine if there were any infected animals.
+}
+
+\item Some samples appear to be missing in the analysis, which contained high viral read numbers: A20, F13, F54.
+
+{\color{black}
+In my analysis, I filter only for samples sequenced metagenomically, and exclude samples that were enriched for virus for viral sequencing.
+Originally, I was filtering all samples that had any virus-enrichment based sequencing.
+In the updated pipeline, for samples like A20 that have both metagenomic and viral sequencing, I now retain the metagenomic sequencing while ignoring the viral sequencing---thanks for pointing out this oversight.
+This adds a few samples including A20, but does not change any of the overall results.
+However, F13 and F54 are still excluded because according to the metadata in the \textit{Experiments} sheet in the Chinese CDC metadata Excel file at \url{https://ngdc.cncb.ac.cn/gsa/browse/CRA010170}, these samples were not sequenced metagenomically (there are no entries with the ``RNA sequencing of total nucleic acids from environmental swabs for metagenomics'' annotation for them in the \textit{Experiments} sheet).
+}
+
+\item Bloom chose to align reads to chordates’ mitochondrial genomes, but did not justify this choice.
+Only mammals are known to be infected by SARS-CoV-2, so an analysis of just mammalian RNA/DNA, as presented in \citet{crits2023genetic}, is most relevant to the question of which species may have shed the virus in a particular sample.
+If the aim was to provide a better picture of the diversity of animals sold at the Huanan market, all metazoa should have been included, to be able to identify seafood for instance (a key early case was a shrimp seller).
+
+However, the end result is that Bloom’s 20\% chordate inclusion threshold for this table results in most of the samples with both wildlife RNA/DNA (e.g.
+raccoon dogs) and viral RNA being excluded.
+The claim that only one sample of all samples with raccoon dog DNA that contains SARS-CoV-2 is therefore inaccurate: please see the supplementary data of either \citet{crits2023genetic} or Bloom et al.
+for comparative results on the true number of such samples (which is at least 5, depending on the given analysis).
+
+{\color{black}
+The reason I limited to chordate rather than metazoa mitochondrial genomes is that all chordates have very similar length mitochondrial genomes, but at the level of metazoa there is wide variation in mitochondrial genome length---this is now explained in the revised Methods.
+
+In any case, the choice of using chordate versus mammalian has no real effect.
+The 20\% cutoff in Table~\ref{tab:susceptible} was just to make the table small enough to easily fit in the main text.
+I have now added a comparable table (Table~\ref{tab:susceptible_mammal}) that lists mammalian rather than chordate composition, and another table that lists SARS-CoV-2 and raccoon dog content for all samples without any cutoff (Table~\ref{tab:raccoon_dog_long}).
+The essential result is unchanged: the samples with abundant raccoon dog mitochondrial material have little or no SARS-CoV-2 reads.
+This is also easily seen in the interactive scatter plots provided in the paper.
+
+There are a few other samples with low raccoon dog content (less than 20\% of either their chordate or mammalian mitochondrial content, see Table~\ref{tab:raccoon_dog_long}) that have some SARS-CoV-2 reads, but these samples also have more abundant genetic material from many other species.
+So certainly they are not sufficient to support an association between viral and raccoon dog material that exceeds that of any of a large number of other species that clearly were not infected (because they are not susceptible).
+This all supports the conclusion that the genetic contents of these samples are insufficient to conclude if any animal was infected.
+}
+
+\item Bloom’s comments on the definition of positive samples overlook the difference in sensitivity between qPCR and sequencing in environmental sampling---qPCR is highly sensitive, with near single molecule detection resolution, and correlates with the absolute load of the virus in the sample.
+Sequencing, however, is both less sensitive (even when deeply sequencing), and highly dependent on the quantity of what else is in the sample.
+Thus, it is very challenging to identify SARS-CoV-2 in samples sequenced from the environment without any kind of enrichment performed: this is because the virus is comparatively rare compared to abundant environmental microbes (bacteria, fungi) as well as DNA/RNA shed by animals and humans.
+This is why most environmental viral sequencing is done after performing a viral enrichment (either physical or amplicon).
+For a similar comparison of unenriched SARS-CoV-2 sequencing, see the supplementary table S1 of Crits-Christoph et al.
+2021 (“Genome Sequencing of Sewage Detects Regionally Prevalent SARS-CoV-2 Variants”), where 5 unenriched samples with viral Ct values in the range of 30-32 were deeply sequenced, and only 1-17 SARS-CoV-2 reads were obtained.
+For this reason, we do not encourage using only sequencing data quantitative values and ignoring qPCR positive samples.
+There is no indication in the \citet{liu2023surveillance} dataset that we have seen to indicate the qPCR results were impacted by contamination.
+Indeed, 112/116 samples that were reported negative for qPCR were also negative by NGS, indicating a high degree of agreement between the two approaches, and that qPCR is merely more sensitive.
+
+In addition, the definition of a positive sample cannot be done independently of the provenance of the sample: even at very low read count, the presence of other positive samples in the same stall at the same date is indicative of sample positivity.
+
+{\color{black}
+I agree that RT-qPCR could be more sensitive than sequencing.
+But the key point is that the ``positive'' sample with the most raccoon dog genetic material (Q61) tested \emph{negative} by RT-qPCR, and was called positive by the Chinese CDC on the basis of having 1 in 200,000,000 reads align to raccoon dog.
+It is not consistent to call that sample ``positive'' if other samples that also tested negative by RT-qPCR but were never sequenced are called ``negative.''
+Samples should only be analyzed together if they were subjected to the same consistent set of assays: so for instance, all samples could be classified by RT-qPCR, or all samples subjected to both assays could be classified based on both.
+This observation is not a criticism of your work: the basis for these classifications only recently became clear with the new \citet{liu2023surveillance} paper.
+I just think we should now consider these points now that they are apparent.
+
+As far as other samples in the same stall, it is clear that SARS-CoV-2 was spread widely across the market: that is why we see it on samples full of material from fish, snakes, etc.
+Therefore, it is my contention that we simply can't conclude much about whether any animal was infected if the virus is spread widely in the environment uncorrelated with any specific animal genetic material.
+This is true regardless of whether we do the analysis at the level of samples or stalls.
+}
+
+\item Citations to news articles are vague or misleading. It would be preferable to stick to arguments presented in scientific documents, rather than rely on newspaper articles.
+
+\begin{enumerate}[label=\alph*)]
+
+\item
+As rationale for his analysis, in the Introduction of his preprint, Bloom claims that “Other scientists pointed out that if the raw data were shared, it would be possible to expand upon the analysis in the 2022 Chinese CDC preprint to determine if the abundance of SARS-CoV-2 genetic material correlated with material from other species (Cohen 2022a,b)”, where the two Cohen citations are news articles.
+The authors quoted in these articles (Robertson, Rambaut, Holmes) did not mention Figure 4a because they thought the analysis was useful enough to be reproduced, but because it was possible that other animals’ genetic material was present in the environmental samples, and these data were not shown.
+
+{
+\color{black}
+The overall rationale of my study was to examine the quantitative association between SARS-CoV-2 content and animal genetic material.
+Figure 4a of the original Chinese CDC study was the only prior work on this topic, and it was widely discussed including in the three Cohen news articles I mentioned, one of which even re-printed the figure.
+I think it is important to provide this as background, and I am careful to distinguish what was discussed in \citet{crits2023genetic} (which did not discuss this figure) versus news articles.
+}
+
+\item
+By conflating their views with misinterpretations of news reports written about their work, Bloom misrepresents the conclusions of \citet{crits2023genetic}:
+\citet{crits2023genetic} did not conclude they had found the intermediate host and it was raccoon dogs: their report clearly stated that they could not “identify the intermediate animal host species from these data”, and that the presence of infected animals was a “plausible explanation” for their results.
+And they explicitly stated that “the most abundant animal in the sequencing data of a particular sample is not necessarily the source of the virus in that sample”.
+
+{
+\color{black}
+I am careful to delineate that \citet{crits2023genetic} did \emph{not} analyze the SARS-CoV-2 content in their report, and so never claimed that animal and viral co-mingling in the environmental samples indicated an infected animal.
+My preprint mentions several times how it is consistent with the results reported in \citet{crits2023genetic}, but adds more analyses that look at association between animal and viral material.
+
+However, it is true that the news articles I cite \emph{did} talk about co-mingling of material suggesting infected animals, and those were widely read.
+I therefore provide the full context to help disentangle the confusion between what was in the scientific report versus the related news articles, which the scientific authors obviously did not write.
+}
+
+\end{enumerate}
+
+\end{enumerate}
+
 \end{document}

### 2023-06-29 paper/paper.tex
fix typo in methods (#3)
Methods incorrectly said raccoon dog mitochondrial genome was manually downloaded when it was actually Amur hedgehog. Code is correct, typo is just in methods.

See https://twitter.com/BenFPiercePhD/status/1654364200209207298

@@
 The interactive versions of the correlation plots at \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html} and \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_single.html} enable the reader to explore the correlations for different sample subsets and methods of calculating the statistics.
-Exploration of these interactive plots shows that it takes a great deal of cherry picking of parameter combinations to put any plausibly infectable hosts at the top of the list of species whose mitochondrial material is most correlated with the abundance of SARS-CoV-2.
+%Exploration of these interactive plots shows that it takes a great deal of cherry picking of parameter combinations to put any plausibly infectable hosts at the top of the list of species whose mitochondrial material is most correlated with the abundance of SARS-CoV-2.
 
@@
 The reason I limited to chordate rather than metazoa mitochondrial genomes is that all chordates have very similar length mitochondrial genomes, but at the level of metazoa there is wide variation in mitochondrial genome length.
-The raccoon dog mitochondrial genome was not present in this set, so it was separately downloaded from Genbank accession KX964606.
+The Amur hedgehog mitochondrial genome was not present in this set, so it was separately downloaded from Genbank accession KX964606.
 All the mitochondrial genomes were then filtered to remove highly similar ones.

### 2023-06-29 paper/reviewer_response.tex
add reviewer comments

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
\usepackage{color}
\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex

\usepackage{amssymb}
\usepackage{hyperref} 
\usepackage[round,semicolon]{natbib}

\newcommand{\comment}[1]{{\color{red}[\textsl{#1}]}}
\newcommand{\response}[1]{{\color{black}#1}}

\title{Response to reviews of ``Association between SARS-CoV-2 and metagenomic content of samples from the Huanan Seafood Market'' for \textit{Virus Evolution}}
\author{}

\begin{document}
\maketitle

\emph{Below, the reviewer and editor comments {\color{blue} are in blue}, and my responses are in black.}

\color{blue}

\subsection*{Reviewer: 1}

Comments to the Author

The manuscript by Jesse Bloom performs a thorough analysis of the metagenomic sequence derived from the Huanan Seafood Market collected shortly after the first cases of SARS-CoV-2 in China.  The manuscript does a fantastic job of boiling down the data and putting it onto interactive online outputs that a normal person can play with.  The basic conclusion is that the data can be used to determine what animals were present in the market, but cannot be used to determine if any of the animals were infected.

The topic of this manuscript is very controversial, so if it is to be taken seriously by readers (regardless of their opinions) it needs to be as unbiased as possible.  While the author has mostly ‘stuck to the fact’ there are a few places where those facts could have been presented more neutrally.

The author states more than once that the first infections were in November or earlier.   The statement in the introduction was, “A significant caveat of the Chinese CDC study is that all the samples were collected on January-1-2020 or later, which is at least a month after the first human infections in Wuhan.”  While this is probably correct, it should not be stated as a fact.  I looked up most of the reference cited and none of them were so definitive.  Among the documents was an unclassified government summary which concluded that the virus ‘probably’ emerged no later than November, 2019.  Without the ‘probably’ or some other modifier, what is written in the manuscript is misrepresenting the government document.

Conspicuously missing from the introduction is any discussion of the epidemiology papers that suggest that the virus originated in the market.  A description could come with commentary, but to simply ignore the work makes the manuscript come across as biased.

I don’t love the 20\% cutoff.  I understand why the table in the paper needs to have the 20\% cutoff to make it a manageable size, but the complete dataset is included as supplementary, so there isn’t any reason for the text to focus on the arbitrary 20\% as much as it does.  I suppose it isn’t essential to do away with it, but people who don’t like the paper are going to say that this is cherry picking.

The abstract states ``For instance, 14 samples have >20\% of their chordate mitochondrial material from raccoon dogs, but only one of these samples contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.''  After reading the manuscript this sentence makes sense, but a lot of people are probably only going to read the abstract, so the 20\% cutoff seems would seem arbitrary, and the 1 in 200M has not context.  In the abstract makes it sounds ridiculously low, but many of the samples had viral copies in the single digits.  If this sentence is included I think it should be reworded or put into context (elaborating on what a `typical' number of reads is).

Related to the above point, it seems to me that the manuscript missed an opportunity to drive home the point that the data cannot be used to infer viral source.  The one way that one could get a smoking gun from this kind of data is if there was a sample that had a whole lot of viral material, and basically only one species associated with it (which didn’t happen).  In fact, the two samples with the highest SC2 levels didn’t have high levels of raccoon dog, human or any other susceptible species.  Using that as a foil one could point out that the positive with the most raccoon dog had only once SC2 copy, while the samples with the most SC2 copies (>1000 times more) had no prevalent candidate hosts associated with it at all.  Just a thought.

\subsection*{Reviewer: 2}

Comments to the Author

In this paper, Bloom performs an independent analysis of metagenomic environmental samples from the Huanan Seafood market obtained through Chinese surveillance efforts. Because of the extensive press coverage and vigorous debates on social media regarding various claims and counter-claims about whether or not these data can be used to argue that potential animal intermediate hosts of SARS-CoV-2 (e.g. infected raccoon dogs) were present in the market, Bloom's analysis is valuable. Without this context, it would not be remarkable or publishable. However, it provides a very important null result, and also highlights a number of questionable scientific practices and overzealous interpretations by others. It should be therefore be published. I also commend Bloom on his unfailing commitment to reproducibility, with all code, and results made available (and usable)

I have several specific comments/suggestions which I believe could strengthen the manuscript and improve its readability

1. I think it is critically important to specifically state that EVEN IF there was strong positive correlation between mammalian read counts and SC-2 read counts, it would still not be possible to use these data to argue that ANY particular animal was infected. Correlation $\ne$ causation, and there are too many confounders to note.

2. There is very robust literature on estimating species abundance from RNASeq-type metagenomic data. Specialized tools (e.g. Kraken 2) exist for efficient and accurate classification of such reads. It would be worthwhile to explain briefly why it is necessary to develop a "de novo" pipeline. Many commonly recognized issues (e.g. short reads which are not diagnostic at species level because they are found in multiple genomes) are well understood and quantified. C-C (presumably) used contigs to address this lack of ``read-level'' precision (they are very SHORT in the data discussed), but that's a hack.

3. One specific issue with the minimap2 mapping to a concatenome is how "non-uniquely" mapping reads are dealt with. "Methods" mention something about using only primary maps, but more discussion is warranted. Otherwise, reads which could be similarly diagnostic of multiple species (assuming there are some), would simply be assigned to the first genome in the concatenated string. Maybe this is not an issue here, but should be discussed.

4. I think the point of inconsistent calling of +/- SARS-CoV-2 samples by the original authors of the samples is VERY important and needs to be emphasized, especially because the positivity pattern was used to make very strong spatial correlation claims by Worobey and others. I would also suggest that instead of simply stating that 1/200M is statistically indistinguishable from 0/200M, it might be more informative to talk about limits of detection. In other words, if we accept 1/200M as a positive, we cannot rule out 0/200M as being positive, and perhaps talk about how many reads you would actually need to make some statistically sound claims (as a function of total read \#). You could put this in Table 1, for example.

5. I think several figures can be removed from the printed manuscript without information loss. For example Figure 1 shows a lot of near ~1 correlations, and Figure 4 shows a lot of "Rorschach blot" scatter-shots of points.

6. For regression analyses, I might suggest something more robust to outliers that linear/rank regression. Theil-Sen estimators, or quantile regression might be better here.

7. Two points of curiosity
   (a) For the samples with sufficient SC-2 reads, is it possible to assemble any contigs and see what lineage the virus is?
   (b) Are there any other viral pathogens present?

\section*{Associate Editor: Koelle, Katia}

Comments to the Author:

Both reviewers are of the opinion that this submitted work is highly valuable and thoroughly performed. One of the reviewers has several suggestions/questions that pertain to the methods used in the sequencing analyses, as well as suggestions for streamlining the results presented. The other reviewer has several text edit suggestions, including referencing the epidemiology papers that argue that the virus originated at the market. Both reviewers point out that the manuscript should emphasize the point that the sequencing data cannot be used to infer the viral source. Based on these reviews, I recommend minor revision of this manuscript.

\color{black}
\bibliographystyle{genetics}
{\small
\bibliography{references.bib}
}

\end{document}

### 2023-06-29 paper/reviewer_response.tex
Reviewer 1 comment 1: better caveat date of emergence estimates

@@
 
+\response{This is a good and accurate summary of the conclusions.}
+
 The topic of this manuscript is very controversial, so if it is to be taken seriously by readers (regardless of their opinions) it needs to be as unbiased as possible.
 While the author has mostly ‘stuck to the fact’ there are a few places where those facts could have been presented more neutrally.
+
+\response{
+I have indeed tried to stick as narrowly to the facts as possible, and appreciate the suggestions below of places where descriptions can be made even more neutral.
+As described below, I have made revisions along the lines of all of these suggestions.
+}
 
@@
 Without the ‘probably’ or some other modifier, what is written in the manuscript is misrepresenting the government document.
+
+\response{
+I have added the qualifier ``probably'' to the references in both the Introduction and Discussion about how the first cases probably occurred no later than November of 2019.
+I have also added several more citations to papers that use various methods to try to date the first human infections.
+All of these papers make a point estimate of the date of the first human infection that is no later than November of 2019, although some of them do estimate credible or confidence intervals that extend into early December of 2019.
+}
 

### 2023-06-29 paper/paper.tex
acknowledge Ben Pierce
Reviewer 1 comment 1: better caveat date of emergence estimates

@@
 They reported that none of the 457 animal samples tested positive for SARS-CoV-2, but that 73 of 923 environmental samples tested positive.
-A significant caveat of the Chinese CDC study is that all the samples were collected on January-1-2020 or later, which is \emph{at least} a month after the first human infections in Wuhan~\citep{odni2022,zhang2020viral,van2020emergence,pipes2021assessing}.
+A significant caveat of the Chinese CDC study is that all the samples were collected on January-1-2020 or later, which is probably \emph{at least} a month after the first human infections in Wuhan~\citep{zhang2020viral,van2020emergence,he2020evaluating,pipes2021assessing,pekar2021timing,odni2022,pekar2022molecular}.
 Despite this significant caveat, the data reported in the Chinese CDC study has been variously interpreted to support arguments that virus in the market was human derived~\citep{liu2022surveillance}, that the outbreak originated from live animals sold in the market~\citep{worobey2022huanan}, or that the virus spread among humans in Mahjong rooms or toilets in the market~\citep{courtier2022sars}.
@@
 Therefore, by the time the samples were collected, SARS-CoV-2 had been spread widely across the market by humans regardless of its original source---as evidenced by the results reported here, which show viral genetic material coincident with material from myriad animals ranging from fish to snakes to mammals.
-The first human infections with SARS-CoV-2 in Wuhan occurred no later than November of 2019~\citep{odni2022,zhang2020viral,van2020emergence,pipes2021assessing}, which is over a month before the Chinese CDC reports that it began to collect samples from the market.
+The first human infections with SARS-CoV-2 in Wuhan probably occurred no later than November of 2019~\citep{zhang2020viral,van2020emergence,he2020evaluating,pipes2021assessing,pekar2021timing,odni2022,pekar2022molecular}, which is over a month before the Chinese CDC reports that it began to collect samples from the market.
 For this reason, further insight into the origins and early spread of SARS-CoV-2 will likely require learning more about events or cases that occurred no later than November or early December of 2019.
@@
 An initial version of this preprint was revised in part based on helpful comments posted on \textit{bioRxiv} by Florence Débarre and Alex Crits-Christoph (see appendix posted as supplementary material).
+I thank Benjamin F. Pierce for identifying an error in the reference genome description in the original version (\url{https://github.com/jbloom/Huanan_market_samples/pull/3}).
 This study utilized the Fred Hutch Scientific Computing infrastructure, which is supported in part by NIH grants S10-OD-020069 and S10-OD-028685.

### 2023-06-29 paper/paper.tex
Reviewer 1 point 2: discuss epi of market more

@@
 But by mid to late January of 2020, it was clear SARS-CoV-2 was spreading from human to human, and had been for some time~\citep{chan2020familial,phan2020importation,li2020early,nishiura2020initial}.
-In addition, Chinese scientists published papers reporting that some of the earliest identified human cases in December 2019 had no link to the Huanan Seafood Market~\citep{huang2020clinical,chen2020epidemiological}.
+In addition, Chinese scientists published papers reporting that some of the earliest identified human cases in December 2019 had no link to the Huanan Seafood Market~\citep{huang2020clinical,chen2020epidemiological,zhang2020viral}.
 Thus began a debate that continues to this day: was the market an initial source of zoonotic infections from animals, or was it simply a superspreading site that amplified earlier human infections from another source~\citep{cohen2020wuhan}?
 
-It is universally agreed that there were human cases of SARS-CoV-2 at the Huanan Seafood Market by mid December of 2019~\citep{li2020early}.
-It is also known that some animals susceptible to SARS-CoV-2 were sold at the market~\citep{xiao2021animal}.
-The unanswered question is if any animals were infected, and if so whether they infected humans or were infected by them.
+Some papers have argued that the fact that many of the known human cases from December 2019 are centered around market indicates it was the original source of the virus~\citep{worobey2022huanan}, and that an early two-mutation polytomy in the SARS-CoV-2 phylogenetic tree suggests multiple zoonotic jumps there~\citep{pekar2022molecular}.
+Other papers have argued that reports of earlier cases unlinked to the market~\citep{huang2020clinical,chen2020epidemiological,zhang2020viral,ma2020SCMP} and uncertainties about the completeness of case data provided by the Chinese government~\citep{cohen2022anywhere,hvistendahl2023} mean that the first human infections could have occurred upstream of the December 2019 cases at the market~\citep{pekar2021timing,pipes2021assessing,bloom2021recovery}.
+
+Despite these disputes, it is universally agreed that there were human cases of SARS-CoV-2 at the Huanan Seafood Market by mid December of 2019~\citep{li2020early}, and that some animals susceptible to SARS-CoV-2 were sold at the market~\citep{xiao2021animal}.
+The key unanswered question is if any animals were infected, and if so whether they infected humans or were infected by them.
 

### 2023-07-02 paper/reviewer_response.tex
Reviewer 1 point 2: discuss epi of market more
Reviewer point 3: de-emphasize 20% cutoff

@@
 
+\response{
+This is a good suggestion.
+I have added new text (third paragraph of revised Introduction) that discusses the interpretations of epidemiological data about the origin of the virus.
+}
+
 I don’t love the 20\% cutoff.
@@
 
-The abstract states ``For instance, 14 samples have >20\% of their chordate mitochondrial material from raccoon dogs, but only one of these samples contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.''  After reading the manuscript this sentence makes sense, but a lot of people are probably only going to read the abstract, so the 20\% cutoff seems would seem arbitrary, and the 1 in 200M has not context.
+The abstract states ``For instance, 14 samples have $>$20\% of their chordate mitochondrial material from raccoon dogs, but only one of these samples contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.''  After reading the manuscript this sentence makes sense, but a lot of people are probably only going to read the abstract, so the 20\% cutoff seems would seem arbitrary, and the 1 in 200M has not context.
 In the abstract makes it sounds ridiculously low, but many of the samples had viral copies in the single digits.
 If this sentence is included I think it should be reworded or put into context (elaborating on what a `typical' number of reads is).
+
+\response{
+As the reviewer notes, the paper contains in the figures and a supplementary table analysis of the association between SARS-CoV-2 content and mitochondrial material composition for all samples without any cutoff, and the 20\% cutoff was applied just to make Table 1 fit in the main text as the full table is too large.
+
+However, I see the reviewer's point that the 20\% cutoff is reasonable for creating the table, but that the main text should not rely heavily on that particular cutoff in describing the results as otherwise the reader may get the incorrect impression that the 20\% cutoff is somehow integral to the analysis.
+I also agree that the abstract should clearly emphasize how nearly all samples contain low SARS-CoV-2 content.
+I have made revisions to the text to reflect these suggestions. 
+The abstract now emphasizes the low overall SARS-CoV-2 content and relatively de-emphasizes the 20\% cutoff with the new wording below.
+Similar changes in wording have been made to the discussion text.
+\begin{quote}
+\textit{
+However, the SARS-CoV-2 content of the environmental samples is generally very low: only 21 of 176 samples contain more than 10 SARS-CoV-2 reads, despite most samples being sequenced to depths exceeding $10^8$  total reads.
+None of the samples with double-digit numbers of SARS-CoV-2 reads have a substantial fraction of their mitochondrial material from any non-human susceptible species.
+Only one of the 14 samples with at least a fifth of the chordate mitochondrial material from raccoon dogs contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
+Instead, SARS-CoV-2 reads are most correlated with reads mapping to various fish, such as catfish and largemouth bass.
+}
+\end{quote}
+}
 
@@
 In fact, the two samples with the highest SC2 levels didn’t have high levels of raccoon dog, human or any other susceptible species.
-Using that as a foil one could point out that the positive with the most raccoon dog had only once SC2 copy, while the samples with the most SC2 copies (>1000 times more) had no prevalent candidate hosts associated with it at all.
+Using that as a foil one could point out that the positive with the most raccoon dog had only once SC2 copy, while the samples with the most SC2 copies ($>$1000 times more) had no prevalent candidate hosts associated with it at all.
 Just a thought.

### 2023-07-02 paper/paper.tex
Reviewer point 3: de-emphasize 20% cutoff

@@
 I validate the presence of genetic material from numerous species, and calculate mammalian mitochondrial compositions similar to those reported by \citet{crits2023genetic}.
-However, the number of SARS-CoV-2 reads is not consistently correlated with reads mapping to non-human susceptible species.
-For instance, 14 samples have $>$20\% of their chordate mitochondrial material from raccoon dogs, but only one of these samples contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
+However, the SARS-CoV-2 content of the environmental samples is generally very low: only 21 of 176 samples contain more than 10 SARS-CoV-2 reads, despite most samples being sequenced to depths exceeding $10^8$  total reads.
+None of the samples with double-digit numbers of SARS-CoV-2 reads have a substantial fraction of their mitochondrial material from any non-human susceptible species.
+Only one of the 14 samples with at least a fifth of the chordate mitochondrial material from raccoon dogs contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
 Instead, SARS-CoV-2 reads are most correlated with reads mapping to various fish, such as catfish and largemouth bass.
@@
 Consistent with analyses by the Chinese CDC~\citep{liu2022surveillance}, I find some correlation between the abundance of SARS-CoV-2 and human genetic material, but this correlation is weaker than for several non-infectable animals and so on its own is insufficient to identify the source of the viral material.
-Mitochondrial material from most susceptible non-human species sold live at the market is negatively correlated with the presence of SARS-CoV-2: 13 of the 14 samples with at least 20\% of their chordate mitochondrial material from raccoon dogs contain no SARS-CoV-2 reads, and the other sample contains just 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
-Likewise, none of the six samples with at least 20\% bamboo rat mitochondrial material contain any SARS-CoV-2 reads.
+Mitochondrial material from most susceptible non-human species sold live at the market is negatively correlated with the presence of SARS-CoV-2: for instance, 13 of the 14 samples with at least a fifth of their chordate mitochondrial material from raccoon dogs contain no SARS-CoV-2 reads, and the other sample contains just 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
 These findings are compatible with the results in \citet{crits2023genetic}, since that study did not report any analysis of SARS-CoV-2 content.

### 2023-07-02 paper/reviewer_response.tex
Reviewer 1 point 4: samples inconclusive on origins

@@
 I suppose it isn’t essential to do away with it, but people who don’t like the paper are going to say that this is cherry picking.
-
 The abstract states ``For instance, 14 samples have $>$20\% of their chordate mitochondrial material from raccoon dogs, but only one of these samples contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.''  After reading the manuscript this sentence makes sense, but a lot of people are probably only going to read the abstract, so the 20\% cutoff seems would seem arbitrary, and the 1 in 200M has not context.
@@
 Just a thought.
+
+\response{
+I agree that the main finding is that the metagenomic content of the environmental samples cannot be used to infer the viral source.
+I have tried to emphasize this point as the major finding.
+For instance, the abstract concludes with the statement that the samples are unable to ``reliably indicated whether any animals were infected by SARS-CoV-2.''
+Likewise, the last two paragraphs of the discussion touch on this point (eg, ``When considered in the larger context, the inability of the environmental samples to inform on the origins of the virus is unsurprising.'').
+However, I do not think it was \textit{a priori} certain that the samples would be uninformative on this question: as the reviewer notes, if there were a very strong association with a single species, that would be close to a smoking gun.
+
+Overall, I agree with the reviewer's points above, and hope the current text, particularly the last two paragraphs of the discussion, adequately discuss these points.
+}
 

### 2023-07-02 paper/reviewer_response.tex
Reviewer 2 point 1

@@
 
+\response{
+I agree that this paper essentially just shows that these samples don't tell us much about the origin of the virus.
+Normally, that would not be a finding of particular interest, but because there has been so much intense speculation about this question this paper is an example where an inconclusive or null result (simply showing that the data can't answer the question) is of interest.
+
+I am glad the reviewer found the code and plots useful, and am hopeful that others have also used them to explore these data more.
+}
+
 I have several specific comments/suggestions which I believe could strengthen the manuscript and improve its readability
@@
 Correlation $\ne$ causation, and there are too many confounders to note.
+
+\response{
+I am not certain agree with this point (see also the last point made by Reviewer 1 and my response).
+
+I fully agree that these particular data aren't sufficient to tell either way whether any animal was infected.
+But one can hypothetically imagine a scenario in which there was an extremely strong correlation between SARS-CoV-2 content and genetic material from a single species.
+For instance, if all the samples with appreciable SARS-CoV-2 were dominated by raccoon dog material and no other samples had any SARS-CoV-2, that would in my view be strong evidence for infected raccoon dogs.
+Likewise if all the samples with appreciable SARS-CoV-2 were dominated by human genetic material it would strongly suggest humans were shedding the virus.
+Of course, in actuality it turns out that there isn't any trend like that, and to the extent there is any correlation it is with species that don't make sense as having been infected.
+So I fully agree that it turns out that these samples are not sufficient to show that any particular animal was infected, and there are plenty of reasons we can see in hindsight why this is the case (samples collected late, low viral content, lots of mixing of material from different species, samples collected at different times, etc).
+
+However, I do think that hypothetically if there had been very strong patterns in the data it could have provided strong evidence for infected animals of a particular species, so I don't really feel comfortable stating that the samples could not have been informative even in principle.
+}
 

### 2023-07-02 paper/reviewer_response.tex
Reviewer 2 point 2: methods rationale

@@
 
+\response{
+This is an excellent question.
+My primary reason for using the \textit{de novo} mitochondrial pipeline rather than something like Kraken2 was to enhance the comparability of my results to those of Crits-Christoph et al, since I thought a major question was likely to be if any discrepancies between their findings and mine were due to methodological differences.
+Crits-Christoph et al did not make the relevant code for the metagenomic analyses publicly available alongside their report, so I wrote a \textit{de novo} pipeline that mirrored what they described in the methods of their report.
+
+There are also several technical reasons why standard use of Kraken2 with off-the-shelf reference genome sets would not work:
+\begin{itemize}
+\item A standard Kraken2 pre-built genomic database would not be suitable for this study, as some of the species of particular interest (eg, bamboo rat) do not even have sequenced genomes on RefSeq.
+\item Use of mitochondrial rather than nuclear genomes is therefore probably preferable because a much wider range of species, including essentially all the potentially susceptible non-human species at the market, have fully sequenced mitochondrial genomes.
+\item Although this is not clearly explained in the methods of Crits-Christoph et al, at the current time at least one of the species that is a focus of their study (Amur hedgehog) does not even have a mitochondrial genome on RefSeq, and that sequence had to be obtained elsewhere manually from Genbank (see methods of my pre-print).
+Therefore, it required some steps beyond just downloading RefSeq mitochondrial genomes to build a reference set comparable to that used by Crits-Christoph et al.
+\item Given these complexities, I wanted to be sure I explicitly understood how all steps were being handled in reference building and handling of multi-mapping reads.
+This made me prefer a pipeline where I had written each step explicitly rather than a package like Kracken2 which streamlines several steps together.
+\end{itemize}
+
+Nonetheless, I think a similar analysis could have been performed using Kraken2, and I mostly took the approach here to ensure as much comparability as possible to the report by Crits-Christoph et al.
+
+I have added text to the portion of the manuscript that explains the methods used for this part of the analysis explaining the points above.
+}
+
 3.

### 2023-07-02 paper/paper.tex
Reviewer 2 point 2: methods rationale

@@
 To create an mitochondrial genome alignment reference set, I followed a procedure partially analogous to that described by \citet{crits2023genetic}.
-All mitochondrial genomes in the RefSeq database were downloaded, and then filtered to retain only genomes from the phylum Chordata (in contrast, \citet{crits2023genetic} describe retaining all metazoa mitochondrial genomes).
+Note that a similar analysis could also be implemented using more standard metagenomic analysis tools such as Kraken2~\citep{wood2019improved}, as was in fact done in the Chinese CDC study~\citep{liu2023surveillance}.
+The reasons for instead implementing the pipeline used here were as follows:
+\begin{itemize}
+\item To ensure comparability to \citet{crits2023genetic} by using an approach that parallels the one they described.
+\item To use a reference set that includes all of the potentially susceptible species sold at the market: many of these species lack full nuclear genomes on RefSeq, and one (Amur hedgehog) doesn't even have a mitochondrial genome on RefSeq.
+Therefore, the manual process used to build the mitochondrial reference set described below was necessary to ensure all these species were in the reference.
+\end{itemize}
+
+For the current study, all mitochondrial genomes in the RefSeq database were downloaded, and then filtered to retain only genomes from the phylum Chordata (in contrast, \citet{crits2023genetic} describe retaining all metazoa mitochondrial genomes).
 The reason I limited to chordate rather than metazoa mitochondrial genomes is that all chordates have very similar length mitochondrial genomes, but at the level of metazoa there is wide variation in mitochondrial genome length.

### 2023-07-02 paper/reviewer_response.tex
Reviewer 2 point 3: mapping quality ensures uniquely mapped only

@@
 
+\response{
+This is a great question, because as the reviewer notes the number of aligned reads will depend on how non-uniquely mapping reads are handled.
+Briefly, these are reads that map to two different species in the reference equally well.
+This happens with some reads, although the rate at which it occurs is reduced by use of only mitochondrial genomes in the reference (which are much shorter than nuclear genomes) and the filtering of the reference set to remove nearly identical mitochondrial genomes.
+
+The current pipeline only counts reads that map better to one species than any other species.
+As the reviewer notes, if we simply took all primary alignments that would not be sufficient to guarantee consideration of only uniquely best mapping reads, because if a read maps equally well to two sequences (ie, species) in the reference then \texttt{minimap2} (the aligner used here) will randomly assign the primary mapping to one of the two.
+(Note \texttt{minimap2} chooses the primary alignment randomly among equally good hits, not by taking the first of the concatenated genome.)
+So primary aligned reads are \emph{not} guaranteed to be reads that have uniquely best mapping locations, and so could match equally well to multiple species.
+
+However, as described briefly in the Methods of the original version of my paper, my analysis only counts primary mappings that \emph{also have a mapping quality of at least 4}.
+The mapping quality is an indicator of how confidently a read maps to a specific location in the reference versus all other locations, and read will only have a mapping quality $>$0 if the primary mapping is better than all other mappings.
+Therefore, the combined requirement that reads be primary and have a mapping quality greater than zero has the effect of only counting reads that uniquely map best to just one species.
+
+I have elaborated to better explain this point in the relevant section of the Methods, as suggested by the reviewer.
+
+Note also that in the \texttt{config.yaml} file for the pipeline, you can change the minimum mapping quality from 4 to 0 and re-run the pipeline.
+If you do that, non-uniquely mapping reads are assigned randomly to one of the best-mapping reference species, which has the net effect (averaged over many reads) of giving a fractional score for these reads (eg, 0.5 for a read that maps to two species).
+If you re-run the pipeline with this option, you will see that although the quantitative results shift some, the qualitative conclusions are largely unchanged.
+}
+
 4.

### 2023-07-02 paper/paper.tex
Reviewer 2 point 3: mapping quality ensures uniquely mapped only

@@
 The accessions for the 4,170 chordate mitochondrial genomes that formed this set are listed in Table~\ref{tab:mito_retained}.
-The counts of reads mapping to each mitochondrial genome for each sequencing run and sample are provided in Tables~\ref{tab:mito_composition_by_run} and \ref{tab:mito_composition_by_sample}, respectively.
+The counts of reads mapping uniquely to each mitochondrial genome for each sequencing run and sample are provided in Tables~\ref{tab:mito_composition_by_run} and \ref{tab:mito_composition_by_sample}, respectively.
 
@@
 The resulting alignments were filtered to only retain primary alignments with mapping quality of at least 4.
+Note that one effect of counting only primary alignments with mapping quality $>$0 (4 in this case) is to only count reads that map uniquely---that is, they map better to the mitochondrial genome of one species than all other species in the reference.
+This is because although \texttt{minimap2} assigns the primary mapping randomly  if there are multiple sequences in the reference with equally good mappings, a read will only have a mapping quality that exceeds zero if it maps to one location better than all other locations.
 

### 2023-07-02 paper/paper.tex
update competing interests

@@
 \section{Competing interests}
-JDB is on the scientific advisory boards of Apriori Bio, Aerium Therapeutics, Invivyd, the Vaccine Company, and Oncorus; consults for GSK; and receives royalty payments as an inventor on Fred Hutch licensed patents related to deep mutational scanning of viral proteins.
+JDB is on the scientific advisory boards of Apriori Bio, Aerium Therapeutics, Invivyd, and the Vaccine Company; consults for GSK; and receives royalty payments as an inventor on Fred Hutch licensed patents related to deep mutational scanning of viral proteins.
 

### 2023-07-02 paper/reviewer_response.tex
Reviewer 2 point 4: consistency of calling samples + vs -

@@
 
+\response{
+I agree that the calling of samples as positive / negative based on the binary presence / absence of any SARS-CoV-2 read is not statistically well supported, as we cannot reject the null hypothesis that the underlying amount of SARS-CoV-2 is significantly lower in a sample with 0 in 200-million SARS-CoV-2 reads versus a sample with 1 in 200-million SARS-CoV-2 reads.
+If we use a Fisher exact test and compare two samples with 200-million total reads one of which has 0 SARS-CoV-2 reads and the other of which has $N$ SARS-CoV-2 reads, we can only reject the null hypothesis that the two samples are drawn from reads with the same underlying distribution of SARS-CoV-2 content at $P \le 0.05$ when $N \ge 6$.
+
+Therefore, as suggested by the reviewer, I have expanded on the relevant portion of the text explaining this fact, and described what threshold is needed for statistical significance on the statement that the SARS-CoV-2 content of the two samples is significantly different.
+
+That said, I have not included any claims about how this fact would affect the conclusions that have been drawn by Worobey and others.
+Certainly it is a factor that needs to be considered---however, many of the samples analyzed by Worobey and co-workers were also analyzed by RT-qPCR in addition to next-generation sequencing, so substantially more analysis of the full dataset and spatial distribution of samples is needed to determine how the above point might affect any conclusions by Worobey et al.
+That analysis is beyond the scope of the current study, so here I have simply noted the lack of statistical rigor in how samples have been called positive / negative from deep-sequencing data by the Chinese CDC, but have refrained from speculating on how exactly that might impact downstream analyses using the Chinese CDC's positive / negative classifications.
+}
+
 5.

### 2023-07-02 paper/paper.tex
Reviewer 2 point 4: consistency of calling samples + vs -

@@
 Inconsistency in criteria used to classify SARS-CoV-2 positivity in Chinese CDC study, illustrated with four example samples.
-There is no consistent rationale for classifying Q61 as positive but E-10-29-2 and A1 as negative: all three were reported negative by RT-qPCR, and A1 was not analyzed by sequencing while the difference in the number of SARS-CoV-2 reads between Q61 and E-10-29-2 is not statistically significant.
+There is no consistent rationale for classifying Q61 as positive but E-10-29-2 and A1 as negative: all three were reported negative by RT-qPCR, and A1 was not analyzed by sequencing while the difference in the number of SARS-CoV-2 reads between Q61 and E-10-29-2 is not statistically significant ($P = 1$ for both Fisher exact or $\chi^2$ test).
 RT-qPCR results are from the second supplementary table of \citet{liu2023surveillance} (or equivalently the first table of \citet{liu2022surveillance}).
@@
 Now that the full data are available, we can examine the criteria used to classify samples as positive or negative for SARS-CoV-2.
-It appears that the Chinese CDC classified samples as positive if they met \emph{either} of two criteria: they tested positive for SARS-CoV-2 by RT-qPCR, \emph{or} they were metagenomically sequenced and contained at least one read mapping to SARS-CoV-2.
+It appears that the Chinese CDC classified samples as positive if they met \emph{either} of two criteria: they tested positive for SARS-CoV-2 by RT-qPCR, \emph{or} they were metagenomically sequenced and contained at least one read mapping to SARS-CoV-2 (Table~\ref{tab:rt-qpcr} gives sequencing and RT-qPCR data for all samples that were either sequenced or called positive by the Chinese CDC).
 However, these criteria are not consistent because not all samples were analyzed by both methods, and the differences in the number of SARS-CoV-2 reads in the sequencing data are often not statistically significant between samples classified as positive versus negative.
 This inconsistency is illustrated in Table~\ref{tab:consistency} for four example samples: F100 is clearly positive (it both tested positive by RT-qPCR and contained thousands of reads mapping to SARS-CoV-2), but there is no statistical rationale for classifying Q61 as positive but E-10-29-2 and A1 as negative.
-See Table~\ref{tab:rt-qpcr} for merged sequencing and RT-qPCR data for all samples that were either sequenced or called as positive by the Chinese CDC.
+
+To see the lack of statistical rationale for classifying Q61 as positive but E-10-29-2 as negative (Table~\ref{tab:consistency}), we can use a Fisher exact test (or alternatively a $\chi^2$ test) to test the null hypothesis that the reads in samples Q61 and E-10-29-2 are drawn from the same underlying distribution of viral and non-viral reads given that we observe 1 in $2.1 \times 10^8$ SARS-CoV-2 reads in Q61 and 0 in $1.9 \times 10^8$ SARS-CoV-2 reads in E-10-29-2.
+Such an analysis shows that we cannot reject the null hypothesis that samples Q61 and E-10-29-2 have reads drawn from the same underlying distribution; this null hypothesis could only be rejected at a significance level of $P \le 0.05$ if sample Q61 contained $\ge$6 SARS-CoV-2 reads.
+The lack of rationale is even more obvious when comparing Q61 and A1 (Table~\ref{tab:consistency}): both samples were negative by RT-qPCR, but Q61 was called positive based on metagenomic sequencing whereas A1 was never sequenced.
+Therefore, if Q61 is called as positive then A1 should be called as indeterminate, since it was never subjected to the assay used to call Q61 as positive.
 

### 2023-07-02 paper/reviewer_response.tex
Reviewer 2 comment 3: move figure to supplement

@@
 I think several figures can be removed from the printed manuscript without information loss.
-For example Figure 1 shows a lot of near ~1 correlations, and Figure 4 shows a lot of "Rorschach blot" scatter-shots of points.
+For example Figure 1 shows a lot of near ~1 correlations, and Figure 4 shows a lot of ``Rorschach blot'' scatter-shots of points.
+
+\response{
+As the reviewer suggests, I have moved Figure 1 of the original manuscript to the supplement, where it is now Figure S1.
+This is the figure that simply shows that the mammalian metagenomic composition is highly correlated across all shared samples between my analysis and that of Crits-Christoph et al.
+
+However, I have chosen to keep Figure 1 of the original manuscript (now Figure 3 of the revised manuscript) as a main figure.
+I agree with the reviewer that this is a large figure with lots of scatter plots that is a bit overwhelming to look at.
+But this is also represents arguably the most key results of the entire analysis, and since \textit{Virus Evolution} has no length limit I think it is better to keep this large full-page figure so all of the key results are in the main manuscript without having to refer to the supplement.
+}
 

### 2023-07-03 paper/reviewer_response.tex
Reviewer 2 comment 4: Theil-Sen correlation

@@
 
+\response{
+The reviewer is correct that there are a lot of ``outlier'' points in the correlation scatter plots.
+This isn't a substantive issue for the large panel of scatter plots in Figure 3 of the revised manuscript (Figure 4 of the original) since the panels show all the points visually and don't rely on the correlation coefficients for anything other than a numerical summary.
+However, this issue does substantively affect Figure 4 of the revised manuscript (Figure 3 of the original), which plots the Pearson and Spearman correlation coefficients.
+
+The reviewer is correct that Theil-Sen estimators or quantile regression are more robust to outliers.
+However, there are two difficulties with using these instead: the paper never shows the fit lines (which is what those approaches estimate), but rather the correlation coefficient.
+But both Theil-Sen estimators nor quantile regression are designed to estimate fit lines, and don't directly estimate a counterpart to a correlation coefficient.
+However, I did locate a formula for a correlation measure based on the Theil-Sen estimator (\url{https://towardsdatascience.com/a-correlation-measure-based-on-theil-sen-regression-31b8b9ed64f1}).
+I have therefore created a version of Figure 4 that uses this Theil-Sen estimator based correlation rather than the Pearson correlation; it is Figure S4 of the revised manuscript.
+The correlations are somewhat different for the Theil-Sen based estimator than the Pearson or Spearman methods (the Theil-Sen measure is zero for many species where there is no correlation except for a few outliers), but the basic finding holds that the most correlated species across the entire dataset do not include non-human species susceptible to SARS-CoV-2.
+
+I have kept the Pearson / Spearman correlation figure as the main text figure for consistency with the Chinese CDC pre-print, and put this new plot as a supplementary figure.
+}
+
 7. Two points of curiosity

### 2023-07-03 paper/paper.tex
Reviewer 2 comment 3: move figure to supplement
Reviewer 2 comment 4: Theil-Sen correlation

@@
 The counts of reads mapping uniquely to each mitochondrial genome for each sequencing run and sample are provided in Tables~\ref{tab:mito_composition_by_run} and \ref{tab:mito_composition_by_sample}, respectively.
-
-\begin{figure*}
-\includegraphics[width=\linewidth]{figures/crits_christoph_vs_current_run_corr.pdf}
-\caption{
-Correlation of the number of read counts mapping to mitochondrial genomes of mammalian species between the current study and \citet{crits2023genetic} for each sequencing run shared across the studies.
-The correlation is calculated only across species for which \citet{crits2023genetic} provide read counts in the third supplementary figure of their report, and only for runs for which both studies find $\ge$50 reads mapping to that set of species.
-See \url{https://jbloom.github.io/Huanan_market_samples/crits_christoph_vs_current_run_corr.html} for an interactive version of this plot that enables mouseover of points for details on specific runs, and adjustment of the threshold for how many total aligned mitochondrial reads a run must contain to be shown in this plot.
-\label{fig:crits_christoph_vs_current_run_corr}
-}
-\end{figure*}
 
@@
 Some key species are labeled; see the interactive version at \url{https://jbloom.github.io/Huanan_market_samples/overall_corr.html} to mouseover all points for details, select different subsets of samples, and calculate the correlations on a linear or log scale.
+See Figure~\ref{fig:theil_sen_corr} for a version of this plot that uses correlations based on a Theil-Sen estimator that is more robust to outliers.
 \label{fig:overall_corr}
@@
 See the interactive version of Figure~\ref{fig:overall_corr} at \url{https://jbloom.github.io/Huanan_market_samples/overall_corr.html} to explore other sample subsets.
+The trends are broadly similar if the correlations are instead calculated based on a Theil-Sen estimator~\citep{theilsen} that is more robust to outliers (Figure~\ref{fig:theil_sen_corr}).
 
@@
 \begin{figure*}
+\includegraphics[width=\linewidth]{figures/crits_christoph_vs_current_run_corr.pdf}
+\caption{
+Correlation of the number of read counts mapping to mitochondrial genomes of mammalian species between the current study and \citet{crits2023genetic} for each sequencing run shared across the studies.
+The correlation is calculated only across species for which \citet{crits2023genetic} provide read counts in the third supplementary figure of their report, and only for runs for which both studies find $\ge$50 reads mapping to that set of species.
+See \url{https://jbloom.github.io/Huanan_market_samples/crits_christoph_vs_current_run_corr.html} for an interactive version of this plot that enables mouseover of points for details on specific runs, and adjustment of the threshold for how many total aligned mitochondrial reads a run must contain to be shown in this plot.
+\label{fig:crits_christoph_vs_current_run_corr}
+}
+\end{figure*}
+
+\clearpage
+
+\begin{figure*}
 \includegraphics[width=\linewidth]{figures/per_species_corr_faceted_later_dates.png}
@@
 \end{figure*}
+
+\clearpage
+
+\begin{figure}
+\centering
+\includegraphics[width=0.6\linewidth]{figures/theil_sen_corr/theil_sen_corr.png}
+\caption{
+A version of Figure~\ref{fig:overall_corr} where the y-axis shows a correlation computed from the Theil-Sen estimator~\citep{theilsen} rather than the Pearson correlation.
+The Theil-Sen estimator is more robust to outliers, so species where the correlation is driven by just a few species will have values of zero in these plots.
+See \url{https://jbloom.github.io/Huanan_market_samples/theil_sen_corr.html} for an interactive version of this plot.
+\label{fig:theil_sen_corr}
+}
+\end{figure}
 

### 2023-07-03 paper/reviewer_response.tex
address last few comments

@@
 
+\response{
+For the first question, it isn't actually necessary to assemble contigs since the SARS-CoV-2 reference genome is known and reads can just be aligned to that.
+However, most of the samples have so few viral reads that most of the genome isn't covered.
+From the environmental sample metagenomic sequencing, only sample B5 has decent coverage over the viral genome.
+However, the Chinese CDC did do viral enrichment and sequencing on some of the samples in data that are not analyzed here (as the viral enrichment means those samples are not suitable for metagenomics).
+The results of that analysis are in their \textit{Nature} paper (Liu et al, 2023).
+
+I did not attempt to analyze the presence of other viral pathogens.
+This is an interesting question, but beyond the scope of the current study.
+}
+
 \section*{Associate Editor: Koelle, Katia}
@@
 
+\response{
+Thanks for the helpful and timely reviews.
+In the revised manuscript, I have addressed all of the major comments from the reviewers, including clarification of details about the analysis and more discussion of studies related to the epidemiology of the market.
+I have made sure that the paper clearly emphasizes that these samples are simply not informative either way for determining the source of the virus in the market.
+}
+
 \color{black}

### 2023-07-03 paper/paper_original_submission.tex
add track changes version of revised manuscript

--- 2023-07-02 paper/paper.tex
@@
 I validate the presence of genetic material from numerous species, and calculate mammalian mitochondrial compositions similar to those reported by \citet{crits2023genetic}.
-However, the SARS-CoV-2 content of the environmental samples is generally very low: only 21 of 176 samples contain more than 10 SARS-CoV-2 reads, despite most samples being sequenced to depths exceeding $10^8$  total reads.
-None of the samples with double-digit numbers of SARS-CoV-2 reads have a substantial fraction of their mitochondrial material from any non-human susceptible species.
-Only one of the 14 samples with at least a fifth of the chordate mitochondrial material from raccoon dogs contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
+However, the number of SARS-CoV-2 reads is not consistently correlated with reads mapping to non-human susceptible species.
+For instance, 14 samples have $>$20\% of their chordate mitochondrial material from raccoon dogs, but only one of these samples contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
 Instead, SARS-CoV-2 reads are most correlated with reads mapping to various fish, such as catfish and largemouth bass.
@@
 But by mid to late January of 2020, it was clear SARS-CoV-2 was spreading from human to human, and had been for some time~\citep{chan2020familial,phan2020importation,li2020early,nishiura2020initial}.
-In addition, Chinese scientists published papers reporting that some of the earliest identified human cases in December 2019 had no link to the Huanan Seafood Market~\citep{huang2020clinical,chen2020epidemiological,zhang2020viral}.
+In addition, Chinese scientists published papers reporting that some of the earliest identified human cases in December 2019 had no link to the Huanan Seafood Market~\citep{huang2020clinical,chen2020epidemiological}.
 Thus began a debate that continues to this day: was the market an initial source of zoonotic infections from animals, or was it simply a superspreading site that amplified earlier human infections from another source~\citep{cohen2020wuhan}?
 
-Some papers have argued that the fact that many of the known human cases from December 2019 are centered around market indicates it was the original source of the virus~\citep{worobey2022huanan}, and that an early two-mutation polytomy in the SARS-CoV-2 phylogenetic tree suggests multiple zoonotic jumps there~\citep{pekar2022molecular}.
-Other papers have argued that reports of earlier cases unlinked to the market~\citep{huang2020clinical,chen2020epidemiological,zhang2020viral,ma2020SCMP} and uncertainties about the completeness of case data provided by the Chinese government~\citep{cohen2022anywhere,hvistendahl2023} mean that the first human infections could have occurred upstream of the December 2019 cases at the market~\citep{pekar2021timing,pipes2021assessing,bloom2021recovery}.
-
-Despite these disputes, it is universally agreed that there were human cases of SARS-CoV-2 at the Huanan Seafood Market by mid December of 2019~\citep{li2020early}, and that some animals susceptible to SARS-CoV-2 were sold at the market~\citep{xiao2021animal}.
-The key unanswered question is if any animals were infected, and if so whether they infected humans or were infected by them.
+It is universally agreed that there were human cases of SARS-CoV-2 at the Huanan Seafood Market by mid December of 2019~\citep{li2020early}.
+It is also known that some animals susceptible to SARS-CoV-2 were sold at the market~\citep{xiao2021animal}.
+The unanswered question is if any animals were infected, and if so whether they infected humans or were infected by them.
 
@@
 They reported that none of the 457 animal samples tested positive for SARS-CoV-2, but that 73 of 923 environmental samples tested positive.
-A significant caveat of the Chinese CDC study is that all the samples were collected on January-1-2020 or later, which is probably \emph{at least} a month after the first human infections in Wuhan~\citep{zhang2020viral,van2020emergence,he2020evaluating,pipes2021assessing,pekar2021timing,odni2022,pekar2022molecular}.
+A significant caveat of the Chinese CDC study is that all the samples were collected on January-1-2020 or later, which is \emph{at least} a month after the first human infections in Wuhan~\citep{odni2022,zhang2020viral,van2020emergence,pipes2021assessing}.
 Despite this significant caveat, the data reported in the Chinese CDC study has been variously interpreted to support arguments that virus in the market was human derived~\citep{liu2022surveillance}, that the outbreak originated from live animals sold in the market~\citep{worobey2022huanan}, or that the virus spread among humans in Mahjong rooms or toilets in the market~\citep{courtier2022sars}.
@@
 The accessions for the 4,170 chordate mitochondrial genomes that formed this set are listed in Table~\ref{tab:mito_retained}.
-The counts of reads mapping uniquely to each mitochondrial genome for each sequencing run and sample are provided in Tables~\ref{tab:mito_composition_by_run} and \ref{tab:mito_composition_by_sample}, respectively.
+The counts of reads mapping to each mitochondrial genome for each sequencing run and sample are provided in Tables~\ref{tab:mito_composition_by_run} and \ref{tab:mito_composition_by_sample}, respectively.
 
@@
 Consistent with analyses by the Chinese CDC~\citep{liu2022surveillance}, I find some correlation between the abundance of SARS-CoV-2 and human genetic material, but this correlation is weaker than for several non-infectable animals and so on its own is insufficient to identify the source of the viral material.
-Mitochondrial material from most susceptible non-human species sold live at the market is negatively correlated with the presence of SARS-CoV-2: for instance, 13 of the 14 samples with at least a fifth of their chordate mitochondrial material from raccoon dogs contain no SARS-CoV-2 reads, and the other sample contains just 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
+Mitochondrial material from most susceptible non-human species sold live at the market is negatively correlated with the presence of SARS-CoV-2: 13 of the 14 samples with at least 20\% of their chordate mitochondrial material from raccoon dogs contain no SARS-CoV-2 reads, and the other sample contains just 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
+Likewise, none of the six samples with at least 20\% bamboo rat mitochondrial material contain any SARS-CoV-2 reads.
 These findings are compatible with the results in \citet{crits2023genetic}, since that study did not report any analysis of SARS-CoV-2 content.
@@
 Therefore, by the time the samples were collected, SARS-CoV-2 had been spread widely across the market by humans regardless of its original source---as evidenced by the results reported here, which show viral genetic material coincident with material from myriad animals ranging from fish to snakes to mammals.
-The first human infections with SARS-CoV-2 in Wuhan probably occurred no later than November of 2019~\citep{zhang2020viral,van2020emergence,he2020evaluating,pipes2021assessing,pekar2021timing,odni2022,pekar2022molecular}, which is over a month before the Chinese CDC reports that it began to collect samples from the market.
+The first human infections with SARS-CoV-2 in Wuhan occurred no later than November of 2019~\citep{odni2022,zhang2020viral,van2020emergence,pipes2021assessing}, which is over a month before the Chinese CDC reports that it began to collect samples from the market.
 For this reason, further insight into the origins and early spread of SARS-CoV-2 will likely require learning more about events or cases that occurred no later than November or early December of 2019.
@@
 To create an mitochondrial genome alignment reference set, I followed a procedure partially analogous to that described by \citet{crits2023genetic}.
-Note that a similar analysis could also be implemented using more standard metagenomic analysis tools such as Kraken2~\citep{wood2019improved}, as was in fact done in the Chinese CDC study~\citep{liu2023surveillance}.
-The reasons for instead implementing the pipeline used here were as follows:
-\begin{itemize}
-\item To ensure comparability to \citet{crits2023genetic} by using an approach that parallels the one they described.
-\item To use a reference set that includes all of the potentially susceptible species sold at the market: many of these species lack full nuclear genomes on RefSeq, and one (Amur hedgehog) doesn't even have a mitochondrial genome on RefSeq.
-Therefore, the manual process used to build the mitochondrial reference set described below was necessary to ensure all these species were in the reference.
-\end{itemize}
-
-For the current study, all mitochondrial genomes in the RefSeq database were downloaded, and then filtered to retain only genomes from the phylum Chordata (in contrast, \citet{crits2023genetic} describe retaining all metazoa mitochondrial genomes).
+All mitochondrial genomes in the RefSeq database were downloaded, and then filtered to retain only genomes from the phylum Chordata (in contrast, \citet{crits2023genetic} describe retaining all metazoa mitochondrial genomes).
 The reason I limited to chordate rather than metazoa mitochondrial genomes is that all chordates have very similar length mitochondrial genomes, but at the level of metazoa there is wide variation in mitochondrial genome length.
-The Amur hedgehog mitochondrial genome was not present in this set, so it was separately downloaded from Genbank accession KX964606.
+The raccoon dog mitochondrial genome was not present in this set, so it was separately downloaded from Genbank accession KX964606.
 All the mitochondrial genomes were then filtered to remove highly similar ones.
@@
 The resulting alignments were filtered to only retain primary alignments with mapping quality of at least 4.
-Note that one effect of counting only primary alignments with mapping quality $>$0 (4 in this case) is to only count reads that map uniquely---that is, they map better to the mitochondrial genome of one species than all other species in the reference.
-This is because although \texttt{minimap2} assigns the primary mapping randomly  if there are multiple sequences in the reference with equally good mappings, a read will only have a mapping quality that exceeds zero if it maps to one location better than all other locations.
 
@@
 An initial version of this preprint was revised in part based on helpful comments posted on \textit{bioRxiv} by Florence Débarre and Alex Crits-Christoph (see appendix posted as supplementary material).
-I thank Benjamin F. Pierce for identifying an error in the reference genome description in the original version (\url{https://github.com/jbloom/Huanan_market_samples/pull/3}).
 This study utilized the Fred Hutch Scientific Computing infrastructure, which is supported in part by NIH grants S10-OD-020069 and S10-OD-028685.
@@
 \section{Competing interests}
-JDB is on the scientific advisory boards of Apriori Bio, Aerium Therapeutics, Invivyd, and the Vaccine Company; consults for GSK; and receives royalty payments as an inventor on Fred Hutch licensed patents related to deep mutational scanning of viral proteins.
+JDB is on the scientific advisory boards of Apriori Bio, Aerium Therapeutics, Invivyd, the Vaccine Company, and Oncorus; consults for GSK; and receives royalty payments as an inventor on Fred Hutch licensed patents related to deep mutational scanning of viral proteins.
 

### 2023-07-03 paper/paper.tex
revisions for Virus Evolution (#4)
* acknowledge Ben Pierce

* add reviewer comments

* Reviewer 1 comment 1: better caveat date of emergence estimates

* Reviewer 1 point 2: discuss epi of market more

* Reviewer point 3: de-emphasize 20% cutoff

* Reviewer 1 point 4: samples inconclusive on origins

* Reviewer 2 point 1

* Reviewer 2 point 2: methods rationale

* Reviewer 2 point 3: mapping quality ensures uniquely mapped only

* update competing interests

* Reviewer 2 point 4: consistency of calling samples + vs -

* Reviewer 2 comment 3: move figure to supplement

* code for Reviewer 2 co

@@
 Note that a similar analysis could also be implemented using more standard metagenomic analysis tools such as Kraken2~\citep{wood2019improved}, as was in fact done in the Chinese CDC study~\citep{liu2023surveillance}.
-The reasons for instead implementing the pipeline used here were as follows:
-\begin{itemize}
-\item To ensure comparability to \citet{crits2023genetic} by using an approach that parallels the one they described.
-\item To use a reference set that includes all of the potentially susceptible species sold at the market: many of these species lack full nuclear genomes on RefSeq, and one (Amur hedgehog) doesn't even have a mitochondrial genome on RefSeq.
+The reasons for instead implementing the pipeline used here were two-fold.
+First, to ensure comparability to \citet{crits2023genetic} by using an approach that parallels the one they described.
+Second, to use a reference set that includes all of the potentially susceptible species sold at the market: many of these species lack full nuclear genomes on RefSeq, and one (Amur hedgehog) doesn't even have a mitochondrial genome on RefSeq.
 Therefore, the manual process used to build the mitochondrial reference set described below was necessary to ensure all these species were in the reference.
-\end{itemize}
 

### 2023-07-13 paper/paper.tex
clarify question re info lacking in 2022 CDC preprint

@@
 The 2022 Chinese CDC preprint simply reported that the number of SARS-CoV-2 reads in deep sequencing of the samples was correlated with the number of human reads (fourth figure of \citet{liu2022surveillance}), but did not specify correlations for other species and did not provide the raw sequencing data.
-Other scientists pointed out that if the raw data were shared, it would be possible to expand upon the analysis in the 2022 Chinese CDC preprint to determine if the abundance of SARS-CoV-2 genetic material correlated with material from other species~\citep{cohen2022anywhere,cohen2022studies}.
+Other scientists pointed out that if the raw data were shared, it would be possible to expand upon the analysis in the 2022 Chinese CDC preprint to determine what other animal species contributed genetic material to environmental samples containing SARS-CoV-2~\citep{cohen2022anywhere,cohen2022studies}.
 

### 2023-07-13 paper/final_submission/paper.tex
made final submission version splitting main text TEX file and supp PDF

--- 2023-07-13 paper/paper.tex
@@
 To determine the relationship between the 696 FASTQ files released by the Chinese CDC on the NGDC on March-29-2023 and the 227 files downloaded earlier that month by \citet{crits2023genetic}, I computed the SHA-512 hashes for all the NGDC FASTQ files and compared them to the hashes reported in Table S1 of \citet{crits2023genetic}.
-This analysis confirmed that the dataset released on the NGDC contained unmodified versions of all 227 FASTQ files downloaded by \citet{crits2023genetic}, plus an additional 469 FASTQ files (Tables~\ref{tab:sha526} and \ref{tab:metadata}).
+This analysis confirmed that the dataset released on the NGDC contained unmodified versions of all 227 FASTQ files downloaded by \citet{crits2023genetic}, plus an additional 469 FASTQ files (Tables S1 and S2).
 
@@
 To determine the metagenomic content of the samples, I performed an analysis that conceptually parallels that described by \citet{crits2023genetic}.
-Throughout this paper, I report results only for sequencing described by the Chinese CDC as ``RNA sequencing of total nucleic acids from environmental swabs for metagenomics,'' and exclude the small number of samples that involve amplicon-based sequencing of SARS-CoV-2 or RNA sequencing for viral whole-genome assembly (Table~\ref{tab:metadata}).
+Throughout this paper, I report results only for sequencing described by the Chinese CDC as ``RNA sequencing of total nucleic acids from environmental swabs for metagenomics,'' and exclude the small number of samples that involve amplicon-based sequencing of SARS-CoV-2 or RNA sequencing for viral whole-genome assembly (Table S2).
 
@@
 This set included all the mammalian mitochondrial genomes from all species for which alignments were reported in \citet{crits2023genetic}.
-The accessions for the 4,170 chordate mitochondrial genomes that formed this set are listed in Table~\ref{tab:mito_retained}.
-The counts of reads mapping uniquely to each mitochondrial genome for each sequencing run and sample are provided in Tables~\ref{tab:mito_composition_by_run} and \ref{tab:mito_composition_by_sample}, respectively.
+The accessions for the 4,170 chordate mitochondrial genomes that formed this set are listed in Table S3.
+The counts of reads mapping uniquely to each mitochondrial genome for each sequencing run and sample are provided in Tables S4 and S5, respectively.
 
@@
 Note that while \citet{crits2023genetic} describe aligning the sequencing data to all metazoa mitochondria, they only report alignment counts for mammals.
-The aligned read counts for mammals from my analysis are highly correlated with those reported by \citet{crits2023genetic} for all sequencing runs with a reasonable number of total counts (Figure~\ref{fig:crits_christoph_vs_current_run_corr}).
+The aligned read counts for mammals from my analysis are highly correlated with those reported by \citet{crits2023genetic} for all sequencing runs with a reasonable number of total counts (Figure S1).
 My analysis and that of \citet{crits2023genetic} are independently implemented, and use different alignment programs, reference genome sets, and quality filtering criteria---so the fact that the two analyses give highly correlated results is reassuring.
@@
 \begin{figure*}
-\includegraphics[width=\linewidth]{figures/composition.pdf}
+\includegraphics[width=\linewidth]{fig1.pdf}
 \caption{
@@
 \subsection{Most environmental samples contain little or no SARS-CoV-2 reads}
-To quantify the SARS-CoV-2 content of the samples, I plotted the percentage of all high-quality reads that aligned to SARS-CoV-2 (Figure~\ref{fig:sars2_aligned}, Table~\ref{tab:sars2_aligned_by_run}, Table~\ref{tab:sars2_aligned_by_sample}).
+To quantify the SARS-CoV-2 content of the samples, I plotted the percentage of all high-quality reads that aligned to SARS-CoV-2 (Figure~\ref{fig:sars2_aligned}, Table S6, Table S7).
 Over three-quarters of the samples had no reads aligning to SARS-CoV-2, and most of the rest only had a small number of viral reads.
@@
 \begin{figure*}
-\centerline{\includegraphics[width=0.85\linewidth]{figures/sars2_aligned.pdf}}
+\centerline{\includegraphics[width=0.85\linewidth]{fig2.pdf}}
 \caption{
@@
 Samples with non-zero SARS-CoV-2 reads are in bold.
-See Table~\ref{tab:susceptible_mammal} for a similar table that shows mammalian rather than chordate composition.
-This table uses a 20\% cutoff due to space considerations; to see similar data tabulated for all samples with no cutoff, see the much larger Table~\ref{tab:raccoon_dog_long} (for raccoon dog) and Table~\ref{tab:mito_composition_by_sample} (for all species).
+See Table S8 for a similar table that shows mammalian rather than chordate composition.
+This table uses a 20\% cutoff due to space considerations; to see similar data tabulated for all samples with no cutoff, see the much larger Table S9 (for raccoon dog) and Table S5 (for all species).
 \label{tab:susceptible}
@@
 There was one sample each with at least 20\% chordate mitochondrial composition from Amur hedgehog and Malayan porcupine that contained SARS-CoV-2 reads (Table~\ref{tab:susceptible}).
-If mitochondrial composition is instead analyzed only among mammals rather than chordates, Q61 is still the only sample with at least 20\% raccoon dog composition that contains any SARS-CoV-2 reads (Table~\ref{tab:susceptible_mammal}).
-There is one sample with at least 20\% of its mammalian (but not chordate) mitochondrial composition from bamboo rat that contains a small number of SARS-CoV-2 reads (Table~\ref{tab:susceptible_mammal}).
+If mitochondrial composition is instead analyzed only among mammals rather than chordates, Q61 is still the only sample with at least 20\% raccoon dog composition that contains any SARS-CoV-2 reads (Table S8).
+There is one sample with at least 20\% of its mammalian (but not chordate) mitochondrial composition from bamboo rat that contains a small number of SARS-CoV-2 reads (Table S8).
 
-Note that the 20\% cutoff applied in Tables~\ref{tab:susceptible} and \ref{tab:susceptible_mammal} is not integral to the analysis, but is just a way to subset on samples containing the most material from the species of interest to make the tables small enough to easily visualize.
-See Tables~\ref{tab:raccoon_dog_long} and \ref{tab:mito_composition_by_sample} for much larger tables that show comparable data for all samples with respect to the raccoon dog or overall mitochondrial composition, respectively.
-For instance, Table~\ref{tab:raccoon_dog_long} shows that there are a few samples with lower raccoon dog mitochondrial composition that also contain some SARS-CoV-2 reads.
+Note that the 20\% cutoff applied in Tables~\ref{tab:susceptible} and S8 is not integral to the analysis, but is just a way to subset on samples containing the most material from the species of interest to make the tables small enough to easily visualize.
+See Tables S9 and S5 for much larger tables that show comparable data for all samples with respect to the raccoon dog or overall mitochondrial composition, respectively.
+For instance, Table S9 shows that there are a few samples with lower raccoon dog mitochondrial composition that also contain some SARS-CoV-2 reads.
 However, ultimately these very large tables are difficult to visualize, and so the relationship between SARS-CoV-2 and mitochondrial content for each species for all samples is probably more facilely visualized using the interactive versions of the scatter plots described in the next subsection.
@@
 Furthermore, the correlations are highly contingent on the sample set and details of how the statistics are calculated.
-For instance, if we exclude the first sampling timepoint and only look at samples calculated on January-12-2020 or later, then the most correlated species is the Oriental rat snake, with the correlation for humans becoming even weaker and that for raccoon dogs remaining negative (Figure~\ref{fig:per_species_corr_faceted_later_dates}).
-The correlations also change if only consider samples containing a non-zero number of SARS-CoV-2 reads, or calculate correlations on a linear rather than log scale (Figure~\ref{fig:different_corrs}).
+For instance, if we exclude the first sampling timepoint and only look at samples calculated on January-12-2020 or later, then the most correlated species is the Oriental rat snake, with the correlation for humans becoming even weaker and that for raccoon dogs remaining negative (Figure S2).
+The correlations also change if only consider samples containing a non-zero number of SARS-CoV-2 reads, or calculate correlations on a linear rather than log scale (Figure S3).
 The interactive versions of the correlation plots at \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html} and \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_single.html} enable the reader to explore the correlations for different sample subsets and methods of calculating the statistics.
@@
 \begin{figure*}
-\includegraphics[width=\linewidth]{figures/per_species_corr_faceted.png}
+\includegraphics[width=\linewidth]{fig3.png}
 \caption{
@@
 \begin{figure}
-\includegraphics[width=\linewidth]{figures/overall_corr/overall_corr.pdf}
+\includegraphics[width=\linewidth]{fig4.pdf}
 \caption{
@@
 Some key species are labeled; see the interactive version at \url{https://jbloom.github.io/Huanan_market_samples/overall_corr.html} to mouseover all points for details, select different subsets of samples, and calculate the correlations on a linear or log scale.
-See Figure~\ref{fig:theil_sen_corr} for a version of this plot that uses correlations based on a Theil-Sen estimator that is more robust to outliers.
+See Figure S4 for a version of this plot that uses correlations based on a Theil-Sen estimator that is more robust to outliers.
 \label{fig:overall_corr}
@@
 See the interactive version of Figure~\ref{fig:overall_corr} at \url{https://jbloom.github.io/Huanan_market_samples/overall_corr.html} to explore other sample subsets.
-The trends are broadly similar if the correlations are instead calculated based on a Theil-Sen estimator~\citep{theilsen} that is more robust to outliers (Figure~\ref{fig:theil_sen_corr}).
+The trends are broadly similar if the correlations are instead calculated based on a Theil-Sen estimator~\citep{theilsen} that is more robust to outliers (Figure S4).
 
@@
 Now that the full data are available, we can examine the criteria used to classify samples as positive or negative for SARS-CoV-2.
-It appears that the Chinese CDC classified samples as positive if they met \emph{either} of two criteria: they tested positive for SARS-CoV-2 by RT-qPCR, \emph{or} they were metagenomically sequenced and contained at least one read mapping to SARS-CoV-2 (Table~\ref{tab:rt-qpcr} gives sequencing and RT-qPCR data for all samples that were either sequenced or called positive by the Chinese CDC).
+It appears that the Chinese CDC classified samples as positive if they met \emph{either} of two criteria: they tested positive for SARS-CoV-2 by RT-qPCR, \emph{or} they were metagenomically sequenced and contained at least one read mapping to SARS-CoV-2 (Table S10 gives sequencing and RT-qPCR data for all samples that were either sequenced or called positive by the Chinese CDC).
 However, these criteria are not consistent because not all samples were analyzed by both methods, and the differences in the number of SARS-CoV-2 reads in the sequencing data are often not statistically significant between samples classified as positive versus negative.
@@
 After retaining all of these manually specified mitochondrial genomes, the pipeline then greedily iterated through all other genomes choosing for retention each genome that was not within a mash distance of 0.07 of another already retained genome.
-The full set of retained mitochondrial genomes is listed in Table~\ref{tab:mito_retained}.
+The full set of retained mitochondrial genomes is listed in Table S3.
 
@@
 I then used \texttt{CoverM} (\url{https://github.com/wwood/CoverM}) to compute the number of aligned reads for each run, requiring alignment lengths of at least 40 nucleotides with at least 95\% identity, and excluding the 100 nucleotides at the contig ends.
-The resulting statistics on alignment counts to SARS-CoV-2 and the mitochondrial genomes for each sequencing run are in Tables~\ref{tab:mito_composition_by_run} and \ref{tab:sars2_aligned_by_run}.
-To get per-sample alignment counts, I aggregated counts for all runs for each sample to get the statistics in Tables~\ref{tab:mito_composition_by_sample} and \ref{tab:sars2_aligned_by_sample}.
+The resulting statistics on alignment counts to SARS-CoV-2 and the mitochondrial genomes for each sequencing run are in Tables S4 and S6.
+To get per-sample alignment counts, I aggregated counts for all runs for each sample to get the statistics in Tables S5 and S7.
 For the mitochondrial genome compositions and all other analyses in this paper, I only included only the samples from metagenomic sequencing, which are annotated with the description ``RNA sequencing of total nucleic acids from environmental swabs for metagenomics'' in the metadata provided by \citet{liu2023surveillance}.
@@
 
-\newgeometry{left=1.2in, right=1.2in, top=48pt, bottom=60pt}
-\onecolumn
-\renewcommand{\thepage}{S\arabic{page}}
-\setcounter{page}{1}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-\renewcommand{\thetable}{S\arabic{table}}
-\setcounter{table}{0}
-
-\clearpage
-
-\begin{figure*}
-\includegraphics[width=\linewidth]{figures/crits_christoph_vs_current_run_corr.pdf}
-\caption{
-Correlation of the number of read counts mapping to mitochondrial genomes of mammalian species between the current study and \citet{crits2023genetic} for each sequencing run shared across the studies.
-The correlation is calculated only across species for which \citet{crits2023genetic} provide read counts in the third supplementary figure of their report, and only for runs for which both studies find $\ge$50 reads mapping to that set of species.
-See \url{https://jbloom.github.io/Huanan_market_samples/crits_christoph_vs_current_run_corr.html} for an interactive version of this plot that enables mouseover of points for details on specific runs, and adjustment of the threshold for how many total aligned mitochondrial reads a run must contain to be shown in this plot.
-\label{fig:crits_christoph_vs_current_run_corr}
-}
-\end{figure*}
-
-\clearpage
-
-\begin{figure*}
-\includegraphics[width=\linewidth]{figures/per_species_corr_faceted_later_dates.png}
-\caption{
-Correlation between percent of all reads mapping to SARS-CoV-2 and the mitochondrial genome of each of the indicated species only among samples collected after the first (Jan-1-2020) sampling date.
-Except for the fact that this figure excludes samples from the first date, this figure is otherwise the same as Figure~\ref{fig:per_species_corr_faceted}.
-See \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html} for an interactive version of this plot with numerous options, including allowing further refinement of the sample date range.
-The plots shown here include only samples with at least 200 aligned mitochondrial reads; that option can be adjusted in the interactive plots.
-\label{fig:per_species_corr_faceted_later_dates}
-}
-\end{figure*}
-
-\clearpage
-
-\begin{figure*}
-\includegraphics[width=\linewidth]{figures/different_corrs/correlations.pdf}
-\caption{
-The correlations between the amount of SARS-CoV-2 and mitochondrial genetic material in the samples depends on how the correlation is calculated.
-The top row shows correlations for human, and the bottom row shows correlations for raccoon dog.
-Each column shows the correlations calculated different ways: with a log or linear scale, calculating the mitochondrial percent to be of all reads or just reads mapping to the mitochondria, and showing all samples or only those with at least one SARS-CoV-2 read.
-See \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_single.html} for interactive versions of the plot panels.
-The plots shown here include only samples with at least 200 aligned mitochondrial reads; that option can be adjusted in the interactive plots.
-\label{fig:different_corrs}
-}
-\end{figure*}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=0.6\linewidth]{figures/theil_sen_corr/theil_sen_corr.png}
-\caption{
-A version of Figure~\ref{fig:overall_corr} where the y-axis shows a correlation computed from the Theil-Sen estimator~\citep{theilsen} rather than the Pearson correlation.
-The Theil-Sen estimator is more robust to outliers, so species where the correlation is driven by just a few species will have values of zero in these plots.
-See \url{https://jbloom.github.io/Huanan_market_samples/theil_sen_corr.html} for an interactive version of this plot.
-\label{fig:theil_sen_corr}
-}
-\end{figure}
-
-\clearpage
-
-\begin{table}
-\caption{
-See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/crits_christoph_data/check_sha512_vs_crits_christoph.csv} for the correspondence between the FASTQ files analyzed by \citet{crits2023genetic} and the files deposited in the NGDC by \citet{liu2023surveillance}.
-The files are matched based on their SHA-526 hashes, which were computed directly for the NGDC files as part of the current study, and were taken from Table S1 of \citet{crits2023genetic} for that study.
-\label{tab:sha526}
-}
-\end{table}
-
-\begin{table}
-\caption{
-See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/metadata/merged_metadata.csv} for the metadata for all of the samples and files uploaded to the NGDC by \citet{liu2023surveillance} under accession CRA010170.
-\label{tab:metadata}
-}
-\end{table}
-
-\begin{table}
-\caption{
-See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/mitochondrial_genomes/retained.csv} for the NCBI accessions of the set of chordate mitochondrial genomes that was used as the alignment reference for the metagenomic analysis.
-\label{tab:mito_retained}
-}
-\end{table}
-
-\begin{table}
-\caption{
-See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/aggregated_counts/mito_composition_by_run.csv} for the counts of reads aligning to each mitochondrial genome for each sequencing run.
-\label{tab:mito_composition_by_run}
-}
-\end{table}
-
-\begin{table}
-\caption{
-See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/aggregated_counts/mito_composition_by_sample.csv} for the counts of reads aligning to each mitochondrial genome for each sample.
-\label{tab:mito_composition_by_sample}
-}
-\end{table}
-
-\begin{table}
-\caption{
-See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/aggregated_counts/sars2_aligned_by_run.csv} for the counts of reads aligning to SARS-CoV-2 for each sequencing run.
-\label{tab:sars2_aligned_by_run}
-}
-\end{table}
-
-\begin{table}
-\caption{
-See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/aggregated_counts/sars2_aligned_by_sample.csv} for the counts of reads aligning to SARS-CoV-2 for each sequencing sample.
-\label{tab:sars2_aligned_by_sample}
-}
-\end{table}
-
-\begin{table}
-\caption{
-Reads mapping to SARS-CoV-2 out of all high-quality (pre-processed) reads for samples with $\ge$20\% of their mammalian mitochondrial composition from a susceptible non-human species as defined in \citet{crits2023genetic}.
-Samples with non-zero SARS-CoV-2 reads are in bold.
-See Table~\ref{tab:susceptible} for a similar table that shows chordate rather than mammalian composition.
-This table uses a 20\% cutoff due to space considerations; to see similar data tabulated for all samples with no cutoff, see the much larger Table~\ref{tab:raccoon_dog_long} (for raccoon dog) and Table~\ref{tab:mito_composition_by_sample} (for all species).
-\label{tab:susceptible_mammal}
-}
-{\scriptsize
-\setlength{\tabcolsep}{5pt}
-\renewcommand{\arraystretch}{1.3}
-\begin{tabular}{llrrr}
-\toprule
-species      &   sample              & \makecell[r]{mammalian \\ mitochondrial \\ reads from \\ species} &  \makecell[r]{reads \\ aligning \\ to SARS2} & \makecell[r]{ total pre-\\processed \\ reads} \\
-\midrule
-\multirow{28}{*}{raccoon dog} & HJ200050-20200112-1 &                                       94\% &                        0 &                     1e+08 \\
-                  & HJ200023-20200112-1 &                                       93\% &                        0 &                   6.9e+07 \\
-                  & HJ200048-20200112-1 &                                       92\% &                        0 &                   1.2e+08 \\
-                  & HJ200019-20200112-1 &                                       90\% &                        0 &                     7e+07 \\
-                  & {\bf Q61} &                                       {\bf 81\%} &                        {\bf 1} &                   {\bf 2.1e+08} \\
-                  & HJ200017-20200112-1 &                                       81\% &                        0 &                   1.1e+08 \\
-                  & 629-3-C &                                       78\% &                        0 &                   2.5e+08 \\
-                  & HJ200044-20200112-1 &                                       76\% &                        0 &                   1.2e+08 \\
-                  & HJ200014-20200112-1 &                                       70\% &                        0 &                   1.1e+08 \\
-                  & HJ200047-20200112-1 &                                       66\% &                        0 &                   1.4e+08 \\
-                  & HJ200011-20200112-1 &                                       61\% &                        0 &                   5.8e+07 \\
-                  & HJ200018-20200112-1 &                                       53\% &                        0 &                   1.4e+08 \\
-                  & HJ200013-20200112-1 &                                       50\% &                        0 &                   1.1e+08 \\
-                  & HJ200012-20200112-1 &                                       50\% &                        0 &                   1.3e+08 \\
-                  & HJ200043-20200112-1 &                                       48\% &                        0 &                   8.3e+07 \\
-                  & HJ200027-20200112-1 &                                       43\% &                        0 &                     9e+07 \\
-                  & HJ200024-20200112-1 &                                       42\% &                        0 &                   9.8e+07 \\
-                  & HJ200001-20200112-1 &                                       38\% &                        0 &                   1.2e+08 \\
-                  & HJ200016-20200112-1 &                                       33\% &                        0 &                   1.2e+08 \\
-                  & HJ200007-20200112-1 &                                       31\% &                        0 &                   1.4e+08 \\
-                  & HJ200020-20200112-1 &                                       31\% &                        0 &                   7.9e+07 \\
-                  & HJ200006-20200112-1 &                                       30\% &                        0 &                   1.3e+08 \\
-                  & HJ200009-20200112-1 &                                       27\% &                        0 &                   1.5e+08 \\
-                  & HJ200045-20200112-1 &                                       27\% &                        0 &                   1.1e+08 \\
-                  & HJ200026-20200112-1 &                                       25\% &                        0 &                     1e+08 \\
-                  & HJ200022-20200112-1 &                                       23\% &                        0 &                   1.2e+08 \\
-                  & HJ200004-20200112-1 &                                       23\% &                        0 &                   2.3e+08 \\
-                  & HJ200065-20200112-1 &                                       22\% &                        0 &                   7.3e+07 \\
-\cline{1-5}
-\multirow{14}{*}{hoary bamboo rat} & 629-13-L &                                       85\% &                        0 &                   1.5e+08 \\
-                  & HJ200041-20200112-1 &                                       56\% &                        0 &                   1.8e+08 \\
-                  & HJ200015-20200112-1 &                                       56\% &                        0 &                   1.4e+08 \\
-                  & HJ200065-20200112-1 &                                       56\% &                        0 &                   7.3e+07 \\
-                  & 629-1-L1 &                                       55\% &                        0 &                   2.5e+08 \\
-                  & 629-5-L4 &                                       54\% &                        0 &                   1.4e+08 \\
-                  & HJ200062-20200112-1 &                                       42\% &                        0 &                   1.8e+08 \\
-                  & HJ200013-20200112-1 &                                       37\% &                        0 &                   1.1e+08 \\
-                  & HJ200049-20200112-1 &                                       31\% &                        0 &                     1e+08 \\
-                  & HJ200046-20200112-1 &                                       28\% &                        0 &                   1.1e+08 \\
-                  & HJ200030-20200112-1 &                                       28\% &                        0 &                   1.1e+08 \\
-                  & HJ200067-20200112-1 &                                       27\% &                        0 &                   7.7e+07 \\
-                  & HJ200025-20200112-1 &                                       21\% &                        0 &                   1.2e+08 \\
-                  & {\bf Q68} &                                       {\bf 20\%} &                        {\bf 6} &                   {\bf 8.7e+07} \\
-\cline{1-5}
-\multirow{15}{*}{Amur hedgehog} & W-8-25-L2 &                                       76\% &                        0 &                   3.1e+08 \\
-                  & HJ200040-20200112-1 &                                       75\% &                        0 &                   1.5e+08 \\
-                  & HJ200038-20200112-1 &                                       63\% &                        0 &                     1e+08 \\
-                  & HJ200039-20200112-1 &                                       62\% &                        0 &                   1.2e+08 \\
-                  & 8-25-CK &                                       62\% &                        0 &                   1.5e+09 \\
-                  & W-8-25-L &                                       59\% &                        0 &                   2.9e+08 \\
-                  & HJ200035-20200112-1 &                                       59\% &                        0 &                   7.2e+07 \\
-                  & {\bf 8-25-M1} &                                      {\bf 53\%} &                      {\bf 24} &                   {\bf 4.4e+08} \\
-                  & HJ200033-20200112-1 &                                       51\% &                        0 &                   1.2e+08 \\
-                  & W-8-25-D2 &                                       50\% &                        0 &                   3.3e+08 \\
-                  & W-8-25-D1 &                                       50\% &                        0 &                   2.9e+08 \\
-                  & HJ200045-20200112-1 &                                       41\% &                        0 &                   1.1e+08 \\
-                  & HJ200032-20200112-1 &                                       35\% &                        0 &                   1.4e+08 \\
-                  & 8-25-M2 &                                       24\% &                        0 &                   1.6e+09 \\
-                  & HJ200036-20200112-1 &                                       21\% &                        0 &                   2.2e+08 \\
-\cline{1-5}
-\multirow{2}{*}{Malayan porcupine} & {\bf Q70} &                                      {\bf 93\%} &                       {\bf 2} &                  {\bf 1.5e+08} \\
-                  & HJ200001-20200112-1 &                                       23\% &                        0 &                   1.2e+08 \\
-\cline{1-5}
-Himalayan marmot & HJ200005-20200112-1 &                                       36\% &                        0 &                   1.2e+08 \\
-\cline{1-5}
-masked palm civet & HJ200042-20200112-1 &                                       22\% &                        0 &                   1.9e+08 \\
-\bottomrule
-\end{tabular}
-}
-\end{table}
-
-\begin{table}
-\caption{
-See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/plots/raccoon_dog_long.csv} for a table giving the raccoon dog mitochondrial composition (both among chordates and mammals) along with the SARS-CoV-2 content for all samples.
-\label{tab:raccoon_dog_long}
-}
-\end{table}
-
-\begin{table}
-\caption{
-See \url{https://github.com/jbloom/Huanan_market_samples/blob/main/results/rt_qpcr/rt_qpcr.csv} for combined RT-qPCR data and number of SARS-CoV-2 reads for all environmental samples that were either RT-qPCR positive or metagenomically sequenced.
-This table was constructed by merging the results for all sequenced samples with all samples listed in the second supplementary table of \citet{liu2023surveillance} (or equivalently the first main table of \citet{liu2022surveillance}), which lists all ``positive samples.''
-Since the Chinese CDC says all environmental samples were tested by RT-qPCR, samples not listed as RT-qPCR positive are presumed to have tested negative by that assay.
-Note that this table does \emph{not} list samples that tested negative by RT-qPCR and were not sequenced.
-\label{tab:rt-qpcr}
-}
-\end{table}
-
-\clearpage
-
-\renewcommand{\thepage}{A\arabic{page}}
-\setcounter{page}{1}
-\renewcommand{\thefigure}{A\arabic{figure}}
-\setcounter{figure}{0}
-\renewcommand{\thetable}{A\arabic{table}}
-\setcounter{table}{0}
-
-\clearpage
-\normalsize
-
-\section{Appendix: response to comments and revision of first version of preprint}
-After the original version of this preprint was posted on \textit{bioRxiv} at \url{https://www.biorxiv.org/content/10.1101/2023.04.25.538336v1}, Florence Débarre and Alex Crits-Christoph posted a set of helpful comments at \url{https://www.biorxiv.org/content/10.1101/2023.04.25.538336v1#comments} using the \textit{bioRxiv} comment feature.
-
-This appendix provides a response to those comments, some of which were used to inform revisions incorporated into this new version of the preprint.
-Note that the computer code / data / paper from the initial version are all also available on GitHub under tag \texttt{bioRxiv\_v1} (\url{https://github.com/jbloom/Huanan_market_samples/tree/bioRxiv_v1}) and the code / data / paper for this revised version are under the tag \texttt{bioRxiv\_v2} (\url{https://github.com/jbloom/Huanan_market_samples/tree/bioRxiv_v2}).
-To see all differences between the two versions, go to \url{https://github.com/jbloom/Huanan_market_samples/compare/bioRxiv_v1...bioRxiv_v2} which shows the differences between the GitHub tags.
-
-In addition to the revisions made in response to the comments below, I also made a minor tweak to ensure that the same \texttt{fastp} pre-processing filters were being applied to both the single- and paired-end sequencing data.
-Overall, none of the modifications to the code led to anything more than minor changes in the overall results.
-
-Below the comments posted on \textit{bioRxiv} by Florence Débarre and Alex Crits-Christoph are in {\bf \color{purple} purple text}, and my responses are in {\bf black text}.
-
-\color{purple}
-
-\subsection{Comments and responses}
-
-In this preprint, Bloom re-analyses a dataset of metagenomic data recently shared on open repositories by \citet{liu2023surveillance}.
-A subset of these data, previously made available on GISAID, was also previously analyzed by \citet{crits2023genetic}---of which we are co-authors.
-
-\begin{enumerate}
-\item Bloom’s analysis largely confirms the genetic identification of wildlife species by \citet{crits2023genetic}.
-
-The identification of animal species differed between \citet{liu2023surveillance} and \citet{crits2023genetic}, in particular regarding the abundance of raccoon dog genetic material in sample Q61.
-Bloom’s analysis, done with similar but independent methods from \citet{crits2023genetic}, largely replicates their findings.
-
-A comprehensive Github repository allows any interested reader to conveniently check Bloom’s results by themselves.
-
-{\color{black}
-I agree that my analysis largely confirm the analysis of the mammalian mitochondrial genetic material in your study.
-As I note in my preprint, it is reassuring to have two independent studies reach largely consistent results on this point.
-As I also note, I think the main value of your earlier work was to provide genetic confirmation of which animals were at the market, and my study basically independently confirms your findings on that point.
-
-I am glad you find the GitHub repository useful!
-}
-
-\item The correlation analysis presented by Bloom contains numerous flaws.
-
-Bloom reproduced the correlation analysis initially presented in \citet{liu2022surveillance}, but Bloom’s analysis, as conducted, is inappropriate to answer the question of which animal hosts shed the viral material detected.
-\citet{crits2023genetic} were missing data to carry out such an analysis, but warned against its use.
-Here we repeat and expand the arguments presented in \citet{crits2023genetic}.
-These points would need to be taken into account for a correlation analysis to be valid, but were not by Bloom (and often just cannot be):
-
-{\color{black}
-At a high level, I agree that the lack of any correlation between SARS-CoV-2 and the genetic material of susceptible animals does not \emph{disprove} the possible existence of an infected animal (for reasons you list below).
-However, it does indicate that the genetic content of these samples is insufficient to reliably indicate whether any animal was infected. That is the claim my preprint makes.
-In the end, if the SARS-CoV-2 content is not associated with the genetic content from susceptible animals, then certainly the genetic material cannot reliably indicate that the animals were infected and shedding the virus.
-
-I think we are in agreement that co-mingling of viral and animal genetic material in these samples is insufficient to indicate if any animals were infected.
-Regardless of perceptions/discussion on social or popular media, your report never claimed proof of infected animals, and my preprint never claimed disproof of infected animals.
-Rather, my understanding is that by fully analyzing the data we are converging on the view that the genetic content of these samples is simply insufficient to reliably determine if any animals were infected.
-}
-
-\begin{enumerate}[label=\alph*)]
-
-\item The outbreak had been ongoing for weeks in the market when the samples were collected, and multiple humans had been infected.
-By then, most virus in the market will come from human cases, especially as stalls containing human cases were targeted first, when sampling was performed on January 1.
-The large number of samples from these stalls with sick humans will highly influence correlational analyses.
-As shown in Worobey et al.
-(2022) on similar data obtained via qPCR, both the distance from known human cases and the distance from wildlife stalls independently contributed to the spatial variation in viral positive samples.
-The number of human cases at the time of sampling means that even if infected animals were present at the market, an underpowered correlation analysis of the sequencing data would be unlikely to have revealed it.
-
-{\color{black} I agree that the late date of sample collection is a major limitation, and mention that several times in my preprint.
-As a say in the preprint, by the time the samples were collected humans had spread the virus widely across the market.
-That means the environmental samples are not particularly informative on the original source of the virus in the market in either direction, which is the central conclusion I draw.}
-
-\item The correlation analysis treats all animals of a given species as an homogeneous group.
-There were however multiple stalls selling the same animal species, but likely not from the same supply sources.
-The presence of uninfected animals in one stall would affect the result of a correlation analysis, but would not invalidate the potential presence of infected animals in another stall.
-
-{\color{black} Certainly I agree that this is a limitation. As I say in the preprint, these environmental samples are not sufficient for reliably indicating if any animals were infected.}
-
-\item The samples themselves are not homogeneous: for instance, within-sample animal diversity is different when the swab was taken directly from fish packaging than from the ground.
-
-In addition, wildlife stall samples tend to be composed of a diverse menagerie of animal species RNA/DNA (a consequence of animals housed directly on top of each other and equipment shared between them), while samples related to human cases elsewhere in the market tend to be fairly simple.
-Therefore, by correlating abundances within just the positive samples, human-shed samples tend to have very high relative abundances for human DNA (or, e.g., human and fish), while animal shed samples tend to have lower abundances for any given species.
-This is a natural consequence of sequencing abundances being relative, not absolute.
-
-{\color{black}
-There is no correlation between the amount of SARS-CoV-2 genetic material and that from potentially susceptible animals like raccoon dogs or bamboo rats regardless of whether the correlation is taken over just SARS-CoV-2 positive samples or all samples.
-This is shown in the new Figure~\ref{fig:overall_corr} and can also be seen in Figure~\ref{fig:per_species_corr_faceted} and perhaps most easily in the interactive version of that figure linked in its legend.
-}
-
-\item SARS-CoV-2 read counts can differ for multiple reasons: there can be different quantities of virus shed in the first place, and different times since virus deposition will influence detection, as RNA degrades over time.
-While it is not possible to precisely date the time of viral deposition, there are two important considerations.
-First, if wildlife introduced the virus first to the market, animal-shed virus would (on average) occur before human sheddings.
-Second, wildlife stalls were sampled multiple days after stalls with reported cases (January 12 vs January 1).
-Put together, this means there would have been substantially more time for virus shed by animals to decay than virus shed by humans.
-
-{
-\color{black}
-There is no correlation between the amount of SARS-CoV-2 and genetic material from potentially susceptible animals like raccoon dogs or bamboo rats regardless of whether the analysis is performed on all samples or restricted to just samples collected on later dates (such as January 12).
-This can be seen in Figure~\ref{fig:overall_corr}, Figure~\ref{fig:per_species_corr_faceted_later_dates}, and perhaps most easily in the interactive plots at  \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html}, which enable subsetting to samples collected on arbitrary dates.
-Of course this fact does not rule out the possibility that animals shed virus earlier and that viral RNA decayed.
-It is also possible animals never shed virus.
-Both possibilities are consistent with my conclusion that these samples are not sufficient to reliably indicate whether any animals were infected.
-}
-
-\item Virus presence is detected through RNA, host presence through mtDNA, which degrade differently over time.
-Relative abundances therefore depend on the time difference between when the different genetic materials were deposited and when they were collected.
-
-{\color{black}
-See response to prior point.
-}
-
-\item Stalls were sampled multiple times. A correlation analysis would need to take into account spatial proximity and in particular model the per-stall effect.
-
-{\color{black}
-At a qualitative level (just for instance looking at the scatter plots), it is obvious that the samples that contain substantial amounts of material from potentially susceptible animals like raccoon dogs or bamboo rats contain no or very little SARS-CoV-2 (see interactive plots at \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_single.html}).
-This is going to remain true no matter how the samples are partitioned or modeled.
-Given this fact, it seems clear that the samples do not suggest shedding by an infected animal. 
-Of course there could have been infected animals earlier or that shed virus that was not collected in these samples, but in either case that makes these samples uninformative for determining if any animal was infected.
-}
-
-\item The correlations reported in Bloom’s figure 5 are calculated solely on samples that contained any positive reads for SARS-CoV-2, and were therefore sequencing-positive for the virus.
-Therefore, this is not a test of which species were more likely to be in samples that tested positive for the virus, but how abundant each species was in the samples that did test positive.
-This is an important distinction.
-
-{\color{black}
-The correlations in Figure~\ref{fig:per_species_corr_faceted} and its interactive version (\url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html}) show all samples, and I have also added panels with all samples to Figure~\ref{fig:overall_corr}.
-It remains the case even when looking at all samples that SARS-CoV-2 content is mostly associated with material from animals that definitely were not infected.
-}
-
-\end{enumerate}
-
-For these reasons, both positive and negative correlations are largely uninterpretable:
-\begin{itemize}
-
-\item A positive correlation is not indicative of infection.
-As an enlightening example (also highlighted in \citet{crits2023genetic}), the most correlated species in Bloom’s analysis is a fish species that is not susceptible to SARS-CoV-2.
-
-\item A negative correlation is not indicative of lack of infection (see above).
-
-\end{itemize}
-
-{\color{black}
-I agree with this essential point: my preprint simply concludes that the genetic content of these samples is insufficient to reliably determine if there were any infected animals.
-}
-
-\item Some samples appear to be missing in the analysis, which contained high viral read numbers: A20, F13, F54.
-
-{\color{black}
-In my analysis, I filter only for samples sequenced metagenomically, and exclude samples that were enriched for virus for viral sequencing.
-Originally, I was filtering all samples that had any virus-enrichment based sequencing.
-In the updated pipeline, for samples like A20 that have both metagenomic and viral sequencing, I now retain the metagenomic sequencing while ignoring the viral sequencing---thanks for pointing out this oversight.
-This adds a few samples including A20, but does not change any of the overall results.
-However, F13 and F54 are still excluded because according to the metadata in the \textit{Experiments} sheet in the Chinese CDC metadata Excel file at \url{https://ngdc.cncb.ac.cn/gsa/browse/CRA010170}, these samples were not sequenced metagenomically (there are no entries with the ``RNA sequencing of total nucleic acids from environmental swabs for metagenomics'' annotation for them in the \textit{Experiments} sheet).
-}
-
-\item Bloom chose to align reads to chordates’ mitochondrial genomes, but did not justify this choice.
-Only mammals are known to be infected by SARS-CoV-2, so an analysis of just mammalian RNA/DNA, as presented in \citet{crits2023genetic}, is most relevant to the question of which species may have shed the virus in a particular sample.
-If the aim was to provide a better picture of the diversity of animals sold at the Huanan market, all metazoa should have been included, to be able to identify seafood for instance (a key early case was a shrimp seller).
-
-However, the end result is that Bloom’s 20\% chordate inclusion threshold for this table results in most of the samples with both wildlife RNA/DNA (e.g.
-raccoon dogs) and viral RNA being excluded.
-The claim that only one sample of all samples with raccoon dog DNA that contains SARS-CoV-2 is therefore inaccurate: please see the supplementary data of either \citet{crits2023genetic} or Bloom et al.
-for comparative results on the true number of such samples (which is at least 5, depending on the given analysis).
-
-{\color{black}
-The reason I limited to chordate rather than metazoa mitochondrial genomes is that all chordates have very similar length mitochondrial genomes, but at the level of metazoa there is wide variation in mitochondrial genome length---this is now explained in the revised Methods.
-
-In any case, the choice of using chordate versus mammalian has no real effect.
-The 20\% cutoff in Table~\ref{tab:susceptible} was just to make the table small enough to easily fit in the main text.
-I have now added a comparable table (Table~\ref{tab:susceptible_mammal}) that lists mammalian rather than chordate composition, and another table that lists SARS-CoV-2 and raccoon dog content for all samples without any cutoff (Table~\ref{tab:raccoon_dog_long}).
-The essential result is unchanged: the samples with abundant raccoon dog mitochondrial material have little or no SARS-CoV-2 reads.
-This is also easily seen in the interactive scatter plots provided in the paper.
-
-There are a few other samples with low raccoon dog content (less than 20\% of either their chordate or mammalian mitochondrial content, see Table~\ref{tab:raccoon_dog_long}) that have some SARS-CoV-2 reads, but these samples also have more abundant genetic material from many other species.
-So certainly they are not sufficient to support an association between viral and raccoon dog material that exceeds that of any of a large number of other species that clearly were not infected (because they are not susceptible).
-This all supports the conclusion that the genetic contents of these samples are insufficient to conclude if any animal was infected.
-}
-
-\item Bloom’s comments on the definition of positive samples overlook the difference in sensitivity between qPCR and sequencing in environmental sampling---qPCR is highly sensitive, with near single molecule detection resolution, and correlates with the absolute load of the virus in the sample.
-Sequencing, however, is both less sensitive (even when deeply sequencing), and highly dependent on the quantity of what else is in the sample.
-Thus, it is very challenging to identify SARS-CoV-2 in samples sequenced from the environment without any kind of enrichment performed: this is because the virus is comparatively rare compared to abundant environmental microbes (bacteria, fungi) as well as DNA/RNA shed by animals and humans.
-This is why most environmental viral sequencing is done after performing a viral enrichment (either physical or amplicon).
-For a similar comparison of unenriched SARS-CoV-2 sequencing, see the supplementary table S1 of Crits-Christoph et al.
-2021 (“Genome Sequencing of Sewage Detects Regionally Prevalent SARS-CoV-2 Variants”), where 5 unenriched samples with viral Ct values in the range of 30-32 were deeply sequenced, and only 1-17 SARS-CoV-2 reads were obtained.
-For this reason, we do not encourage using only sequencing data quantitative values and ignoring qPCR positive samples.
-There is no indication in the \citet{liu2023surveillance} dataset that we have seen to indicate the qPCR results were impacted by contamination.
-Indeed, 112/116 samples that were reported negative for qPCR were also negative by NGS, indicating a high degree of agreement between the two approaches, and that qPCR is merely more sensitive.
-
-In addition, the definition of a positive sample cannot be done independently of the provenance of the sample: even at very low read count, the presence of other positive samples in the same stall at the same date is indicative of sample positivity.
-
-{\color{black}
-I agree that RT-qPCR could be more sensitive than sequencing.
-But the key point is that the ``positive'' sample with the most raccoon dog genetic material (Q61) tested \emph{negative} by RT-qPCR, and was called positive by the Chinese CDC on the basis of having 1 in 200,000,000 reads align to raccoon dog.
-It is not consistent to call that sample ``positive'' if other samples that also tested negative by RT-qPCR but were never sequenced are called ``negative.''
-Samples should only be analyzed together if they were subjected to the same consistent set of assays: so for instance, all samples could be classified by RT-qPCR, or all samples subjected to both assays could be classified based on both.
-This observation is not a criticism of your work: the basis for these classifications only recently became clear with the new \citet{liu2023surveillance} paper.
-I just think we should now consider these points now that they are apparent.
-
-As far as other samples in the same stall, it is clear that SARS-CoV-2 was spread widely across the market: that is why we see it on samples full of material from fish, snakes, etc.
-Therefore, it is my contention that we simply can't conclude much about whether any animal was infected if the virus is spread widely in the environment uncorrelated with any specific animal genetic material.
-This is true regardless of whether we do the analysis at the level of samples or stalls.
-}
-
-\item Citations to news articles are vague or misleading. It would be preferable to stick to arguments presented in scientific documents, rather than rely on newspaper articles.
-
-\begin{enumerate}[label=\alph*)]
-
-\item
-As rationale for his analysis, in the Introduction of his preprint, Bloom claims that “Other scientists pointed out that if the raw data were shared, it would be possible to expand upon the analysis in the 2022 Chinese CDC preprint to determine if the abundance of SARS-CoV-2 genetic material correlated with material from other species (Cohen 2022a,b)”, where the two Cohen citations are news articles.
-The authors quoted in these articles (Robertson, Rambaut, Holmes) did not mention Figure 4a because they thought the analysis was useful enough to be reproduced, but because it was possible that other animals’ genetic material was present in the environmental samples, and these data were not shown.
-
-{
-\color{black}
-The overall rationale of my study was to examine the quantitative association between SARS-CoV-2 content and animal genetic material.
-Figure 4a of the original Chinese CDC study was the only prior work on this topic, and it was widely discussed including in the three Cohen news articles I mentioned, one of which even re-printed the figure.
-I think it is important to provide this as background, and I am careful to distinguish what was discussed in \citet{crits2023genetic} (which did not discuss this figure) versus news articles.
-}
-
-\item
-By conflating their views with misinterpretations of news reports written about their work, Bloom misrepresents the conclusions of \citet{crits2023genetic}:
-\citet{crits2023genetic} did not conclude they had found the intermediate host and it was raccoon dogs: their report clearly stated that they could not “identify the intermediate animal host species from these data”, and that the presence of infected animals was a “plausible explanation” for their results.
-And they explicitly stated that “the most abundant animal in the sequencing data of a particular sample is not necessarily the source of the virus in that sample”.
-
-{
-\color{black}
-I am careful to delineate that \citet{crits2023genetic} did \emph{not} analyze the SARS-CoV-2 content in their report, and so never claimed that animal and viral co-mingling in the environmental samples indicated an infected animal.
-My preprint mentions several times how it is consistent with the results reported in \citet{crits2023genetic}, but adds more analyses that look at association between animal and viral material.
-
-However, it is true that the news articles I cite \emph{did} talk about co-mingling of material suggesting infected animals, and those were widely read.
-I therefore provide the full context to help disentangle the confusion between what was in the scientific report versus the related news articles, which the scientific authors obviously did not write.
-}
-
-\end{enumerate}
-
-\end{enumerate}
-
 \end{document}

### 2023-07-13 paper/final_submission/paper.tex
move references from BBL to TEX file

@@
 
-\bibliography{references}
+\begin{thebibliography}{44}
+
+\bibitem[{Bloom}(2021)]{bloom2021recovery}
+{Bloom, J.~D.}, 2021 Recovery of deleted deep sequencing data sheds more light
+  on the early {Wuhan SARS-CoV-2} epidemic. Molecular Biology and Evolution
+  {\bf 38}: 5211--5224.
+
+\bibitem[{Chan {\em et~al.\/}}(2020){Chan, Yuan, Kok, To, Chu, Yang, Xing, Liu,
+  Yip, Poon, {\em et~al.\/}}]{chan2020familial}
+{Chan, J. F.-W., S.~Yuan, K.-H. Kok, K.~K.-W. To, H.~Chu, {\em et~al.\/}}, 2020
+  A familial cluster of pneumonia associated with the 2019 novel coronavirus
+  indicating person-to-person transmission: a study of a family cluster. The
+  Lancet {\bf 395}: 514--523,
+  \url{https://www.thelancet.com/journals/lancet/article/PIIS0140-6736(20)30154-9/fulltext}.
+
+\bibitem[{Chen {\em et~al.\/}}(2020){Chen, Zhou, Dong, Qu, Gong, Han, Qiu,
+  Wang, Liu, Wei, {\em et~al.\/}}]{chen2020epidemiological}
+{Chen, N., M.~Zhou, X.~Dong, J.~Qu, F.~Gong, {\em et~al.\/}}, 2020
+  Epidemiological and clinical characteristics of 99 cases of 2019 novel
+  coronavirus pneumonia in {Wuhan, China}: a descriptive study. Lancet {\bf
+  395}: 507--513,
+  \url{https://www.thelancet.com/journals/lancet/article/PIIS0140-6736(20)30211-7/fulltext}.
+
+\bibitem[{Chen {\em et~al.\/}}(2018){Chen, Zhou, Chen, {\rm and}
+  Gu}]{chen2018fastp}
+{Chen, S., Y.~Zhou, Y.~Chen, {\rm and} J.~Gu}, 2018 fastp: an ultra-fast
+  all-in-one {FASTQ} preprocessor. Bioinformatics {\bf 34}: i884--i890,
+  \url{https://academic.oup.com/bioinformatics/article/34/17/i884/5093234}.
+
+\bibitem[{Cohen}(2020)]{cohen2020wuhan}
+{Cohen, J.}, 2020 Wuhan seafood market may not be source of novel virus
+  spreading globally. Science {\bf 10}: 10.1126/science.abb0611,
+  \url{https://www.science.org/content/article/wuhan-seafood-market-may-not-be-source-novel-virus-spreading-globally}.
+
+\bibitem[{Cohen}(2022{a})]{cohen2022anywhere}
+{Cohen, J.}, 2022{a} {Anywhere but here: China now insists the pandemic
+  didn’t start within its borders. Its scientists are publishing a flurry of
+  papers pointing the finger elsewhere}. Science
+  \url{https://www.science.org/content/article/pandemic-start-anywhere-but-here-argue-papers-chinese-scientists-echoing-party-line}.
+
+\bibitem[{Cohen}(2022{b})]{cohen2022studies}
+{Cohen, J.}, 2022{b} Studies bolster pandemic origin in {W}uhan animal market.
+  Science {\bf 375}: 946 -- 947,
+  \url{https://www.science.org/doi/epdf/10.1126/science.adb1760}; see Tweet at
+  \url{https://twitter.com/sciencecohen/status/1498658469775298568} for image
+  of figure from 2022 Chinese CDC pre-print.
+
+\bibitem[{Cohen}(2023{a})]{cohen2023new}
+{Cohen, J.}, 2023{a} {A new pandemic origin report is stirring controversy.
+  Here are key takeaways}. Science,
+  \url{https://www.science.org/content/article/covid-origin-report-controversy}.
+
+\bibitem[{Cohen}(2023{b})]{cohen2023unearthed}
+{Cohen, J.}, 2023{b} {Unearthed genetic sequences from China market may point
+  to animal origin of COVID-19}. Science
+  \url{https://www.science.org/content/article/covid-19-origins-missing-sequences}.
+
+\bibitem[{Courtier-Orgogozo {\rm and} de~Ribera}(2022)]{courtier2022sars}
+{Courtier-Orgogozo, V. {\rm and} F.~A. de~Ribera}, 2022 {SARS-CoV-2 infection
+  at the Huanan seafood market}. Environmental Research {\bf 214}: 113702,
+  \url{https://www.sciencedirect.com/science/article/pii/S0013935122010295}.
+
+\bibitem[{Crits-Christoph {\em et~al.\/}}(2023){Crits-Christoph, Gangavarapu,
+  Pekar, Moshiri, Singh, Levy, Goldstein, Suchard, Popescu, Robertson, {\em
+  et~al.\/}}]{crits2023genetic}
+{Crits-Christoph, A., K.~Gangavarapu, J.~E. Pekar, N.~Moshiri, R.~Singh, {\em
+  et~al.\/}}, 2023 {Genetic evidence of susceptible wildlife in SARS-CoV-2
+  positive samples at the Huanan Wholesale Seafood Market, Wuhan: Analysis and
+  interpretation of data released by the Chinese Center for Disease Control}.
+  Zenodo {\bf DOI 10.5281/zenodo.7754299},
+  \url{https://zenodo.org/record/7754299#.ZEUz_uzMLX0}.
+
+\bibitem[{GISAID}(2023)]{gisaid_statement}
+{GISAID}, 2023 {GISAID comments on the speculations surrounding data
+  availability}.
+  \url{https://gisaid.org/statements-clarifications/data-availability/}.
+
+\bibitem[{Grabherr {\em et~al.\/}}(2011){Grabherr, Haas, Yassour, Levin,
+  Thompson, Amit, Adiconis, Fan, Raychowdhury, Zeng, {\em
+  et~al.\/}}]{grabherr2011trinity}
+{Grabherr, M.~G., B.~J. Haas, M.~Yassour, J.~Z. Levin, D.~A. Thompson, {\em
+  et~al.\/}}, 2011 Trinity: reconstructing a full-length transcriptome without
+  a genome from {RNA-Seq} data. Nature Biotechnology {\bf 29}: 644,
+  \url{https://www.nature.com/articles/nbt.1883}.
+
+\bibitem[{He {\rm and} Dunn}(2020)]{he2020evaluating}
+{He, M. {\rm and} L.~F. Dunn}, 2020 Evaluating incidence and impact estimates
+  of the coronavirus outbreak from official and non-official chinese data
+  sources. SSRN {\bf DOI 10.2139/ssrn.3540636},
+  \url{https://papers.ssrn.com/sol3/Papers.cfm?abstract_id=3540636}.
+
+\bibitem[{Huang {\em et~al.\/}}(2020){Huang, Wang, Li, Ren, Zhao, Hu, Zhang,
+  Fan, Xu, Gu, {\em et~al.\/}}]{huang2020clinical}
+{Huang, C., Y.~Wang, X.~Li, L.~Ren, J.~Zhao, {\em et~al.\/}}, 2020 Clinical
+  features of patients infected with 2019 novel coronavirus in {Wuhan, China}.
+  Lancet {\bf 395}: 497--506,
+  \url{https://www.thelancet.com/article/S0140-6736(20)30183-5/fulltext}.
+
+\bibitem[{Hvistendahl {\rm and} Mueller}(2023)]{hvistendahl2023}
+{Hvistendahl, M. {\rm and} B.~Mueller}, 2023 Chinese censorship is quietly
+  rewriting the {Covid-19} story. The New York Times
+  \url{https://www.nytimes.com/2023/04/23/world/europe/chinese-censorship-covid.html}.
+
+\bibitem[{Kang {\em et~al.\/}}(2020){Kang, Cheng, {\rm and}
+  McNeil}]{kang2020china}
+{Kang, D., M.~Cheng, {\rm and} S.~McNeil}, 2020 China clamps down in hidden
+  hunt for coronavirus origins.
+  \url{https://apnews.com/article/united-nations-coronavirus-pandemic-china-only-on-ap-bats-24fbadc58cee3a40bca2ddf7a14d2955},
+  The actual China State Council order described in the article is at
+  \url{https://www.documentcloud.org/documents/7340337-State-Research-regulations.html}.
+
+\bibitem[{Li}(2018)]{li2018minimap2}
+{Li, H.}, 2018 Minimap2: pairwise alignment for nucleotide sequences.
+  Bioinformatics {\bf 34}: 3094--3100,
+  \url{https://academic.oup.com/bioinformatics/article/34/18/3094/4994778}.
+
+\bibitem[{Li {\em et~al.\/}}(2020){Li, Guan, Wu, Wang, Zhou, Tong, Ren, Leung,
+  Lau, Wong, {\em et~al.\/}}]{li2020early}
+{Li, Q., X.~Guan, P.~Wu, X.~Wang, L.~Zhou, {\em et~al.\/}}, 2020 {Early
+  transmission dynamics in Wuhan, China, of novel coronavirus--infected
+  pneumonia}. New England Journal of Medicine {\bf 382}: 1199--1207,
+  \url{https://www.nejm.org/doi/full/10.1056/nejmoa2001316}.
+
+\bibitem[{Liu {\em et~al.\/}}(2022){Liu, Liu, Lei, Jia, He, Liu, Shi, Tan, Zou,
+  {\em et~al.\/}}]{liu2022surveillance}
+{Liu, W., P.~Liu, W.~Lei, Z.~Jia, X.~He, {\em et~al.\/}}, 2022 {Surveillance of
+  SARS-CoV-2 in the environment and animal samples of the Huanan Seafood
+  Market}. Research Square {\bf DOI 10.21203/rs.3.rs-1370392/v1},
+  \url{https://www.researchsquare.com/article/rs-1370392/v1}.
+
+\bibitem[{Liu {\em et~al.\/}}(2023{a}){Liu, Liu, Lei, Jia, He, Shi, Tan, Zou,
+  Wong, Wang, {\em et~al.\/}}]{liu2023surveillance}
+{Liu, W.~J., P.~Liu, W.~Lei, Z.~Jia, X.~He, {\em et~al.\/}}, 2023{a}
+  {Surveillance of SARS-CoV-2 at the Huanan Seafood Market}. Nature {\bf DOI:
+  10.1038/s41586-023-06043-2},
+  \url{https://www.nature.com/articles/s41586-023-06043-2}.
+
+\bibitem[{Liu {\em et~al.\/}}(2023{b}){Liu, Liu, Lei, Jia, He, Shi, Tan, Zou,
+  Wong, Wang, {\em et~al.\/}}]{liu2023surveillance-chinaxiv}
+{Liu, W.~J., P.~Liu, W.~Lei, Z.~Jia, X.~He, {\em et~al.\/}}, 2023{b}
+  {Surveillance of SARS-CoV-2 at the Huanan Seafood Market}. ChinaXiv {\bf DOI:
+  10.12074/202303.10351V1}, \url{https://chinaxiv.org/abs/202303.10351V1}.
+
+\bibitem[{Ma}(2020)]{ma2020SCMP}
+{Ma, J.}, 2020 {Coronavirus: China’s first confirmed Covid-19 case traced
+  back to November 17}. South China Morning Post,
+  \url{https://www.scmp.com/news/china/society/article/3074991/coronavirus-chinas-first-confirmed-covid-19-case-traced-back},
+  archived at
+  \url{https://web.archive.org/web/20200315011702/https://www.scmp.com/news/china/society/article/3074991/coronavirus-chinas-first-confirmed-covid-19-case-traced-back}.
+
+\bibitem[{M{\"o}lder {\em et~al.\/}}(2021){M{\"o}lder, Jablonski, Letcher,
+  Hall, Tomkins-Tinch, Sochat, Forster, Lee, Twardziok, Kanitz, {\em
+  et~al.\/}}]{molder2021sustainable}
+{M{\"o}lder, F., K.~P. Jablonski, B.~Letcher, M.~B. Hall, C.~H. Tomkins-Tinch,
+  {\em et~al.\/}}, 2021 Sustainable data analysis with {Snakemake}.
+  F1000Research {\bf 10}, \url{https://f1000research.com/articles/10-33}.
+
+\bibitem[{Mueller}(2023)]{mueller2023nytimes}
+{Mueller, B.}, 2023 New data links pandemic’s origins to raccoon dogs at
+  {Wuhan} market. The New York Times
+  \url{https://www.nytimes.com/2023/03/16/science/covid-wuhan-market-raccoon-dogs-lab-leak.html}.
+
+\bibitem[{Nishiura {\em et~al.\/}}(2020){Nishiura, Linton, {\rm and}
+  Akhmetzhanov}]{nishiura2020initial}
+{Nishiura, H., N.~M. Linton, {\rm and} A.~R. Akhmetzhanov}, 2020 {Initial
+  cluster of novel coronavirus (2019-nCoV) infections in Wuhan, China is
+  consistent with substantial human-to-human transmission}. Journal of Clinical
+  Medicine {\bf 9}: 488, \url{https://www.mdpi.com/2077-0383/9/2/488}.
+
+\bibitem[{{ODNI}}(2022)]{odni2022}
+{{ODNI}}, 2022 Unclassified summary of assessment on {COVID-19} origins.
+  \url{https://www.dni.gov/files/ODNI/documents/assessments/Unclassified-Summary-of-Assessment-on-COVID-19-Origins.pdf}.
+
+\bibitem[{Ondov {\em et~al.\/}}(2016){Ondov, Treangen, Melsted, Mallonee,
+  Bergman, Koren, {\rm and} Phillippy}]{ondov2016mash}
+{Ondov, B.~D., T.~J. Treangen, P.~Melsted, A.~B. Mallonee, N.~H. Bergman, {\em
+  et~al.\/}}, 2016 Mash: fast genome and metagenome distance estimation using
+  {MinHash}. Genome biology {\bf 17}: 1--14,
+  \url{https://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-0997-x}.
+
+\bibitem[{Pekar {\em et~al.\/}}(2021){Pekar, Worobey, Moshiri, Scheffler, {\rm
+  and} Wertheim}]{pekar2021timing}
+{Pekar, J., M.~Worobey, N.~Moshiri, K.~Scheffler, {\rm and} J.~O. Wertheim},
+  2021 Timing the {SARS-CoV-2} index case in hubei province. Science {\bf 372}:
+  412--417, \url{https://www.science.org/doi/full/10.1126/science.abf8003}.
+
+\bibitem[{Pekar {\em et~al.\/}}(2022){Pekar, Magee, Parker, Moshiri,
+  Izhikevich, Havens, Gangavarapu, Malpica~Serrano, Crits-Christoph, Matteson,
+  {\em et~al.\/}}]{pekar2022molecular}
+{Pekar, J.~E., A.~Magee, E.~Parker, N.~Moshiri, K.~Izhikevich, {\em et~al.\/}},
+  2022 The molecular epidemiology of multiple zoonotic origins of {SARS-CoV-2}.
+  Science {\bf 377}: 960--966,
+  \url{https://www.science.org/doi/full/10.1126/science.abp8337}.
+
+\bibitem[{Phan {\em et~al.\/}}(2020){Phan, Nguyen, Luong, Nguyen, Nguyen, Le,
+  Nguyen, Cao, {\rm and} Pham}]{phan2020importation}
+{Phan, L.~T., T.~V. Nguyen, Q.~C. Luong, T.~V. Nguyen, H.~T. Nguyen, {\em
+  et~al.\/}}, 2020 Importation and human-to-human transmission of a novel
+  coronavirus in vietnam. New England Journal of Medicine {\bf 382}: 872--874,
+  \url{https://www.nejm.org/doi/full/10.1056/nejmc2001272}.
+
+\bibitem[{Pipes {\em et~al.\/}}(2021){Pipes, Wang, Huelsenbeck, {\rm and}
+  Nielsen}]{pipes2021assessing}
+{Pipes, L., H.~Wang, J.~P. Huelsenbeck, {\rm and} R.~Nielsen}, 2021 Assessing
+  uncertainty in the rooting of the {SARS-CoV-2} phylogeny. Molecular Biology
+  and Evolution {\bf 38}: 1537--1543,
+  \url{https://academic.oup.com/mbe/article/38/4/1537/6028993}.
+
+\bibitem[{Plaue}(2020)]{theilsen}
+{Plaue, M.}, 2020 {A correlation measure based on Theil-Sen regression}.
+  Torwards Data Science,
+  \url{https://towardsdatascience.com/a-correlation-measure-based-on-theil-sen-regression-31b8b9ed64f1}.
+
+\bibitem[{ProMED}(2019)]{ProMED2109}
+{ProMED}, 2019 Undiagnosed pneumonia---{China (Hubei)}: Request for
+  information. \url{https://promedmail.org/promed-post/?id=6864153}.
+
+\bibitem[{van Dorp {\em et~al.\/}}(2020){van Dorp, Acman, Richard, Shaw, Ford,
+  Ormond, Owen, Pang, Tan, Boshier, {\em et~al.\/}}]{van2020emergence}
+{van Dorp, L., M.~Acman, D.~Richard, L.~P. Shaw, C.~E. Ford, {\em et~al.\/}},
+  2020 Emergence of genomic diversity and recurrent mutations in {SARS-CoV-2}.
+  Infection, Genetics and Evolution {\bf 83}: 104351,
+  \url{https://www.sciencedirect.com/science/article/pii/S1567134820301829}.
+
+\bibitem[{VanderPlas {\em et~al.\/}}(2018){VanderPlas, Granger, Heer, Moritz,
+  Wongsuphasawat, Satyanarayan, Lees, Timofeev, Welsh, {\rm and}
+  Sievert}]{vanderplas2018altair}
+{VanderPlas, J., B.~Granger, J.~Heer, D.~Moritz, K.~Wongsuphasawat, {\em
+  et~al.\/}}, 2018 Altair: interactive statistical visualizations for python.
+  Journal of open source software {\bf 3}: 1057,
+  \url{https://joss.theoj.org/papers/10.21105/joss.01057}.
+
+\bibitem[{{WHO}}(2020)]{who2020}
+{{WHO}}, 2020 {Pneumonia of unknown cause -- China}.
+  \url{https://www.who.int/emergencies/disease-outbreak-news/item/2020-DON229}.
+
+\bibitem[{Wood {\em et~al.\/}}(2019){Wood, Lu, {\rm and}
+  Langmead}]{wood2019improved}
+{Wood, D.~E., J.~Lu, {\rm and} B.~Langmead}, 2019 Improved metagenomic analysis
+  with {K}raken 2. Genome Biology {\bf 20}: 1--13.
+
+\bibitem[{Worobey {\em et~al.\/}}(2022){Worobey, Levy, Malpica~Serrano,
+  Crits-Christoph, Pekar, Goldstein, Rasmussen, Kraemer, Newman, Koopmans, {\em
+  et~al.\/}}]{worobey2022huanan}
+{Worobey, M., J.~I. Levy, L.~Malpica~Serrano, A.~Crits-Christoph, J.~E. Pekar,
+  {\em et~al.\/}}, 2022 {The Huanan Seafood Wholesale Market in Wuhan was the
+  early epicenter of the COVID-19 pandemic}. Science {\bf 377}: 951--959,
+  \url{https://www.science.org/doi/10.1126/science.abp8715}.
+
+\bibitem[{Wu}(2023)]{wu2023atlantic}
+{Wu, K.~J.}, 2023 The strongest evidence that an animal started the pandemic.
+  The Atlantic
+  \url{https://www.theatlantic.com/science/archive/2023/03/covid-origins-research-raccoon-dogs-wuhan-market-lab-leak/673390/}.
+
+\bibitem[{{Wuhan Municipal Health Commission}}(2019)]{healthcommission2019}
+{{Wuhan Municipal Health Commission}}, 2019 {Wuhan Municipal Health
+  Commission's briefing on the current situation of pneumonia in our city}. The
+  original report has been deleted, but archived versions are available on the
+  Wayback Machine at
+  \url{https://web.archive.org/web/20200131202951/http:/wjw.wuhan.gov.cn/front/web/showDetail/2019123108989}.
+
+\bibitem[{{Wuhan Municipal Health Commission}}(2020)]{healthcommission2020}
+{{Wuhan Municipal Health Commission}}, 2020 {Wuhan Municipal Health
+  Commission's report on viral pneumonia of unknown cause}. The original report
+  has been deleted, but archived versions are available on the Wayback Machine
+  at
+  \url{https://web.archive.org/web/20200111023106/http://wjw.wuhan.gov.cn/front/web/showDetail/2020011109035}.
+
+\bibitem[{Xiao {\em et~al.\/}}(2021){Xiao, Newman, Buesching, Macdonald, {\rm
+  and} Zhou}]{xiao2021animal}
+{Xiao, X., C.~Newman, C.~D. Buesching, D.~W. Macdonald, {\rm and} Z.-M. Zhou},
+  2021 {Animal sales from Wuhan wet markets immediately prior to the COVID-19
+  pandemic}. Scientific Reports {\bf 11}: 1--7,
+  \url{https://www.nature.com/articles/s41598-021-91470-2}.
+
+\bibitem[{Zhang {\em et~al.\/}}(2020){Zhang, Tan, Ling, Lu, Liu, Yi, Jia, Wu,
+  Shi, Xu, {\em et~al.\/}}]{zhang2020viral}
+{Zhang, X., Y.~Tan, Y.~Ling, G.~Lu, F.~Liu, {\em et~al.\/}}, 2020 Viral and
+  host factors related to the clinical outcome of {COVID-19}. Nature {\bf 583}:
+  437--440, \url{https://www.nature.com/articles/s41586-020-2355-0}.
+
+\end{thebibliography}
+
 }

### 2023-11-03 paper/final_submission/paper_include_graphics_commented_out_table_removed.tex
updates to adhere to formatting for final _Virus Evolution_ submission formatting

--- 2023-07-13 paper/final_submission/paper.tex
@@
 \begin{figure*}
-\includegraphics[width=\linewidth]{fig1.pdf}
+%\includegraphics[width=\linewidth]{fig1.pdf}
 \caption{
@@
 \begin{figure*}
-\centerline{\includegraphics[width=0.85\linewidth]{fig2.pdf}}
+%\centerline{\includegraphics[width=0.85\linewidth]{fig2.pdf}}
 \caption{
@@
 \label{tab:susceptible}
-}
-{\scriptsize
-\setlength{\tabcolsep}{5pt}
-\renewcommand{\arraystretch}{1.3}
-\begin{tabular}{llrrr}
-\toprule
-species      &   sample              & \makecell[r]{chordate \\ mitochondrial \\ reads from \\ species} &  \makecell[r]{reads \\ aligning \\ to SARS2} & \makecell[r]{ total pre-\\processed \\ reads} \\
-\midrule
-\multirow{14}{*}{raccoon dog} & HJ200048-20200112-1 &                                80\% &                        0 &                   $1.2\times 10^8$ \\
-                 & HJ200050-20200112-1 &                                69\% &                        0 &                     $1.0 \times 10^8$ \\
-                 & HJ200017-20200112-1 &                                61\% &                        0 &                   $1.1 \times 10^8$ \\
-                 & HJ200023-20200112-1 &                                58\% &                        0 &                   $6.9 \times 10^7$ \\
-                 & HJ200011-20200112-1 &                                41\% &                        0 &                   $5.8 \times 10^7$ \\
-                 & HJ200012-20200112-1 &                                39\% &                        0 &                   $1.3 \times 10^8$ \\
-                 & {\bf Q61} &                                {\bf 32\%} &                        {\bf 1} &                   $\mathbf{2.1 \times 10^8}$ \\
-                 & HJ200019-20200112-1 &                                30\% &                        0 &                     $7.0 \times 10^7$ \\
-                 & HJ200006-20200112-1 &                                29\% &                        0 &                   $1.3 \times 10^8$ \\
-                 & HJ200001-20200112-1 &                                28\% &                        0 &                   $1.2 \times 10^8$ \\
-                 & HJ200018-20200112-1 &                                26\% &                        0 &                   $1.4 \times 10^8$ \\
-                 & HJ200044-20200112-1 &                                25\% &                        0 &                   $1.2 \times 10^8$ \\
-                 & HJ200047-20200112-1 &                                22\% &                        0 &                   $1.4 \times 10^8$ \\
-                 & 629-3-C &                                22\% &                        0 &                   $2.5 \times 10^8$ \\
-\cline{1-5}
-\multirow{6}{*}{hoary bamboo rat} & HJ200065-20200112-1 &                                48\% &                        0 &                   $7.3 \times 10^7$ \\
-                 & HJ200062-20200112-1 &                                40\% &                        0 &                   $1.8 \times 10^8$ \\
-                 & 629-5-L4 &                                35\% &                        0 &                   $1.4 \times 10^8$ \\
-                 & 629-13-L &                                33\% &                        0 &                   $1.5 \times 10^8$ \\
-                 & 629-1-L1 &                                30\% &                        0 &                   $2.5 \times 10^8$ \\
-                 & HJ200049-20200112-1 &                                23\% &                        0 &                     $1.0 \times 10^8$ \\
-\cline{1-5}
-\multirow{6}{*}{Amur hedgehog} & W-8-25-L2 &                                56\% &                        0 &                   $3.1 \times 10^8$ \\
-                 & HJ200040-20200112-1 &                                51\% &                        0 &                   $1.5 \times 10^8$ \\
-                 & HJ200039-20200112-1 &                                30\% &                        0 &                   $1.2 \times 10^8$ \\
-                 & {\bf 8-25-M1} &                                {\bf 30\%} &                      {\bf 24} &                   $\mathbf{4.4 \times 10^8}$ \\
-                 & HJ200038-20200112-1 &                                23\% &                        0 &                     $1.0 \times 10^8$ \\
-                 & W-8-25-D2 &                                22\% &                        0 &                   $3.3 \times 10^8$ \\
-\cline{1-5}
-Malayan porcupine & {\bf Q70} &                                {\bf 85\%} &                        {\bf 2} &                   $\mathbf{1.5 \times 10^8}$ \\
-\cline{1-5}
-Himalayan marmot & HJ200005-20200112-1 &                                30\% &                        0 &                   $1.2 \times 10^8$ \\
-\bottomrule
-\end{tabular}
 }
@@
 \begin{figure*}
-\includegraphics[width=\linewidth]{fig3.png}
+%\includegraphics[width=\linewidth]{fig3.png}
 \caption{
@@
 \begin{figure}
-\includegraphics[width=\linewidth]{fig4.pdf}
+%\includegraphics[width=\linewidth]{fig4.pdf}
 \caption{
@@
 \label{tab:consistency}
-}
-{\footnotesize
-\setlength{\tabcolsep}{5pt}
-\renewcommand{\arraystretch}{1.3}
-\begin{tabular}{llll}
-\toprule
-sample & \makecell[l]{RT-qPCR test \\ result (Ct)} & \makecell[l]{sequencing reads \\ mapping to SARS-CoV-2 \\ out of high-quality reads} & \makecell[l]{classification in \\ Chinese CDC study} \\
-\midrule
-F100 & positive (34.7) & 7,200 out of $2.6 \times 10^8$ & positive \\
-Q61 & negative & 1 out of $2.1 \times 10^8$ & positive \\
-E-10-29-2 & negative & 0 out of $1.9 \times 10^8$ & negative \\
-A1 & negative & not sequenced & negative \\
-\bottomrule
-\end{tabular}
 }

### 2023-11-03 paper/final_submission/table_2.tex
updates to adhere to formatting for final _Virus Evolution_ submission formatting

--- 2023-07-13 paper/final_submission/paper.tex
@@
 
-\title{Association between SARS-CoV-2 and metagenomic content of samples from the Huanan Seafood Market}
-
-\author[]{\Large Jesse D. Bloom}
-
-\affil[]{Fred Hutchinson Cancer Center; Howard Hughes Medical Institute
-
-}
-
-\keywords{}
-
-\runningtitle{} % For use in the footer
-\runningauthor{}
-
-\begin{abstract}
-The role of the Huanan Seafood Market in the early SARS-CoV-2 outbreak remains unclear.
-Recently the Chinese CDC released data from deep sequencing of environmental samples collected from the market after it was closed on January-1-2020~\citep{liu2023surveillance}.
-Prior to this release, \citet{crits2023genetic} analyzed data from a subset of the samples.
-Both studies concurred that the samples contained genetic material from a variety of species, including some like raccoon dogs that are susceptible to SARS-CoV-2.
-However, neither study systematically analyzed the relationship between the amount of genetic material from SARS-CoV-2 and different animal species.
-Here I implement a fully reproducible computational pipeline that jointly analyzes the number of reads mapping to SARS-CoV-2 and the mitochondrial genomes of chordate species across the full set of samples.
-I validate the presence of genetic material from numerous species, and calculate mammalian mitochondrial compositions similar to those reported by \citet{crits2023genetic}.
-However, the SARS-CoV-2 content of the environmental samples is generally very low: only 21 of 176 samples contain more than 10 SARS-CoV-2 reads, despite most samples being sequenced to depths exceeding $10^8$  total reads.
-None of the samples with double-digit numbers of SARS-CoV-2 reads have a substantial fraction of their mitochondrial material from any non-human susceptible species.
-Only one of the 14 samples with at least a fifth of the chordate mitochondrial material from raccoon dogs contains any SARS-CoV-2 reads, and that sample only has 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
-Instead, SARS-CoV-2 reads are most correlated with reads mapping to various fish, such as catfish and largemouth bass.
-These results suggest that while metagenomic analysis of the environmental samples is useful for identifying animals or animal products sold at the market, co-mingling of animal and viral genetic material is unlikely to reliably indicate whether any animals were infected by SARS-CoV-2.
-\end{abstract}
-
 \begin{document}
 
-\maketitle
-\thispagestyle{firststyle}
-%\marginmark
-\firstpagefootnote
-
-\correspondingauthoraffiliation{}{\href{mailto:jbloom@fredhutch.org}{jbloom@fredhutch.org}}
-\vspace{-33pt}% Only used for adjusting extra space in the left column of the first page
-
-\lettrine[lines=2]{\color{color2}I}{}nitial reports from Chinese officials about the outbreak that eventually became the SARS-CoV-2 pandemic described patients associated with the Huanan Seafood Market, and said there was no evidence of significant human-to-human transmission~\citep{ProMED2109, healthcommission2019, healthcommission2020, who2020}.
-These two claims implied that the infections likely originated from a non-human source within the market.
-
-But by mid to late January of 2020, it was clear SARS-CoV-2 was spreading from human to human, and had been for some time~\citep{chan2020familial,phan2020importation,li2020early,nishiura2020initial}.
-In addition, Chinese scientists published papers reporting that some of the earliest identified human cases in December 2019 had no link to the Huanan Seafood Market~\citep{huang2020clinical,chen2020epidemiological,zhang2020viral}.
-Thus began a debate that continues to this day: was the market an initial source of zoonotic infections from animals, or was it simply a superspreading site that amplified earlier human infections from another source~\citep{cohen2020wuhan}?
-
-Some papers have argued that the fact that many of the known human cases from December 2019 are centered around market indicates it was the original source of the virus~\citep{worobey2022huanan}, and that an early two-mutation polytomy in the SARS-CoV-2 phylogenetic tree suggests multiple zoonotic jumps there~\citep{pekar2022molecular}.
-Other papers have argued that reports of earlier cases unlinked to the market~\citep{huang2020clinical,chen2020epidemiological,zhang2020viral,ma2020SCMP} and uncertainties about the completeness of case data provided by the Chinese government~\citep{cohen2022anywhere,hvistendahl2023} mean that the first human infections could have occurred upstream of the December 2019 cases at the market~\citep{pekar2021timing,pipes2021assessing,bloom2021recovery}.
-
-Despite these disputes, it is universally agreed that there were human cases of SARS-CoV-2 at the Huanan Seafood Market by mid December of 2019~\citep{li2020early}, and that some animals susceptible to SARS-CoV-2 were sold at the market~\citep{xiao2021animal}.
-The key unanswered question is if any animals were infected, and if so whether they infected humans or were infected by them.
-
-In early 2022, the Chinese CDC posted a preprint on \textit{Research Square} describing their sampling of the market beginning immediately after its closure on January-1-2020~\citep{liu2022surveillance}.
-The Chinese CDC collected samples from both the environment and animals / animal products.
-They reported that none of the 457 animal samples tested positive for SARS-CoV-2, but that 73 of 923 environmental samples tested positive.
-A significant caveat of the Chinese CDC study is that all the samples were collected on January-1-2020 or later, which is probably \emph{at least} a month after the first human infections in Wuhan~\citep{zhang2020viral,van2020emergence,he2020evaluating,pipes2021assessing,pekar2021timing,odni2022,pekar2022molecular}.
-Despite this significant caveat, the data reported in the Chinese CDC study has been variously interpreted to support arguments that virus in the market was human derived~\citep{liu2022surveillance}, that the outbreak originated from live animals sold in the market~\citep{worobey2022huanan}, or that the virus spread among humans in Mahjong rooms or toilets in the market~\citep{courtier2022sars}.
-
-One aspect of the data gathered by the Chinese CDC that was \emph{not} thoroughly analyzed in their 2022 preprint was the metagenomic content of the environmental samples. 
-The 2022 Chinese CDC preprint simply reported that the number of SARS-CoV-2 reads in deep sequencing of the samples was correlated with the number of human reads (fourth figure of \citet{liu2022surveillance}), but did not specify correlations for other species and did not provide the raw sequencing data.
-Other scientists pointed out that if the raw data were shared, it would be possible to expand upon the analysis in the 2022 Chinese CDC preprint to determine what other animal species contributed genetic material to environmental samples containing SARS-CoV-2~\citep{cohen2022anywhere,cohen2022studies}.
-
-At some point after posting their 2022 preprint, the Chinese CDC uploaded raw sequencing data for some environmental samples to the GISAID database, where they were subsequently downloaded and analyzed by another group of scientists.
-News of this analysis leaked to the media, which published stories emphasizing the co-mingling of genetic material from raccoon dogs and SARS-CoV-2 in one of the samples~\citep{wu2023atlantic,mueller2023nytimes}.
-The next week, the scientists published their report~\citep{crits2023genetic}, which described bioinformatic analyses showing that some environmental samples contained genetic material from animals susceptible to SARS-CoV-2, including raccoon dogs.
-However, although \citet{crits2023genetic} analyzed the mammalian metagenomic content of the partial set of samples they obtained from GISAID, they did not report any analysis of the SARS-CoV-2 content of the samples.
-
-A week later, the Chinese CDC posted an updated version of their preprint on the \textit{ChinaXiv} server~\citep{liu2023surveillance-chinaxiv} and released the full set of raw sequencing data in public databases.
-This updated preprint, and a version published by \textit{Nature} the next week~\citep{liu2023surveillance} emphasized different aspects of the data than \citet{crits2023genetic}.
-Specifically, although the Chinese CDC concurred that some of the environmental samples contained material from raccoon dogs and other susceptible species, they stated that material from many species was found, and that raccoon dog material was more commonly identified in SARS-CoV-2 negative samples.
-However, like \citet{crits2023genetic}, the new version of the Chinese CDC paper~\citep{liu2023surveillance} did not analyze the association between the amount of SARS-CoV-2 and animal material in the samples, and even removed the figure from the original preprint~\citep{liu2022surveillance} that showed that the number of SARS-CoV-2 reads was correlated with reads from both humans and several unidentified species.
-
-Here I systematically analyze the relationship between the metagenomic and SARS-CoV-2 content of all environmental samples for which deep sequencing data are available.
-This analysis confirms that samples contain genetic material from a wide range of species, including some that are susceptible to SARS-CoV-2.
-However, the samples that contain abundant material from raccoon dogs and other non-human susceptible species usually contain little or no SARS-CoV-2 reads.
-Furthermore, the number of SARS-CoV-2 reads is not consistently correlated with reads mapping to non-human susceptible species, but is instead generally most correlated with reads from species that are implausible candidates for having been infected with SARS-CoV-2.
-These results suggest that SARS-CoV-2 was widespread at the market by January 2020, and therefore that co-mingling of viral and animal genetic material in environmental samples collected at that time is unlikely to be informative about the original source of the outbreak.
-
-\section{Results}
-
-\subsection{Deep sequencing data deposited in public databases on March-29-2023 by the Chinese CDC are a superset of the data analyzed by Crits-Christoph et al earlier that month}
-On March-29-2023, the Chinese CDC released raw deep sequencing data for samples taken from the Huanan Market in early 2020.
-These data were released on the NGDC database under accession CRA010170 (\url{https://ngdc.cncb.ac.cn/gsa/browse/CRA010170}).
-
-I created a fully reproducible computational pipeline that downloaded and analyzed the full dataset from the NGDC (my pipeline is available at \url{https://github.com/jbloom/Huanan_market_samples}).
-The dataset consists of 696 FASTQ files from 395 deep sequencing runs of 176 samples, with the total size of the gzipped files exceeding three terabytes.
-
-Prior to the full posting of the dataset on the NGDC by the Chinese CDC, \citet{crits2023genetic} downloaded and analyzed 227 FASTQ files that had been uploaded to the GISAID database by the Chinese CDC.
-When news of the analysis by \citet{crits2023genetic} became public, GISAID reportedly revoked access to the data and stated that the set of files that had been downloaded was incomplete~\citep{cohen2023new,gisaid_statement}.
-
-To determine the relationship between the 696 FASTQ files released by the Chinese CDC on the NGDC on March-29-2023 and the 227 files downloaded earlier that month by \citet{crits2023genetic}, I computed the SHA-512 hashes for all the NGDC FASTQ files and compared them to the hashes reported in Table S1 of \citet{crits2023genetic}.
-This analysis confirmed that the dataset released on the NGDC contained unmodified versions of all 227 FASTQ files downloaded by \citet{crits2023genetic}, plus an additional 469 FASTQ files (Tables S1 and S2).
-
-\subsection{Analysis of mitochondrial sequences validates mammalian metagenomics by Crits-Christoph et al, and also finds abundant non-mammalian chordates}
-To determine the metagenomic content of the samples, I performed an analysis that conceptually parallels that described by \citet{crits2023genetic}.
-Throughout this paper, I report results only for sequencing described by the Chinese CDC as ``RNA sequencing of total nucleic acids from environmental swabs for metagenomics,'' and exclude the small number of samples that involve amplicon-based sequencing of SARS-CoV-2 or RNA sequencing for viral whole-genome assembly (Table S2).
-
-Briefly, the deep sequencing data were pre-processed to retain only high-quality reads.
-These reads were aligned to the SARS-CoV-2 genome and a representative set of chordate mitochondrial genomes.
-%The set of mitochondrial genomes was assembled by downloading all NCBI RefSeq genomes, filtering for those from chordates, and removing genomes that were highly similar to other retained genomes (within a Mash distance of $<$0.07).
-This set included all the mammalian mitochondrial genomes from all species for which alignments were reported in \citet{crits2023genetic}.
-The accessions for the 4,170 chordate mitochondrial genomes that formed this set are listed in Table S3.
-The counts of reads mapping uniquely to each mitochondrial genome for each sequencing run and sample are provided in Tables S4 and S5, respectively.
-
-I compared the mitochondrial compositions of the samples determined by my analysis to those reported by \citet{crits2023genetic}.
-Note that while \citet{crits2023genetic} describe aligning the sequencing data to all metazoa mitochondria, they only report alignment counts for mammals.
-The aligned read counts for mammals from my analysis are highly correlated with those reported by \citet{crits2023genetic} for all sequencing runs with a reasonable number of total counts (Figure S1).
-My analysis and that of \citet{crits2023genetic} are independently implemented, and use different alignment programs, reference genome sets, and quality filtering criteria---so the fact that the two analyses give highly correlated results is reassuring.
-
-Despite the strong correlation between my analysis and that by \citet{crits2023genetic} on a common set of mammalian mitochondrial genomes, it is important to emphasize that the results are dependent on the genomes to which the reads are aligned.
-For instance, Figure~\ref{fig:composition} shows the composition of sample Q61 as determined by aligning reads to different genome sets.
-My analysis recapitulates the finding of \citet{crits2023genetic} that raccoon dogs make the largest contribution to the mammalian mitochondrial genetic material in this sample---but if the analysis is expanded to all chordates, then duck mitochondrial genetic material is more abundant (Figure~\ref{fig:composition}A).
-But raccoon dog is more abundant than duck if the analysis is instead performed by assembling contigs and then aligning to the full genomes of raccoon dogs, ducks, and several other species, in line with the findings of \citet{crits2023genetic}.
-These results are a reminder that the measured metagenomic composition is contingent upon the method and reference set used.
-Throughout the rest of this paper, I use compositions calculated from reads mapping to mitochondrial genomes rather than contigs mapped to the full genomes for two reasons: (1) this parallels the approach of \citet{crits2023genetic} who only reported mitochondrial composition for samples other than Q61, and (2) a number of relevant species only have mitochondrial but not full genomes available in NCBI.
-
-The interactive figure at \url{https://jbloom.github.io/Huanan_market_samples/mito_composition.html} enables the reader to examine the mitochondrial composition of each sample over both mammals and chordates.
-
-\begin{figure*}
-\includegraphics[width=\linewidth]{fig1.pdf}
-\caption{
-Metagenomic composition of sample Q61.
-{\bf (A)}
-Composition as determined by aligning reads to mitochondrial genomes.
-From left to right: composition determined in current study across all chordates, composition determined in current study across all mammals, composition reported in \citet{crits2023genetic} for just mammals shown in the first figure of their report, and composition reported \citet{crits2023genetic} across all mammals in the third supplementary table of their report.
-{\bf (B)}
-Composition determined in the current study by aligning assembled contigs to the four indicated genomes; this composition is similar to that reported in the third figure of \citet{crits2023genetic}.
-See \url{https://jbloom.github.io/Huanan_market_samples/mito_composition.html} for an interactive version of (A) that allows similar pie charts to be viewed for any sample.
-See \url{https://jbloom.github.io/Huanan_market_samples/genomic_contig_composition.html} for an interactive version of (B).
-\label{fig:composition}
-}
-\end{figure*}
-
-\subsection{Most environmental samples contain little or no SARS-CoV-2 reads}
-To quantify the SARS-CoV-2 content of the samples, I plotted the percentage of all high-quality reads that aligned to SARS-CoV-2 (Figure~\ref{fig:sars2_aligned}, Table S6, Table S7).
-Over three-quarters of the samples had no reads aligning to SARS-CoV-2, and most of the rest only had a small number of viral reads.
-
-Despite the fact that most samples were sequenced to depths that exceeded 100,000,000 reads, only two samples had $>$1,000 reads mapping to SARS-CoV-2 (F100 and B5).
-These samples were both collected on January-1-2020 and had a chordate mitochondrial composition dominated by catfish and largemouth bass (see interactive version of Figure~\ref{fig:sars2_aligned} at \url{https://jbloom.github.io/Huanan_market_samples/sars2_aligned_vertical.html}).
-
-\begin{figure*}
-\centerline{\includegraphics[width=0.85\linewidth]{fig2.pdf}}
-\caption{
-Percentage of high-quality reads that align to SARS-CoV-2 for each sample.
-Points are colored according to the date that the sample was collected.
-Note the x-axis uses a symlog scale.
-See \url{https://jbloom.github.io/Huanan_market_samples/sars2_aligned_vertical.html} for an interactive version of this plot where you can mouseover points for details including the mitochondrial composition of each sample, and select only samples from specific dates or from locations.
-\label{fig:sars2_aligned}
-}
-\end{figure*}
-
-Most samples with the highest SARS-CoV-2 content were collected on the Chinese CDC's first sampling date of January-1-2020, but some samples with SARS-CoV-2 reads were collected on later dates (Figure~\ref{fig:sars2_aligned}).
-The non-January-1 sample with the most SARS-CoV-2 was RLC-3, which was collected on February-2-2022 and had chordate mitochondrial composition dominated by rat snake, spotted dove, and human (interactive version of Figure~\ref{fig:sars2_aligned}).
-
-\begin{table}
-\caption{
-Reads mapping to SARS-CoV-2 out of all high-quality (pre-processed) reads for samples with $\ge$20\% of their chordate mitochondrial composition from a susceptible non-human species as defined in \citet{crits2023genetic}.
-Samples with non-zero SARS-CoV-2 reads are in bold.
-See Table S8 for a similar table that shows mammalian rather than chordate composition.
-This table uses a 20\% cutoff due to space considerations; to see similar data tabulated for all samples with no cutoff, see the much larger Table S9 (for raccoon dog) and Table S5 (for all species).
-\label{tab:susceptible}
-}
-{\scriptsize
-\setlength{\tabcolsep}{5pt}
-\renewcommand{\arraystretch}{1.3}
-\begin{tabular}{llrrr}
-\toprule
-species      &   sample              & \makecell[r]{chordate \\ mitochondrial \\ reads from \\ species} &  \makecell[r]{reads \\ aligning \\ to SARS2} & \makecell[r]{ total pre-\\processed \\ reads} \\
-\midrule
-\multirow{14}{*}{raccoon dog} & HJ200048-20200112-1 &                                80\% &                        0 &                   $1.2\times 10^8$ \\
-                 & HJ200050-20200112-1 &                                69\% &                        0 &                     $1.0 \times 10^8$ \\
-                 & HJ200017-20200112-1 &                                61\% &                        0 &                   $1.1 \times 10^8$ \\
-                 & HJ200023-20200112-1 &                                58\% &                        0 &                   $6.9 \times 10^7$ \\
-                 & HJ200011-20200112-1 &                                41\% &                        0 &                   $5.8 \times 10^7$ \\
-                 & HJ200012-20200112-1 &                                39\% &                        0 &                   $1.3 \times 10^8$ \\
-                 & {\bf Q61} &                                {\bf 32\%} &                        {\bf 1} &                   $\mathbf{2.1 \times 10^8}$ \\
-                 & HJ200019-20200112-1 &                                30\% &                        0 &                     $7.0 \times 10^7$ \\
-                 & HJ200006-20200112-1 &                                29\% &                        0 &                   $1.3 \times 10^8$ \\
-                 & HJ200001-20200112-1 &                                28\% &                        0 &                   $1.2 \times 10^8$ \\
-                 & HJ200018-20200112-1 &                                26\% &                        0 &                   $1.4 \times 10^8$ \\
-                 & HJ200044-20200112-1 &                                25\% &                        0 &                   $1.2 \times 10^8$ \\
-                 & HJ200047-20200112-1 &                                22\% &                        0 &                   $1.4 \times 10^8$ \\
-                 & 629-3-C &                                22\% &                        0 &                   $2.5 \times 10^8$ \\
-\cline{1-5}
-\multirow{6}{*}{hoary bamboo rat} & HJ200065-20200112-1 &                                48\% &                        0 &                   $7.3 \times 10^7$ \\
-                 & HJ200062-20200112-1 &                                40\% &                        0 &                   $1.8 \times 10^8$ \\
-                 & 629-5-L4 &                                35\% &                        0 &                   $1.4 \times 10^8$ \\
-                 & 629-13-L &                                33\% &                        0 &                   $1.5 \times 10^8$ \\
-                 & 629-1-L1 &                                30\% &                        0 &                   $2.5 \times 10^8$ \\
-                 & HJ200049-20200112-1 &                                23\% &                        0 &                     $1.0 \times 10^8$ \\
-\cline{1-5}
-\multirow{6}{*}{Amur hedgehog} & W-8-25-L2 &                                56\% &                        0 &                   $3.1 \times 10^8$ \\
-                 & HJ200040-20200112-1 &                                51\% &                        0 &                   $1.5 \times 10^8$ \\
-                 & HJ200039-20200112-1 &                                30\% &                        0 &                   $1.2 \times 10^8$ \\
-                 & {\bf 8-25-M1} &                                {\bf 30\%} &                      {\bf 24} &                   $\mathbf{4.4 \times 10^8}$ \\
-                 & HJ200038-20200112-1 &                                23\% &                        0 &                     $1.0 \times 10^8$ \\
-                 & W-8-25-D2 &                                22\% &                        0 &                   $3.3 \times 10^8$ \\
-\cline{1-5}
-Malayan porcupine & {\bf Q70} &                                {\bf 85\%} &                        {\bf 2} &                   $\mathbf{1.5 \times 10^8}$ \\
-\cline{1-5}
-Himalayan marmot & HJ200005-20200112-1 &                                30\% &                        0 &                   $1.2 \times 10^8$ \\
-\bottomrule
-\end{tabular}
-}
-\end{table}
-
-The Q61 sample that was the focus of the raccoon-dog centered media coverage~\citep{wu2023atlantic,mueller2023nytimes} preceding the report by \citet{crits2023genetic} is one of 70 samples collected on January-12-2020 from the west wing of the market.
-Six of these samples contained SARS-CoV-2 reads, while the other 64 had no SARS-CoV-2 reads (Figure~\ref{fig:sars2_aligned}).
-Sample Q61, which has a mitochondrial metagenomic composition dominated by duck and raccoon dog, had 1 of $\sim2.1 \times 10^8$ high-quality reads mapping to SARS-CoV-2 (interactive version of Figure~\ref{fig:sars2_aligned}).
-The samples from January-12-2020 with the most SARS-CoV-2 reads were Q68, Q37, and Q64; species that contributed $\ge$10\% of chordate mitochondrial reads in these samples were chicken, dog, duck, Chinese salamander, rabbit, and various snakes.
-
- Overall, there were SARS-CoV-2 reads in just 3 of the 28 samples with at least 20\% of their chordate mitochondrial composition from the non-human susceptible species thought to have been sold live in the market (Table~\ref{tab:susceptible}).
-There were 14 samples with at least 20\% chordate mitochondrial composition from raccoon dogs, but only sample Q61 contained any SARS-CoV-2 reads (1 of $\sim2.1 \times 10^8$ high-quality reads).
-None of the six samples with at least 20\% chordate mitochondrial composition from bamboo rats contained any SARS-CoV-2 reads.
-There was one sample each with at least 20\% chordate mitochondrial composition from Amur hedgehog and Malayan porcupine that contained SARS-CoV-2 reads (Table~\ref{tab:susceptible}).
-If mitochondrial composition is instead analyzed only among mammals rather than chordates, Q61 is still the only sample with at least 20\% raccoon dog composition that contains any SARS-CoV-2 reads (Table S8).
-There is one sample with at least 20\% of its mammalian (but not chordate) mitochondrial composition from bamboo rat that contains a small number of SARS-CoV-2 reads (Table S8).
-
-Note that the 20\% cutoff applied in Tables~\ref{tab:susceptible} and S8 is not integral to the analysis, but is just a way to subset on samples containing the most material from the species of interest to make the tables small enough to easily visualize.
-See Tables S9 and S5 for much larger tables that show comparable data for all samples with respect to the raccoon dog or overall mitochondrial composition, respectively.
-For instance, Table S9 shows that there are a few samples with lower raccoon dog mitochondrial composition that also contain some SARS-CoV-2 reads.
-However, ultimately these very large tables are difficult to visualize, and so the relationship between SARS-CoV-2 and mitochondrial content for each species for all samples is probably more facilely visualized using the interactive versions of the scatter plots described in the next subsection.
-
-\subsection{Correlations of abundance of SARS-CoV-2 to mitochondrial genetic material from various species}
-To more systematically examine the relationship between SARS-CoV-2 and genetic material from different chordates, I calculated the correlation between the number of reads mapping to SARS-CoV-2 versus the mitochondrial genome of each species across all samples.
-If the correlation is calculated on a log-log scale as was done in the original Chinese CDC preprint~\citep{liu2022surveillance}, then the five species whose genetic material is most correlated with the abundance of SARS-CoV-2 are (in order) largemouth bass, catfish, cow, carp, and snakehead fish (Figure~\ref{fig:per_species_corr_faceted}).
-None of these species are likely hosts for SARS-CoV-2: non-mammals are not thought to be infectable, and cows were probably sold as animal products rather than live animals.
-There is a modest correlation between the abundance of SARS-CoV-2 and human mitochondrial material, but this correlation is weaker than for several other species (Figure~\ref{fig:per_species_corr_faceted}).
-There is a negative correlation between the abundance of SARS-CoV-2 and mitochondrial material from raccoon dogs and hoary bamboo rats, and at most a weak positive correlation for other susceptible species thought to have been sold live at the market (Figure~\ref{fig:per_species_corr_faceted}).
-
-Furthermore, the correlations are highly contingent on the sample set and details of how the statistics are calculated.
-For instance, if we exclude the first sampling timepoint and only look at samples calculated on January-12-2020 or later, then the most correlated species is the Oriental rat snake, with the correlation for humans becoming even weaker and that for raccoon dogs remaining negative (Figure S2).
-The correlations also change if only consider samples containing a non-zero number of SARS-CoV-2 reads, or calculate correlations on a linear rather than log scale (Figure S3).
-The interactive versions of the correlation plots at \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html} and \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_single.html} enable the reader to explore the correlations for different sample subsets and methods of calculating the statistics.
-%Exploration of these interactive plots shows that it takes a great deal of cherry picking of parameter combinations to put any plausibly infectable hosts at the top of the list of species whose mitochondrial material is most correlated with the abundance of SARS-CoV-2.
-
-\begin{figure*}
-\includegraphics[width=\linewidth]{fig3.png}
-\caption{
-Correlation between percent of all reads mapping to SARS-CoV-2 and the mitochondrial genome of each of the indicated species.
-Each point represents a different environmental sample, and the orange text shows the Pearson correlation.
-The scales are log10, and values of zero (which cannot be plotted on a log scale) are shown as half the minimum non-zero value observed across all samples.
-See \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_faceted.html} for an interactive version of this plot that enables mouseover of points for sample details, selection only of samples collected on specific dates or containing at least one SARS-CoV-2 reads, adjustment of scales from log to linear, and adjustment of mitochondrial percent to be of reads mapping to any mitochondria rather than of all reads.
-See \url{https://jbloom.github.io/Huanan_market_samples/per_species_corr_single.html} for similar plots for individual species.
-The plots shown here include only samples with at least 200 aligned mitochondrial reads; that option can be adjusted in the interactive plots.
-\label{fig:per_species_corr_faceted}
-}
-\end{figure*}
-
-\subsection{Fully annotated correlation figure analogous to that in the original Chinese CDC preprint}
-The original Chinese CDC preprint from 2022 had in its fourth figure a plot showing the correlation between the amount of SARS-CoV-2 and genetic material from various animal species in the environmental samples~\citep{liu2022surveillance}.
-The preprint annotated one point in the plot to indicate there was a correlation between the amount of SARS-CoV-2 and human genetic material.
-However, the preprint failed to annotate the identity of any other species with genetic material that correlated with SARS-CoV-2 abundance.
-This omission of full species annotations in the correlation plot was widely noted, including in two news articles in \textit{Science} in 2022~\citep{cohen2022anywhere,cohen2022studies}, and a third article in 2023 that re-printed the incompletely annotated plot~\citep{cohen2023unearthed}.
-However, neither of the two more recent studies of the metagenomic data have addressed this omission: \citet{crits2023genetic} did not report any analysis of the abundance of SARS-CoV-2 reads, and the published 2023 version of the Chinese CDC preprint simply dropped the plot altogether~\citep{liu2023surveillance}.
-
-\begin{figure}
-\includegraphics[width=\linewidth]{fig4.pdf}
-\caption{
-Correlations between SARS-CoV-2 content and mitochondrial content for all species.
-Top row show just samples containing at least one SARS-CoV-2 read and bottom row shows samples regardless of SARS-CoV-2 content; left shows samples from all sampling dates and right shows just samples from the January-12-2020 date when most of the wildlife sampling occurred.
-This plot is designed to mimic the fourth figure of \citet{liu2022surveillance}.
-Some key species are labeled; see the interactive version at \url{https://jbloom.github.io/Huanan_market_samples/overall_corr.html} to mouseover all points for details, select different subsets of samples, and calculate the correlations on a linear or log scale.
-See Figure S4 for a version of this plot that uses correlations based on a Theil-Sen estimator that is more robust to outliers.
-\label{fig:overall_corr}
-}
-\end{figure}
-
-To remedy this omission, I used the chordate mitochondrial compositions calculated in the current study to generate plots analogous to that in the original Chinese CDC preprint.
-When the correlations are taken across all sampling dates only for samples containing at least one SARS-CoV-2 read (as done in the original Chinese CDC preprint), then the genetic material of several non-human species (largemouth bass, catfish, cow, sheep, pig) is more correlated with SARS-CoV-2 reads than is human material (Figure~\ref{fig:overall_corr}, upper left).
-If the plot is expanded to all samples (regardless of whether or not they contain SARS-CoV-2 reads), the trends are broadly similar, with fish remaining the most correlated species (Figure~\ref{fig:overall_corr}, lower left).
-It is also informative to take the correlations only across samples collected on January-12-2020, which was the date of most intensive sampling of the wildlife stalls (this is when Q61 was collected).
- For January-12-2020 samples that contain SARS-CoV-2 reads, humans are among the species whose genetic material is most correlated with SARS-CoV-2 reads, but goat and spotted dove have roughly equivalent correlations (Figure~\ref{fig:overall_corr}, upper right).
- If we include January-12-2020 samples regardless of SARS-CoV-2 content, then snakehead fish and Malayan porcupine are among the most correlated species.
-Raccoon dog and bamboo rat genetic material are not positively correlated with SARS-CoV-2 reads in any of the sample sets.
-See the interactive version of Figure~\ref{fig:overall_corr} at \url{https://jbloom.github.io/Huanan_market_samples/overall_corr.html} to explore other sample subsets.
-The trends are broadly similar if the correlations are instead calculated based on a Theil-Sen estimator~\citep{theilsen} that is more robust to outliers (Figure S4).
-
-Note that the correlation plots in Figure~\ref{fig:overall_corr} are not exactly identical to those in the original Chinese CDC preprint.
-Since \citet{liu2022surveillance} do not provide sufficiently detailed methods to fully reproduce their analysis, it is impossible to definitively determine the source of the discrepancy---but it is likely due to the inclusion of different genomes in the reference sets used to calculate the metagenomic composition of the samples.
-
-\subsection{The low SARS-CoV-2 read counts raise questions about the consistency of the approach use to call sample positivity}
-The Chinese CDC study includes a table that classifies which environmental samples were positive for SARS-CoV-2 (this is first main table in \citet{liu2022surveillance} and the second supplementary table in \citet{liu2023surveillance}).
-Subsequent studies that have re-analyzed the Chinese CDC data have re-used their classifications of sample positivity~\citep{worobey2022huanan,courtier2022sars,crits2023genetic}.
-
-\begin{table}
-\caption{
-Inconsistency in criteria used to classify SARS-CoV-2 positivity in Chinese CDC study, illustrated with four example samples.
-There is no consistent rationale for classifying Q61 as positive but E-10-29-2 and A1 as negative: all three were reported negative by RT-qPCR, and A1 was not analyzed by sequencing while the difference in the number of SARS-CoV-2 reads between Q61 and E-10-29-2 is not statistically significant ($P = 1$ for both Fisher exact or $\chi^2$ test).
-RT-qPCR results are from the second supplementary table of \citet{liu2023surveillance} (or equivalently the first table of \citet{liu2022surveillance}).
-\label{tab:consistency}
-}
-{\footnotesize
-\setlength{\tabcolsep}{5pt}
-\renewcommand{\arraystretch}{1.3}
 \begin{tabular}{llll}
@@
 \end{tabular}
-}
-\end{table}
-
-Now that the full data are available, we can examine the criteria used to classify samples as positive or negative for SARS-CoV-2.
-It appears that the Chinese CDC classified samples as positive if they met \emph{either} of two criteria: they tested positive for SARS-CoV-2 by RT-qPCR, \emph{or} they were metagenomically sequenced and contained at least one read mapping to SARS-CoV-2 (Table S10 gives sequencing and RT-qPCR data for all samples that were either sequenced or called positive by the Chinese CDC).
-However, these criteria are not consistent because not all samples were analyzed by both methods, and the differences in the number of SARS-CoV-2 reads in the sequencing data are often not statistically significant between samples classified as positive versus negative.
-This inconsistency is illustrated in Table~\ref{tab:consistency} for four example samples: F100 is clearly positive (it both tested positive by RT-qPCR and contained thousands of reads mapping to SARS-CoV-2), but there is no statistical rationale for classifying Q61 as positive but E-10-29-2 and A1 as negative.
-
-To see the lack of statistical rationale for classifying Q61 as positive but E-10-29-2 as negative (Table~\ref{tab:consistency}), we can use a Fisher exact test (or alternatively a $\chi^2$ test) to test the null hypothesis that the reads in samples Q61 and E-10-29-2 are drawn from the same underlying distribution of viral and non-viral reads given that we observe 1 in $2.1 \times 10^8$ SARS-CoV-2 reads in Q61 and 0 in $1.9 \times 10^8$ SARS-CoV-2 reads in E-10-29-2.
-Such an analysis shows that we cannot reject the null hypothesis that samples Q61 and E-10-29-2 have reads drawn from the same underlying distribution; this null hypothesis could only be rejected at a significance level of $P \le 0.05$ if sample Q61 contained $\ge$6 SARS-CoV-2 reads.
-The lack of rationale is even more obvious when comparing Q61 and A1 (Table~\ref{tab:consistency}): both samples were negative by RT-qPCR, but Q61 was called positive based on metagenomic sequencing whereas A1 was never sequenced.
-Therefore, if Q61 is called as positive then A1 should be called as indeterminate, since it was never subjected to the assay used to call Q61 as positive.
-
-\section{Discussion}
-I have described the first full analysis of the association between the abundance of SARS-CoV-2 and genetic material from different animal species in environmental samples collected by the Chinese CDC from the Huanan Seafood Market.
-This analysis expands upon prior work by \citet{liu2023surveillance} and \citet{crits2023genetic} in several ways.
-
-First, I establish that the full set of sequencing data uploaded to public databases by the Chinese CDC at the end of March 2023 contains unmodified versions of all files earlier analyzed by \citet{crits2023genetic}, as well as a large number of additional files.
-Therefore, regardless of the disputes about the original access to these files~\citep{cohen2023new,gisaid_statement,crits2023genetic}, all the files earlier posted on GISAID are now available without alteration in public databases.
-
-Second, I validate the finding of both \citet{crits2023genetic} and \citet{liu2023surveillance} that the environmental samples contain genetic material from many species, including humans, various fish, various snakes, cows, goats, pigs, sheep, birds such as ducks and spotted doves, raccoon dogs, bamboo rats, and a long list of other animals.
-I am only able to quantitatively compare my analyses of metagenomic compositions to the mammalian compositions reported by \citet{crits2023genetic}, since \citet{liu2023surveillance} provide neither numerical results nor relevant computer code, and \citet{crits2023genetic} only provide numerical results for mammalian species.
-For mammalian species, the compositions determined by my analyses are highly correlated with those reported by \citet{crits2023genetic}, which provides a reassuring robustness check on both analyses.
-However, it is important to emphasize that metagenomic compositions depend on the methods and reference genomes used, so future analyses using different approaches (e.g., full versus mitochondrial genomes) would likely yield somewhat different results.
-I suggest future work should provide full computer code and tabulated results for all species (as done in the current study) to facilitate reproducibility and comparability across studies.
-
-Third, I perform the first comprehensive analysis of the association between the abundance of SARS-CoV-2 and mitochondrial genetic material across all environmental samples.
-This analysis reveals that the greatest co-mingling of viral and animal material involves species that were almost certainly not infected by SARS-CoV-2, such as fish (e.g., largemouth bass, catfish) and livestock (e.g., cows, sheep, goats).
-Consistent with analyses by the Chinese CDC~\citep{liu2022surveillance}, I find some correlation between the abundance of SARS-CoV-2 and human genetic material, but this correlation is weaker than for several non-infectable animals and so on its own is insufficient to identify the source of the viral material.
-Mitochondrial material from most susceptible non-human species sold live at the market is negatively correlated with the presence of SARS-CoV-2: for instance, 13 of the 14 samples with at least a fifth of their chordate mitochondrial material from raccoon dogs contain no SARS-CoV-2 reads, and the other sample contains just 1 of $\sim$200,000,000 reads mapping to SARS-CoV-2.
-These findings are compatible with the results in \citet{crits2023genetic}, since that study did not report any analysis of SARS-CoV-2 content.
-However, they are somewhat inconsistent with related media articles that emphasized co-mingling of raccoon dog and viral material~\citep{wu2023atlantic,mueller2023nytimes}---in fact, raccoon dogs are one of the species with the least co-mingling of their genetic material and SARS-CoV-2.
-The basic finding that SARS-CoV-2 material is not associated with material from susceptible non-human species sold live at the market is largely robust to examining subsets of the samples (such as just those collected on specific dates), and I provide interactive plots to facilitate visualizing the data in different ways.
-
-Fourth, I create a version of the SARS-CoV-2 versus species correlation plot in the fourth figure of the original Chinese CDC preprint~\citep{liu2022surveillance}, but with full annotation of all species.
-The identity of the non-human species correlated with SARS-CoV-2 content had previously been the subject of speculation~\citep{cohen2022anywhere, cohen2022studies, cohen2023unearthed}.
-It turns out these species are fish and livestock that are unlikely candidates for having been infected with SARS-CoV-2.
-Even if the plot is restricted to just samples collected on the date of most intense sampling of wildlife stalls (January-12-2020), no non-human susceptible species is among those most correlated with SARS-CoV-2 content.
-
-Fifth, my analysis calls into question the consistency of the criteria used to classify environmental samples as positive versus negative.
-For instance, sample Q61 became the subject of widespread media coverage~\citep{wu2023atlantic, mueller2023nytimes} because it contains raccoon dog genetic material and was classified as SARS-CoV-2 positive by the Chinese CDC.
-However, this sample tested negative by RT-qPCR and appears to have been called positive on the basis of containing 1 of $\sim$200,000,000 reads that mapped to SARS-CoV-2.
-But all environmental samples contain a mix of genetic material from numerous sources, so it is not consistent to classify this particular sample as positive when hundreds of other samples that also tested negative by RT-qPCR are classified as negative because they were never sequenced or had SARS-CoV-2 content statistically indistinguishable from 1 of $\sim$200,000,000 reads.
-I suggest that future work analyzing the spatial distribution of SARS-CoV-2 across the Huanan Seafood Market~\citep{worobey2022huanan,courtier2022sars,crits2023genetic} consider the quantitative content of samples (such as determined from deep sequencing in the current study, or the Ct values in \citet{liu2023surveillance}), and only include negative samples subjected to the full set of measurements used to classify other samples as positive.
-
-Overall my study validates the approach of \citet{crits2023genetic} of using metagenomic analysis of environmental samples to identify animals and animal products sold at the Huanan Seafood Market.
-As noted by \citet{crits2023genetic}, this innovative approach could usefully inform tracing of animals supplied to the market.
-However, the results described here suggest that the utility of these metagenomic analyses does not extend to indicating whether animals at the market were actually infected by virus.
-For instance, the presence of 1 in $\sim$200,000,000 sequencing reads mapping to SARS-CoV-2 in a sample containing raccoon dog genetic material does not suggest raccoon dogs were infected, given that material from many other species that certainly were not infected (such as fish) is far more consistently co-mingled with SARS-CoV-2.
-Of course, the lack of association also does not disprove the possibility of infected animals at the market, particularly at a date substantially preceding the Chinese CDC's collection of samples---it simply suggests that analysis of the combined viral and animal content of the available environmental samples is not informative for shedding light on this question either way.
-
-When considered in larger context, the inability of the environmental samples to inform on the origins of the virus is unsurprising.
-These samples were all collected on January-1-2020 or later, which is at least several weeks after the Huanan Seafood Market became a superspreading site for human infections~\citep{li2020early}.
-Therefore, by the time the samples were collected, SARS-CoV-2 had been spread widely across the market by humans regardless of its original source---as evidenced by the results reported here, which show viral genetic material coincident with material from myriad animals ranging from fish to snakes to mammals.
-The first human infections with SARS-CoV-2 in Wuhan probably occurred no later than November of 2019~\citep{zhang2020viral,van2020emergence,he2020evaluating,pipes2021assessing,pekar2021timing,odni2022,pekar2022molecular}, which is over a month before the Chinese CDC reports that it began to collect samples from the market.
-For this reason, further insight into the origins and early spread of SARS-CoV-2 will likely require learning more about events or cases that occurred no later than November or early December of 2019.
-
-\section{Limitations of this study}
-This study has limitations related to the data, methodology, and samples themselves.
-
-For the data, all sequencing files and related annotations derive from information shared by the Chinese CDC.
-The description of how samples were processed prior to sequencing lacks detail: for instance, \citet{liu2023surveillance} say ``human nucleic acid was removed'' but not precisely how this was done.
-Additionally, the data were released after China's State Council ordered in March of 2020 that all publications and information related to COVID-19 be reviewed by a centralized task force~\citep{kang2020china}.
-It is possible that this centralized Chinese government review influenced which data were released~\citep{hvistendahl2023}.
-
-For the methodology, the metagenomic compositions were calculated by alignment to a reference set of mitochondrial genomes from a large but still incomplete set of species.
-It is possible that some species that deposited genetic material do not have exact matches in this reference set, so slightly different results might be obtained if a different reference set was used.
-Similarly, this study analyzed chordate mitochondrial genomes: different results would be obtained if the reference genome set was expanded (e.g., to all metazoa) or shrunk (e.g., only to mammals), or if the reads were aligned to full rather than mitochondrial genomes.
-In addition, the results might slightly change with different parameters for read alignment, quality filtering, etc.
-To ensure transparency of the methodology in this study, I have provided a fully reproducible computational pipeline (\url{https://github.com/jbloom/Huanan_market_samples}).
-In addition, I have provided interactive plots of the results to help the reader explore the effects of different parameter choices on the final results (\url{https://jbloom.github.io/Huanan_market_samples/}).
-
-The major limitation of the samples is that they were all collected on January-1-2020 or later, which is well after the first human SARS-CoV-2 infections in Wuhan.
-The lateness of the sampling relative to the origin of the outbreak limits the conclusions that can be drawn.
-
-{\small
-
-\section{Methods}
-\subsection{Code and data availability}
-See \url{https://github.com/jbloom/Huanan_market_samples} for a GitHub repository containing a fully reproducible Snakemake~\citep{molder2021sustainable} computational pipeline implementing the analysis described in this paper, starting with downloading of the raw data from NGDC and proceeding all the way through to rendering of the plots shown in the figures.
-That GitHub repository also includes files containing key numerical results.
-See \url{https://jbloom.github.io/Huanan_market_samples/} for interactive plots of the results rendered using Altair~\citep{vanderplas2018altair}.
-
-\subsection{Processing and alignment of deep sequencing data}
-The FASTQ files with the raw deep sequencing data for all samples were downloaded from the NGDC project CRA010170 (\url{https://ngdc.cncb.ac.cn/gsa/browse/CRA010170}).
-The FASTQ files were pre-processed with \texttt{fastp}~\citep{chen2018fastp} to remove low-quality reads; all tabulations of number of reads in this paper refer to the high-quality reads that passed this pre-processing step.
-
-To create an mitochondrial genome alignment reference set, I followed a procedure partially analogous to that described by \citet{crits2023genetic}.
-Note that a similar analysis could also be implemented using more standard metagenomic analysis tools such as Kraken2~\citep{wood2019improved}, as was in fact done in the Chinese CDC study~\citep{liu2023surveillance}.
-The reasons for instead implementing the pipeline used here were two-fold.
-First, to ensure comparability to \citet{crits2023genetic} by using an approach that parallels the one they described.
-Second, to use a reference set that includes all of the potentially susceptible species sold at the market: many of these species lack full nuclear genomes on RefSeq, and one (Amur hedgehog) doesn't even have a mitochondrial genome on RefSeq.
-Therefore, the manual process used to build the mitochondrial reference set described below was necessary to ensure all these species were in the reference.
-
-For the current study, all mitochondrial genomes in the RefSeq database were downloaded, and then filtered to retain only genomes from the phylum Chordata (in contrast, \citet{crits2023genetic} describe retaining all metazoa mitochondrial genomes).
-The reason I limited to chordate rather than metazoa mitochondrial genomes is that all chordates have very similar length mitochondrial genomes, but at the level of metazoa there is wide variation in mitochondrial genome length.
-The Amur hedgehog mitochondrial genome was not present in this set, so it was separately downloaded from Genbank accession KX964606.
-All the mitochondrial genomes were then filtered to remove highly similar ones.
-To do this, \texttt{Mash}~\citep{ondov2016mash} was used to compute the mash distances between all pairs of mitochondrial genomes.
-To ensure the most relevant mitochondrial genomes were retained, I first specified for manual retention the genomes for all the species listed in the third supplementary file of \citet{crits2023genetic}, as well as genomes for some additional relevant species (these species are listed under the \textit{mitochondrial\_genomes\_to\_keep} key in \url{https://github.com/jbloom/Huanan_market_samples/blob/main/config.yaml}).
-After retaining all of these manually specified mitochondrial genomes, the pipeline then greedily iterated through all other genomes choosing for retention each genome that was not within a mash distance of 0.07 of another already retained genome.
-The full set of retained mitochondrial genomes is listed in Table S3.
-
-The retained mitochondrial genomes were then concatenated, and added to the SARS-CoV-2 reference genome (Genbank accession NC\_045512v2) with its polyA tail trimmed to avoid spurious alignments.
-This concatenation of the SARS-CoV-2 and mitochondrial genomes served as the reference for all sequencing read alignment.
-
-The sequencing reads were then aligned to this genome using \texttt{minimap2}~\citep{li2018minimap2} in the \texttt{sr} (short-read) mode.
-The resulting alignments were filtered to only retain primary alignments with mapping quality of at least 4.
-Note that one effect of counting only primary alignments with mapping quality $>$0 (4 in this case) is to only count reads that map uniquely---that is, they map better to the mitochondrial genome of one species than all other species in the reference.
-This is because although \texttt{minimap2} assigns the primary mapping randomly  if there are multiple sequences in the reference with equally good mappings, a read will only have a mapping quality that exceeds zero if it maps to one location better than all other locations.
-
-I then used \texttt{CoverM} (\url{https://github.com/wwood/CoverM}) to compute the number of aligned reads for each run, requiring alignment lengths of at least 40 nucleotides with at least 95\% identity, and excluding the 100 nucleotides at the contig ends.
-The resulting statistics on alignment counts to SARS-CoV-2 and the mitochondrial genomes for each sequencing run are in Tables S4 and S6.
-To get per-sample alignment counts, I aggregated counts for all runs for each sample to get the statistics in Tables S5 and S7.
-For the mitochondrial genome compositions and all other analyses in this paper, I only included only the samples from metagenomic sequencing, which are annotated with the description ``RNA sequencing of total nucleic acids from environmental swabs for metagenomics'' in the metadata provided by \citet{liu2023surveillance}.
-Species were only retained in the annotated reference set if they had at least 20\% of all aligned mitochondrial reads and at least 4000 covered based for at least one run, or if they are one of the species specified under the \textit{mitochondrial\_genomes\_to\_keep} key in \url{https://github.com/jbloom/Huanan_market_samples/blob/main/config.yaml}.
-The species specified under that key include all susceptible species studied in \citet{crits2023genetic}, so this criteria does not lead too dropping of any of the species in their study.
-
-\subsection{Assembling and aligning contigs for sample Q61}
-For sample Q61, contigs were also assembled and aligned to full genomes to facilitate comparison to \citet{crits2023genetic}, who also performed a similar analysis only for this sample.
-The contigs were assembled using \texttt{Trinity}~\citep{grabherr2011trinity}, and then aligned to using \texttt{minimap2}~\citep{li2018minimap2} to concatenated full genomes for chicken, dog, raccoon dog, and duck.
-Contig alignments were only reported if they had a mapping quality of at least 10, an alignment length of at least 300 nucleotides, and an identity of at least 98\%.
-
-\section{Acknowledgments}
-I thank \citet{liu2023surveillance} for publicly sharing via the NGDC database the deep sequencing data analyzed in this study.
-An initial version of this preprint was revised in part based on helpful comments posted on \textit{bioRxiv} by Florence Débarre and Alex Crits-Christoph (see appendix posted as supplementary material).
-I thank Benjamin F. Pierce for identifying an error in the reference genome description in the original version (\url{https://github.com/jbloom/Huanan_market_samples/pull/3}).
-This study utilized the Fred Hutch Scientific Computing infrastructure, which is supported in part by NIH grants S10-OD-020069 and S10-OD-028685.
-The author is an Investigator of the Howard Hughes Medical Institute.
-
-\section{Competing interests}
-JDB is on the scientific advisory boards of Apriori Bio, Aerium Therapeutics, Invivyd, and the Vaccine Company; consults for GSK; and receives royalty payments as an inventor on Fred Hutch licensed patents related to deep mutational scanning of viral proteins.
-
-\bibliography{references}
-}
 
